### 2018-03-14 article/ms.tex
add article from overleaf

%% This document is part of the snob project
%% Copyright 2017 the authors. All rights reserved.
\documentclass{elsarticle}
\journal{Artificial Intelligence Journal}
\usepackage{bm,url,amsmath,amssymb}
%\usepackage{boldsymbol}
\usepackage{color,colortbl} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.
\usepackage{algorithm,algorithmicx}
\usepackage{algpseudocode}
% For revision history
\IfFileExists{vc.tex}{\input{vc.tex}}{
    \newcommand{\githash}{UNKNOWN}
    \newcommand{\giturl}{UNKNOWN}}

\newcommand{\todo}[1]{\textcolor{red}{#1}}
\usepackage[T1]{fontenc}
\newcommand{\nbb}[2]{
    % \fbox{\bfseries\sffamily\scriptsize#1}
    \fcolorbox{black}{cyan}{\bfseries\sffamily\scriptsize#1}
    {\sf$\blacktriangleright$\textcolor{blue}{\textit{#2}}$\blacktriangleleft$}
    % \marginpar{\fbox{\bfseries\sffamily#1}}
}

\newcommand{\nbt}[2]{
    % \fbox{\bfseries\sffamily\scriptsize#1}
    \fcolorbox{black}{red}{\bfseries\sffamily\scriptsize#1}
    {\sf$\blacktriangleright$\textcolor{blue}{\textit{#2}}$\blacktriangleleft$}
    % \marginpar{\fbox{\bfseries\sffamily#1}}
}
\newcommand{\nbd}[2]{
    % \fbox{\bfseries\sffamily\scriptsize#1}
    \fcolorbox{black}{green}{\bfseries\sffamily\scriptsize#1}
    {\sf$\blacktriangleright$\textcolor{blue}{\textit{#2}}$\blacktriangleleft$}
    % \marginpar{\fbox{\bfseries\sffamily#1}}
}
\newcommand{\alda}[1]{\nbb{Aldeida}{#1}}

% Define commands
\newcommand{\article}{\emph{Article}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}

% Surveys
\newcommand{\apogee}{\acronym{APOGEE}}
\newcommand{\ges}{\acronym{GES}}
\newcommand{\hermes}{\acronym{HERMES}}
\newcommand{\galah}{\acronym{GALAH}}
\newcommand{\fourmost}{\acronym{4MOST}}
\newcommand{\weave}{\acronym{WEAVE}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\Cspace}{$\mathcal{C}$-space }
\newcommand{\Cspaceno}{$\mathcal{C}$-space}
% Common terms
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}

\def\teff{T_{\rm eff}}
\def\logg{\log{g}}

% Have some maaaath.
\def\infinity{\rotatebox{90}{8}}
\def\veccov{\vect{C}}
\def\vecmean{\vect{\mu}}
\def\vectheta{\vect{\theta}}
\def\weight{w}
\def\weights{\vect{\weight}}
\def\datum{y}
\def\data{\vect{\datum}}
\def\likelihood{\mathcal{L}}
\newcommand{\fisher}[1]{\mathcal{F}\left(#1\right)}
\newcommand{\detfisher}[1]{\left|\fisher{#1}\right|}
\newcommand{\prior}[1]{p\left(#1\right)}

% Affiliation(s)
\newcommand{\moca}{
    \affil{School of Physics and Astronomy, Monash University, 
        Melbourne, Clayton VIC 3800, Australia}}
\newcommand{\claytonfit}{
    \affil{Faculty of Information Technology, Monash University,
        Melbourne, Clayton VIC 3800, Australia}}
\newcommand{\caulfieldfit}{
    \affil{Faculty of Information Technology, Monash University,
        Melbourne, Caulfield East VIC 3145, Australia}}

\begin{document}
    \begin{frontmatter}
%       \title{Searching for the Origins of the Galaxy with the Minimum Message Length}
%       \title{Searching for the Origins of the Galaxy with Minimum Message Length mixture modelling}
        \title{Reconstructing the Origins of the Galaxy with Minimum Message Length mixture modelling and new search techniques}

%% Group authors per affiliation:
\author[moca]{Andy Casey}
\author[fit]{David L. Dowe}
\author[fit]{Aldeida Aleti\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{aldeida.aleti@monash.edu}
\author[moca]{John Lattanzio}
\address[moca]{School of Physics and Astronomy, Monash University, Melbourne, Clayton VIC 3800, Australia}
\address[fit]{Faculty of Information Technology, Monash University, Melbourne, Clayton VIC 3800, Australia}

        \begin{abstract}

        \end{abstract}

        \begin{keyword}
            Search\sep Minimum Message Length\sep MML\sep mixture models\sep Galactic Archaeology
        \end{keyword}
    \end{frontmatter}
\section{Introduction} 
\label{sec:introduction}

The contributions of this paper are:
\begin{itemize}
\item Solving the galactic archaeology (GA) problem using mixture modelling
% bear in mind or don't forget that GA will mean `genetic algorithms' to many readers
\item Minimum Message Length (MML) formulation
\item 2 search methods
\end{itemize}
\section{Galactic Archaeology}
% "Galaxy produces ‘molecular forests’"
% https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history of our Galaxy by identifying the remains of a) ancient star formation events and b) smaller galaxies accreted by our own Galaxy during its youth. This is possible because stars form in clusters of 10s to 1000s at a time, all sharing the same composition, being that of the gas cloud from which they formed. This composition is due to the action of previous generations of stars in fusing light elements into heavy ones, and leads an improved understanding of nuclear processes in stars.
%and which stars they operate in.

The idea of Galactic Archaeology relies on identifying stars born together based on their composition. Internal mechanisms change the abundance of the lighter elements, but the heavier ones remain mostly unchanged. These provide a ``chemical tag'' or a ``stellar DNA'' that can be used to identify the stars.

Observations have shown that each star cluster has a uniform and distinct chemical fingerprint~\cite{pancino2010chemical}. Our aim is to identify stars with the same composition, as these
% were
are widely believed to have been
almost certainly
formed at the same time in the same cluster. We expect this cluster has now dispersed, but the tell-tale chemical tags will be present and will allow us to identify the groups of stars that are the relics of ancient star forming activity.

Little work  has been done to address this problem so far, apart from the work of \cite{mitschang2014quantitative}, who used the \emph{Manhattan distance} metric to  quantify the chemical difference between any two stars, and as a result estimate the level of homogeneity expected within a cluster. This metric calculates the mean weighted absolute difference of abundance levels, $\sigma_C$, between two stars $S^a$ and $S^b$ as follows:
% Given that $\sigma$ is so widely used to mean standard deviation, this notation (to follow) seems a bit unfortunate
$\sigma_C = \sum_{c \in C}w_c|S^a_c-S^b_c|/||C||$, where $C$ is the set of chemical species, $S^a_c$ is the chemical abundance of element $c$ in star $S^a$, $w_c$ is the weight given to species $c$, 
%which is considered 1 for all elements in the current work, 
and $||C||$ is the total number of chemical species. The Manhattan distance values are high for stars from different clusters, and low for stars from the same cluster. Note that this metric is not particularly robust or versatile; for example it is not even invariant to a simple rotation of the co-ordinate axes
% - and nor does it take into account that a difference between two stars of (say) 0.01\% in a rare element might be
% vastly more significant than the same difference in a vastly more abundant element
% Retraction.  It does endeavour to take that into account, via the weights, $w_c$, but not by setting them all to 1.
An important issue is how to set the weights for each chemical species $w_c$, and how they affect 
the results (Mitschang et al.~\cite{mitschang2014quantitative} used $w_c=1$). Another issue is setting the boundary between clusters. It is not clear what a high and low Manhattan distance is; at the moment 
these values are largely subjective. In this paper, we express the problem in terms of mixture modelling and infer the number of clusters, their mixing proportions (or relative abundances) and each cluster's statistical distribution using Minimum Message Length (MML) and two novel search techniques.

\section{Related Work in Mixture Modelling}

Mixture modelling is composed of three main parts: (i) an estimator of component parameters of a mixture, (ii) an objective functions that scores a mixture, and (iii) a search strategy which identifies the best number of components and their weights.

Different criteria for estimating component parameters have been used in the literature, with traditional methods using maximum likelihood estimation (MLE)
% Perhaps because ML could refer to many things (e.g., machine learning), maximum likelihood is often (I think usually) called MLE
~\cite{akaike1971determination} or maximum a posteriori probability (MAP)~\cite{shimony1994finding}.
% I re-write the draft below.
% Traditional methods, such as ML or MAP, have not been designed to consider the complexity of the model, but only focus on the goodness-of-fit. They have to be modified with ad-hoc principles in order to ensure that the data is not over-fitted. This is a very important property for astrophysical applications. 
%
% In this work, we focus on the Bayesian Minimum Message Length Principle~\cite{Wallace05}, which unlike MAP estimators, are invariant under non-linear data transformations.
%, and different from MLE, can be used with different probability distributions.  
In this work, we focus on the Bayesian information-theoretic Minimum Message Length (MML) principle~\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a, Wallace05}.
%
The MML principle states that the best explanation of the data is the one that leads to the shortest two-part message~\cite{Wallace05}, where a \textit{message} takes into account both the complexity of the model and its explanatory power (corresponding respectively to the lengths of the first and second parts of the message).
MML has been shown to perform well on a variety of empirical analyses (see, e.g., \cite{viswanathan1999finding,fitzgibbon2004minimum}, with
% further examples and
references to further examples in \cite{Wallace05,dowe2007bayes,Dowe2008a,Dowe2011a}.
Arguments about the statistical consistency
(i.e.,~as the number of data points increases the distributions of the estimates become increasingly concentrated near the true value)
of MML are given in
\cite{DoweWallace1997a,Dowe2011a}.
Unlike the rival Bayesian approach MAP, for continuous parameters MML is a statistically invariant (meaning that $1$-to-$1$ transformations of co-ordinates, such as polar $(r, \theta)$ to Cartesian $(x, y)$, leave the answer unchanged) Bayesian approach~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}. It allows the modelling of prior beliefs in the calculation of the best model parameters. If one has a particular reason to expect one value rather than another, a prior probability can be chosen to represent that belief. Of course, we can also model the data as objectively as Bayesianism permits.

Two other well-known methods based on information theory are the Akaike Information Criterion (AIC)~\cite{akaike1974} and Bayesian Information Criterion
(BIC)~\cite{schwarz1978estimating}, both of which endeavour to temper Maximum Likelihood's tendency to over-fit and under-estimate the noise.
% Unlike Maximum Likelihood, MML takes into account model complexity - and, unlike MAP, it is invariant under non-linear continuous transformations.
In MML terms, Maximum Likelihood can be thought of as MML without paying for the first part of the message.

AIC~\cite{akaike1974} chooses a model 
% that minimizes
with the motivation of minimizing
the Kullback-Leibler
% distance
divergence
between
% KL divergence is generally asymmetric
the model and the truth, and is defined as:

\begin{equation}
AIC = -2 (\ln ( f(\vec{x} | \vec{\theta} )) + 2 k
\end{equation}
where the likelihood $f(\vec{x} | \vec{\theta})$ is the probability of the data $\vec{x}$ given a model parameterised by $\vec{\theta}$,
and $k$ is the number of free parameters in the model.

BIC~\cite{schwarz1978estimating} uses a constant penalty equal to $(\log N)/2$ for each free parameter in the model,
where $N$ is the number of data points.
Minimum Description Length (MDL)~\cite{rissanen1978modeling} was independently introduced in the same year (1978) as BIC with a very similar formula, a decade after MML \cite{WallaceBoulton1968}.
Both AIC and BIC/MDL associate parameter costs with the number of free parameters and not their values. In this case, all models of a particular probability distribution have the same cost, regardless of the respective means, covariance matrices and Fisher information,
which may hinder accurate inference.

% Comparisons of MML to other methods
% clearly indicate MML's theoretical and empirical superiority.
% Let's not overly annoy the reviewers and readers.
Examples of MML's promise against rival methods include comparisons with AIC~\cite{dowe2007bayes}, BIC \cite{agusta2002mml}, MDL~\cite{viswanathan1999finding} and MLE~\cite{dowe2007bayes} - and also in statistical econometric time series autoregression \cite{fitzgibbon2004minimum}, machine learning problems \cite{viswanathan1999finding},
angular data \cite{WallaceDowe1993,DoweOliverWallace1996}
and using this modelling \cite{WallaceDowe1993} to infer the order in which various parts of a protein fold~\cite{edgoose1998mml}. 
%
A further example is the problem of deciding whether or not some 1-dimensional data generated uniformly in the interval $[ 0, 1 ]$ has a gap in it. Curiously, even when there is clearly no gap, AIC infers that there is a gap, whereas MML gives the right answer \cite{dowe2007bayes}.

Very importantly, the MML message structure enables us to robustly deal with missing data \cite{WallaceDowe1994b,WallaceDowe1997,WallaceDowe2000}. Other methods often replace missing values with the mean - in turn, reducing the true standard deviation - or various other imputed methods. In MML, on the other hand, missing data can easily be left out from the calculations from modelling done so far, simply encoding the missing data as missing \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}, which is very useful when modelling astrophysical data, where missing observations are quite common.

The parameters of a mixture model can be estimated using Expectation-Maximisation (EM)~\cite{dempster1977maximum}. The EM algorithm has been used with MML to infer Gaussian mixtures under different simplifying assumptions, such as assuming that the covariance matrices are diagonal~\cite{WallaceDowe1994b, WallaceDowe1997a, WallaceDowe2000} (where it is sometimes called an `adjust cycle') and \cite{oliver1996unsupervised}, or approximating the probabilities of mixture parameters~\cite{figueiredo2002unsupervised,roberts1998bayesian}.

The search algorithm employed in the majority of these approaches is based on running EM iteratively for different numbers of components~\cite{WallaceDowe1994b, WallaceDowe1997a, WallaceDowe2000, oliver1996unsupervised,roberts1998bayesian,biernacki2000assessing}. The search method proposed by Figueiredo and Jain~\cite{figueiredo2002unsupervised}, on the other hand, starts with a large number of components, and iteratively deletes components. An extension of this method was proposed by Kasarapu and Allison~\cite{kasarapu2015minimum}, which starts with one components, and selectively splits, deletes, or merges components in an attempt to minimise the message length. While this method removes the overhead introduces by the search algorithm of Figueiredo and Jain~\cite{figueiredo2002unsupervised}, the search remains greedy, and can be prone to premature convergence in suboptimal solutions. The search methods introduced in this paper, incorporate a perturbation step which encourages explorations of new local optima.

{\color{blue}References that we/you seem to be missing include
\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a}
% these three above are probably the three most-cited MML papers
and
WallaceDowe1993,WallaceDowe1994b,DoweOliverWallace1996,
% these three works above are the earliest MML works on angular data and mixture modelling with angular data
DoweWallace1997
% the above is a work about MML statistical consistency
WallaceDowe1997,WallaceDowe2000
% these two papers above are MML mixture modelling papers
Dowe2008a,Dowe2011a
% these two papers above are MML survey pieces

Things that could also be added include
ConwaySloane1988
% the above is a book on lattices and lattice constants
EdwardsDowe1998
% this was (and still is) a work on single latent factors in mixture modelling with application to astro' data
McLachlanPeel2000
% this is quite a popular book on finite mixture models
}
\section{Minimum Message Length (MML)}
A complex model is not an optimal one, unless its complexity is justifiable by the added explanatory power. There is an established connection between simplicity and truth, as beautifully elucidated by the Occam's Razor principle. MML is a statistical technique that formalizes and quantifies this principle.

The basic idea is to consider the transmission of a set of data and a model that describes it. Clearly one can make increasingly complex models to obtain better fits -- think of epicycles in the solar system. Suppose we try to best represent N data points.

Traditional curve fitting methods would give a polynomial of degree N-1, which is perhaps the most precise, but it is also the most complex, and the one that most models the noise, by spuriously {\it over-fitting non-existing patterns.\/}

Over-fitting can result in a model that may be far from the truth. MML will only consider a more complex model if the encoding of the complex model and the data is more efficient  than the less complex model, i.e. it finds the minimum message length needed to transmit the data and the model. It is this \textit{true} model that MML aims to find.

The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace05}, 
where a \textit{message} takes into account both the complexity of the model, and its explanatory power.

In essence, MML infers the most concise way of encoding the model, $\theta$, and the data, $x$, given the model, $x|\theta$. A complex model will produce a longer first part of the message than a simple model, since more quantities must be stated. On the other hand, the length of the second part of the message decreases with more accurate and complex models, since less data has to be described fully.

The \textit{message} must encode two parts: the model, and the data. The encoding of the message is based on Shannon's information theory. According to Shannon, the information we gain from an event $e$ is $h(e) = - \log_2 p(e)$, where $p(e)$ is the probability of that event. The information content is biggest for the improbable outcomes, and smallest for outcomes that we are almost certain about. An outcome that has a probability close to one has close to zero information content, since we don't learn anything new from it, whereas the rarer events have much higher information content.

%%%% This section below looks like a variant of a lift-out from the ARC DP16 grant proposal.
%%%% One problem with the presented text is that 4 does not equal 6.
%%%% Mightn't it be easier to talk about Morse Code?
%%%% In certain parts of the astronomy literature, such an example (when tidied up) might be suitable, but I think that - in the AI Journal - such an example needs to be no longer than short. I think that, here, all we need to do is mention Morse code and (e.g.) the Huffman code example from \cite[sec. 2.1.4]{Wallace05}.
% Consider the simplified illustrative example of rolling an unbiased four-sided die (d4), where each of the four outcomes $(1,2,3,4,5,6)$ have equal probability $p_i=0.25$. The information content in each datum is $- \log_2 0.25 = 2$. In total $\sum_{i=1}^4-\log_2 (p_i) = 2 + 2 + 2 + 2 = 8$ bits are required to describe the four outcomes. The model description could be $\{00,01,10,11\}$:~this is the first part of the message.

% Now assume that the die is biased, with probabilities $p(1)=0.5, p(2)=0.25, p(3)=0.125$, and $p(4)=0.125$. In this case, the number of bits required to represent each alternative is different. One bit is required for 1, since $-\log_2 0.5 = 1$, and the codeword could be the single binary value 1. Similarly, 2 bits are required for the outcome 2, which can be represented with codeword $01$. It follows that possible codewords for the remaining alternatives, all using the bit ``1'' to indicate the end of each datum, are $001$ for 3, $0001$ for 4, $00001$ for 5, and $000001$ for six.

% The length of the model required to describe the outcomes of the biased die would be $\sum_{i=1}^4-\log_2 (p_i) = 1 + 2 + 3 + 3 = 9$ bits, and the codewords would be $\{0,10,110,111\}$. The second description is 1 bit longer than the first, but, since the most frequent events have a shorter codeword, the second part of the message (the data given the model) is shorter. The overall message length of the second description is shorter than the first, when applied to the biased die.

% To illustrate this point let's encode 1000 outcomes of a biased die, where 500 outcomes are 1, 250 are 2, 125 are 3, and 125 are 4. With the first model each outcome is encoded with two bits, so the length of the second part of the message is $1000\times 2=2000$. The total message length is the sum of the model length and the length of the data given the model, or 2008 bits. But if we encode the data with the second model, the length of the data given the model is $500+(250\times2)+(125\times3)+(125\times3)=1750$. The total length is $1750+9=1759$. Thus the second model, although it is more complex (takes more bits to describe), creates a shorter encoded message, 
% and hence is a better description than the shorter model.
%%%% As per above comment, I think that the text from the above comment to here can all go.

Formally, the length of the first part of the message is $- \log_2 h(\vec{\theta})$, and that of the second part is $- \log_2 f(\vec{x}|\vec{\theta})$, where $h(\theta)$ is the prior probability of model $\theta$, and $f(x|\theta)$ is the likelihood of data $x$ assuming a model parameterised by $\vec{\theta}$. The total message length is the sum of the two parts, i.e. $- \log_2 h(\vec{\theta}) - \log_2f(\vec{x} | \vec{\theta})$, which is minimised precisely when $h(\vec{\theta}) f(\vec{x}| \vec{\theta})$ is maximised.
The multiplication of the prior and the likelihood is proportional to the posterior {\em probability} $g(\theta|x)$. An important but often misunderstood subtle point is that we are quantising (or rounding off in) parameter space, and so $h(\theta)$ corresponds to a {\em probability} and {\em not} a density. Thus, the MML estimate can in some sense be thought of as maximising $h(\theta)f(x|\theta)$ and in turn (in some sense) equal to the posterior mode. This is similar in spirit to the (Bayesian) Maximum A Posteriori (MAP) posterior mode, but
(partly because both terms $- \log_2h(\vec{\theta})$ and $\log_2 f(\vec{x}|\vec{\theta})$ here correspond to probabilities)
with the benefit of statistical invariance under re-parameterisation
- for further discussion, see, e.g., ~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes} or immediately below.

\section{MML for Mixture Modelling}
MML is used to identify groups (also known as ``classes'') of stars with the  same composition. Mixture modelling enables the partitioning of the data into overlapping groups.

In MML we model this by constructing a two-part message, with the first part describing the mixture model (or hypothesis, $\theta$) and the second part representing the encoded data ($x$) given the hypothesis. The MML \textit{message} contains the following components:

\begin{itemize}
\item[\textbf{1a)}] A statement of the number of groups (or classes, or Gaussian mixture components), $K$. Each class is assumed to have the same prior probability, equal to $2^{-K}$. The length of this part is equal to the negative logarithm of the prior, i.e. $K$. 
\item[\textbf{1b)}] The relative abundances (or relative weights or mixing proportions) of the amount(s) of data in each group.
Only $(K - 1)$ weights need be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
The length of this part is equal (within a small constant) to the negative log of the multiplication of the prior probabilities of each group ($h(\vec{p})$) divided by the square root of the expected
Fisher information ($\sqrt{F(\vec{p_k})}$). 
        The Fisher information matrix contains the expected second-order partial derivatives of the log-likelihood function, and quantifies how sharply the likelihood function peaks, and as a result how precisely the parameters should be stated.

    In the case of uniform prior probabilities $h(\vec{p}) = (K-1)!$, and the Fisher information $F(\vec{p}) = N^{K-1} / (\prod_{k=1}^{K} p_k)$. It follows that the length of this part is equal to $- \log( h(\vec{p}) \times (1 / \sqrt{{\kappa_K}^K F(\vec{p})}) ) = - \log((K-1)!) + (K/2) \log(\kappa_K) + ((K-1)/2) \log(N) - (1/2) \sum_{k=1}^{K} \log(p_k)$, where $\kappa_K$ is a lattice constant (varying between $1/12$ and $1/(2 \pi e)$).

\item[\textbf{1c)}]  The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian components.

\item[\textbf{2)}] The data given the model and its parameters.
% , whose cost is the well-known statistical log-likelihood. In the case of a Normal distribution, this % would appear to a first approximation to equal 
%   $$-\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}).$$
To a first approximation, this would appear to be
$ \min_{k} -\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) \rm{,}$
minimising (over cluster indices $k$) $ -\log (p_k(x)) $ (which is the cost of encoding the particular cluster, $k$) plus the negative log-likelihood given cluster $k$'s parameters.
However, as per \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}\cite[sec. 6.8]{Wallace05}, the cost can be reduced to
$ -\log ( \Sigma_{k=1}^{K} p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) $
\end{itemize}

Parts 1a, 1b and 1c encode the parameters from the clustering process, and part 2 encodes the likelihood (the data given the hypothesis), which represents the goodness-of-fit.

The estimation of the multinomial parameters from part 1b is done by quantifying the trade-off between stating the parameters imprecisely, so as to have higher prior probability in the relevant uncertainty region and keep the first part of the message short, and the goodness of fit to the data, which is typically better if we state the parameter estimates more precisely.

Quantitatively, this is done by minimising the sum of the lengths of the three components (1a, 1b, 1c and 2 above). If MML finds that the data for a specific element do not affect the selection of groups, then it will {\it ignore that element.\/} This could tell us something important - such as that there is a large random component for that element or that there are significant observational errors ruining the usefulness of that measurement. These are significant insights into the data that are, importantly, quantified by the MML method.

\subsection{Message Length}

The MML principle requires that we encode everything required to reconstruct the message, including our prior beliefs on the model parameters. Consider that we have $N$ data points each with $D$ dimensions which are to be modelled by a finite number of Gaussian mixtures. Therefore the message for a finite number of Gaussian mixtures must encode:

% \begin{enumerate}
% k has been used previously in describing AIC and BIC.
% If k is the number of parameters, then it will be neither (the number of clusters) nor an index running from 1 to (the number of clusters).
% \item
% \begin{itemize}
% \item (1a) ~ The number of Gaussian mixture components, $K$.
% \item (1b) ~ The relative weights $\weights$ of the $K - 1$ Gaussian components. Only
%       $K - 1$ weights need be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
% \item (1c) ~ The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
%       components.
% \end{itemize}
% \item (2) ~ The data, given the model parameters.
%  \item The lattice quantisation constant $\kappa(Q)$ for the number of model       parameters $Q$.
% \end{enumerate}

We provide the message length of each component in turn before outlining our
full objective function. We assume a prior on the number of mixtures $K$ of
% $\prior{K} \propto 2^{-K}$
$\prior{K} = 2^{-K}$
\cite[p279, sec. 6.8.2, (a)]{Wallace05} such that the length of the 
optimal lossless message to encode $K$ is 
\begin{equation}
    I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant}.
\end{equation}

We assume a uniform prior on the individual weights $\weight_{k}$. The weights
$\weights$ are drawn from a multinomial distribution such that they can be 
optimally encoded in a message with length~\todo{\cite{someone}}:
\begin{equation}
  I(\weights) 
    = \frac{K - 1}{2}\log{N} 
    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
    - \log{\Gamma{\left(K\right)}} \quad .
\end{equation}

In order to properly encode the means $\vecmean$ and covariance matrices
$\veccov$ for all $K$ mixtures, we must encode both our prior belief on 
those parameters and the determinant of the expected Fisher information 
matrix. For the $k$-th mixture the joint message length is
\begin{eqnarray}
  I(\vecmean_k,\veccov_k) &=& -\log{\left(\frac{\prior{{\vecmean_k,\veccov_k}}}{\sqrt{\detfisher{{\vecmean_k,\veccov_k}}}}\right)} \nonumber \\ 
  I(\vecmean_k,\veccov_k) &=& -\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\log{\detfisher{{\vecmean_k,\veccov_k}}}
\end{eqnarray}

\noindent{}and for all mixtures:
\begin{equation}
  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \quad .
  \label{eq:I_component_params}
\end{equation}

We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
that 
    $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
However,  the MML principle \todo{demands} proper priors. In practice we make 
this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
we do so without actually specifying the value of the bounds because they add 
only a constant value to the total message length\todo{\cite{someone}}.
We adopt a conjugate inverted Wishart prior for the covariance matrix of an
individual mixture
- e.g.,
%[]
\cite[Section 5.2.3]{Schafer_1997}:

\begin{equation}
  \prior{{\vecmean_k, \veccov_k}} \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
  \label{eq:covariance-prior}
\end{equation}

% Casey: Reword following paragraph because it is clumsy.
Following Eq. \ref{eq:I_component_params}, we require the logarithm of the
determinant of the expected Fisher information matrix for $\vecmean_k$ and
$\veccov_k$ in order to encode the full message length. We approximate the
determinant of the Fisher information $\detfisher{{\vecmean_k, \veccov_k}}$ 
as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
we take the second derivative of the log-likelihood function 
$\likelihood(\data|\vecmean,\veccov)$, which yields:
\begin{equation}
  \detfisher{\vecmean_k} = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
\end{equation}
We make use of an analytical expression for $\detfisher{\veccov_k}$
\cite{Dwyer_1967,Magnus_1988}:
\begin{equation}
  \detfisher{\veccov_k} = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad .
\end{equation}
Combining these expressions gives our approximation for the determinant of
the expected Fisher information $\detfisher{{\vecmean_k,\veccov_k}}$:
\begin{eqnarray}
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & \detfisher{\vecmean_k} \cdot \detfisher{\veccov_k} \nonumber \\
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
  \label{eq:mixture-component-det-fisher}
\end{eqnarray}

Substituting Eqs. \ref{eq:covariance-prior} and \ref{eq:mixture-component-det-fisher} 
into Eq. \ref{eq:I_component_params} gives the message length to encode the component
parameters $\vecmean$ and $\veccov$ for all $K$ mixtures:

\begin{eqnarray}
  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \nonumber \\
  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
    + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
    -\frac{KD}{2}\log{2} \nonumber \\
  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
\end{eqnarray}

If we assume that the data have homoskedastic noise properties, then the 
precision of each measurement $\mathcal{E}$ relates the probability of a
datum $p(\datum_n)$ can be related to the probability density of
the datum, given the model 
$p(\datum_n) = \mathcal{E}^{D}p(y_i|\mathcal{M})$.
In this work we adopt $\mathcal{E} = 0.01$, but we note that the value of
$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
a datum given the model is,
\begin{equation}
  I(\data_n) = -\log{p(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
\end{equation}

\noindent{}and for the entire data:
\begin{eqnarray}
  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
\end{eqnarray}

The final part of the message is to encode the lattice quantisation constant,
which arises from the approximation to the (so-called) strict MML, where
parameters are quantised into intervals in high dimensional space in order to
be losslessly encoded.  We use an approximation of the lattice constant
\citep[see Sections 5.1.12 and 3.3.4 of ]
%[]
{Wallace_2005} such that,
\begin{equation}
  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
\end{equation}

\noindent{}where $Q$ is the total number of free parameters in the model. Each
Gaussian component in the mixture requires $D$ parameters to specify $\vecmean$,
and a covariance matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ 
parameters. Because only $K - 1$ weights need to be encoded, the total number of free
parameters is
\begin{equation}
    Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1 \quad .
    \label{eq:number-of-parameters}
\end{equation}

%  Here

 Each $k$-th mixture is a $D$-dimensional Gaussian that contributes a relative weight $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$. We will assume that the data have homoscedastic noise properties.

The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.

The message length for the data is given by:

Here we introduce the individual components that contribute to the message length.

Combining these expressions, the complete message length for $K$ mixtures is given by

\begin{eqnarray}
I_K & = & K\log{2} % I_k
    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
    +\mathcal{L}(\data|\vectheta) \nonumber \\
    & -& DN\log{y_{err}} % Likelihood
   +  \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right]\nonumber \\% I_t
    &-& \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
\end{eqnarray}

\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
$\mathcal{L}$ is the log-likelihood of a multivariate Gaussian distribution.

\section{Search Methods}

\subsection{The MML$_{K+D}$ Method}
Say we wanted to calculate whether another mixture was warranted. If another
mixture were preferred then we would want:

\begin{eqnarray}
  \Delta{}I_{K+1} - I_{K} < 0
\end{eqnarray}

The expression is given as:
\begin{eqnarray*}
\Delta{I_{K+1} - I_K} & = & (K + 1)\log{2} - K\log{2} \\ % I_k^(new) - %I_k^(old)
  &&+ \frac{(K)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K+1}\log{w_k}^{(new)} - \log{|\Gamma(K+1)|} \\ % I_w^(new)
  &&- \frac{(K - 1)}{2}\log{N} + \frac{1}{2}\sum_{k=1}^{K}\log{w_k} + \log{|\Gamma(K)|}\\ % I_w^(old)
  &&+ \mathcal{L}^{(new)} - DN\log{y_{err}} \\ % Likelihood (new)\\
  &&- \mathcal{L}^{(old)} + DN\log{y_{err}} \\ % Likelihood (old) \\
  &&+ \frac{1}{2}\sum_{k=1}^{K+1(new)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(new)
  &&- \frac{1}{2}\sum_{k=1}^{K(old)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(old)
  &&- \frac{Q^{(new)}}{2}\log(2\pi) + \frac{\log{Q^{(new)}\pi}}{2} \\ % lattice, minus the lattice of part 2
  &&+ \frac{Q^{(old)}}{2}\log(2\pi) - \frac{\log{Q^{(old)}\pi}}{2} % lattice, minus the lattice of part 2
\end{eqnarray*}

\noindent{}By making use of $\log{\Gamma(K)} - \log{\Gamma(K + 1)} = -\log{K}$ and re-arranging the expression:

\begin{eqnarray}
\Delta{}I_{K+1} - I_K &=& \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber\\
&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
& +& \frac{\log(2\pi)}{2}(Q^{(old)} - Q^{(new)}) + \frac{\pi}{2}\left(\log{Q^{(new)}} - \log{Q^{(old)}}\right)
\label{eq:13}
\end{eqnarray}

Expanding the $Q$ terms:

\begin{eqnarray}
Q^{(old)} - Q^{(new)} &=& \frac{1}{2}DK(D + 3) + K - 1 - \frac{1}{2}D(K + 1)(D + 3) + (K + 1) - 1 \nonumber \\
Q^{(old)} - Q^{(new)} &=& -\frac{1}{2}D(D+3) + 2K  - 1
\label{eq:14}
\end{eqnarray}

\noindent{}And making use of the following logarithmic identities,
\begin{eqnarray}
  \log{Q^{(new)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3) + K\right)} \nonumber \\
                  &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} + \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \\
  \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}DK(D + 3) + K - 1\right)} \nonumber \\
                  &=& \log{\left(\frac{1}{2}DK(D + 3)\right)} + \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}
\end{eqnarray}

\noindent{}gives us,

\begin{eqnarray}
  \log{Q^{(new)}} &- \log{Q^{(old)}} = \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
                                    &+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}.
\end{eqnarray}
The second row of terms can be ignored because they are very small (typically less than 1 bit). This is because as $K \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 1$, thus $\log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \rightarrow \log{2}$. Similarly as $D \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 0$.

As $K \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 1$ and as $D \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 0$ and thus $\log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \approx 0$.

\noindent{}Ignoring these minor terms:

\begin{eqnarray}
  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{(K + 1)} - \log{K}
  \label{eq:19}
\end{eqnarray}

\noindent{}Substituting Eqs. \ref{eq:19} and \ref{eq:14} into \ref{eq:13} yields:

\begin{eqnarray}
\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber \\
&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
& +& \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
\end{eqnarray}

\begin{algorithm}[H]
\caption{The MML$_{K+D}$ Method.}
\label{alg:MMLKD}
\begin{algorithmic}[1]
\State {\bf \sc input:} $K$ \Comment{Number of initial components.}
\State current $\leftarrow$ $K$ component mixture
\While{TRUE}
\EndWhile
\State {\bf \sc output:} 
\end{algorithmic}
\end{algorithm}

\subsection{Method of Moments}
% \alda{This is a copy paste from David's email with minor changes}
%
% Re the case of two (or more) classes with the same mean:
% In 2D, a simple case in point is two classes with the same mean
% such that $\sigma_{1, x} = \sigma_{2, y} >> \sigma_{2, x} = \sigma_{1, y}$.
% Andy said that this and higher dimensional stuff can be picked up
% by dropping to 1 dimension along the eigenvector of largest eigenvalue.
%
%
% I wonder also about the merits of dropping to 2 dimensions (if we have at least 2 attributes), being the eigenvectors of the two largest eigenvalues. 
%
% One could possibly check whether uniform radially directly from the original 2D projection without having to project it to a circle (i.e., it might be pretty easy to work out what the angle of each point would be post-transformation without having to do the transformation).  The 2D projection is 1D in the sense that we only care about the angle around the circle.  And then some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform in radial distribution.  If not uniform, then perhaps grounds for splitting.
%
% Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
% as above but rather projecting on eigenvector of largest eigenvalue), we
% discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution}.
% So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
%

We informally outline here the method of moments form of search heuristic and then look at how it might be applied.

Let's suppose we have a 1D component.  We'll shift it (without loss of generality, w.l.o.g.) so that its mean is 0 and we'll shrink or expand it (again, w.l.o.g.) so that its s.d. ($\sigma$) is 1 (and variance v is 1).
We consider now splitting it into K components with mean 0,
mixing weights (or proportions) $w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
Let's set $K = 2$, so $w_2 = 1 - w_1$.

Noting that the 2nd, 4th, 6th and 8th moments of a true Gaussian distribution are 1, 3, 15, 105,
suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6$ and $105 \alpha_8, ... $.
We get some equations - if we have the correct number of equations,
they're simultaneous equations.

If we get too many equations (by using too many moments), then it
becomes something of a fitting or regression problem to choose
$w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.

Our simultaneous equations are:

\begin{align*}
w_1 v_1       + (1 - w_1) v_2       &=    1\\
w_1 (v_1)^2  + (1 - w_1) (v_2)^2  &=    3 \alpha_4\\
w_1 (v_1)^3  + (1 - w_1) (v_2)^3  &=   15 \alpha_6\\
w_1 (v_1)^4  + (1 - w_1) (v_2)^4  &=  105 \alpha_8
\end{align*}

% These might be messy to solve on the fly  but one could pre-process by solving many of these beforehand (actually, approximately solving these beforehand, as it's only
% a heuristic, anyway) at the very start of the program while loading in the data.

So, for $1000 \times 1000 = 10^6$ different values of $\alpha_4$ and $\alpha_6$
or for $100 \times 100 \times 100 = 10^6$ different values of $\alpha_4, \alpha_6$ and $\alpha_8$
one could pre-compute {\em approximate} 1,000,000ish `solutions' to the above simultaneous equations.

Then, when running the program with real actual values of $\alpha_4$ and $\alpha_6$ (and $\alpha_8$),
one could just grab something from nearby in this pre-computed look-up table
as putative values for $w_1, v_1, v_2$ (and etc., if need be).

% As an example of this algorithm, consider the mixture $0.6 N(-3, 2^2) + 0.4 N(7, 3^2)$.
% The best fitting single Normal distribution to this would be $N(1, \sqrt{30}^2)$.
%
As an
% another
example of this algorithm, consider a first moment of $0$, a $2^{\rm nd}$ moment of $6$ and a $4^{\rm th}$ moment of $3 \times 37.2 = 111.6$.
Solving the simultaneous equations gives us $0.6 N(0, 2^2) +  0.4 N(0, 3^2)$.

\subsection{The k-means++ algorithm}

The k-means++ algorithm extends the k-means method with an efficient way of choosing centres. Let $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected. The k-means++ has the following steps:

\begin{enumerate}
    \item Selects a centre $c_1$ uniformly at random from the $N$ data points.
    \item Selects a new centre $c_i$ from the data points with probability $\frac{D(x)^2}{\sum_{x\in X}}D(x)^2$
    \item Repeat Step 2. until k centres have been selected.
    \item Proceed as with the standard k-means algorithm.
\end{enumerate}

k-means++ is $O(\log k)$-Competitive.

\section{Experimental Design}

To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets. As the output of the algorithms depends on the starting seed, and some of the methods can be affected by randomness, we perform 50 runs for each search scheme and dataset.

The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms. The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.

\subsection{Benchmark Method}
We use the search algorithm proposed by Kasarapu and Allison as a benchmark for comparison, which
has been shown to be the state-of-the art
% in what sense has this been shown?
~\cite{kasarapu2015minimum}. 
\begin{algorithm}[H]
\caption{The benchmark method~\cite{kasarapu2015minimum}.}
\label{alg:kasarapu}
\begin{algorithmic}[1]
\State {\bf \sc input:} $K$ \Comment{Number of components.}
\State current $\leftarrow$ one component mixture
\While{TRUE}
\State components $\leftarrow$ current
  \For{$i\in K$}\label{line:splitBegin}\Comment{Exhaustively split all components}
    \State splits[i]$\leftarrow$ \Call{Split}{current,components[i]}
  \EndFor
\State  bestSplit$\leftarrow$\Call{Best}{splits}\label{line:splitEnd} \Comment{Record the best split.}
\If{K>1}
  \For{$i\in K$}\label{line:deleteBegin}\Comment{Exhaustively delete all components}
    \State deletes[i]$\leftarrow$ \Call{Delete}{current,components[i]}
  \EndFor
\State  bestDelete$\leftarrow$\Call{Best}{deletes} \label{line:deleteEnd}\Comment{Record the best delete.}
\EndIf
  \For{$i\in K$}\label{line:MergeBegin}\Comment{Exhaustively merge all components}
    \State j$\leftarrow$ \Call{closestComponent}{i}
    \State merges[i]$\leftarrow$ \Call{Merge}{merges}
  \EndFor
\State  bestMerge$\leftarrow$\Call{Best}{merges} \label{line:MergeEnd}\Comment{Record the best merge.}
\State bestPerturbation $\leftarrow$ \Call{Best}{bestSplit, bestDelete,bestMerge} \Comment{Select the best perturbation.}
\State $\Delta I \leftarrow$ \Call{MessageLength}{bestPerturbation}-\Call{MessageLength}{current} \Comment{Check for improvement.}
\If{$\Delta I < 0$}
\State current $\leftarrow$ bestPerturbation
\State {\bf \sc Continue}
\EndIf
\State {\bf \sc Break} \label{line:end}
\EndWhile
\State {\bf \sc output:} current
\end{algorithmic}
\end{algorithm}

The method is presented in Algorithm~\ref{alg:kasarapu}. It starts with one component and iteratively splits (lines \ref{line:splitBegin}-\ref{line:splitEnd}), deletes (lines \ref{line:deleteBegin}-\ref{line:deleteEnd}), and merges (lines \ref{line:MergeBegin}-\ref{line:MergeEnd}) components until no improvements in the message length is observed (line~\ref{line:end}).

\subsection{Dataset}

\alda{Andy, can you please include the details?}
\paragraph{Simulations}

\paragraph{GA data}

\subsection{Criteria for Comparison}

We employ two criteria for comparing the proposed search methods to the benchmark: the message length and the Kullback Leibler divergence~\cite{kullback1951information}.

\paragraph{Message Length}

\paragraph{Kullback Leibler divergence}

\section{Results}
\section{References}
\bibliographystyle{elsarticle-num}
\bibliography{mmlbibliography}

\end{document}

### 2018-03-14 gmmmml/mixture_search.py
minor

@@ -509,7 +509,7 @@ class GaussianMixture(object):
             
             if kwds["visualization_handler"] is not None:
 
-                target_K = weight.size + np.arange(1, 200)
+                target_K = weight.size + np.arange(1, 25)
                 self._predict_message_length(target_K, cov, weight, y.shape[0], 
                     ll, message_length, **kwds)
 
@@ -624,7 +624,7 @@ class GaussianMixture(object):
 
             p_I = I + delta_I
 
-            assert weight.size < 25 or np.all(delta_I < 0), "Found the end?"
+            #assert weight.size < 25 or np.all(delta_I < 0), "Found the end?"
 
         else:
             p_I = None
@@ -803,8 +803,12 @@ class GaussianMixture(object):
         x, y = (Ku, sldcu/Ku)
 
         p0 = self._state_meta.get("_predict_slogdetcovs_p0", [y[0], 0.5, 0])
-        p_opt, p_cov = op.curve_fit(_approximate_ldc_per_mixture_component, x, y,
-            p0=p0, maxfev=100000, sigma=x.astype(float)**-1)
+        try:
+            p_opt, p_cov = op.curve_fit(_approximate_ldc_per_mixture_component,
+                x, y, p0=p0, maxfev=100000, sigma=x.astype(float)**-2)
+
+        except RuntimeError:
+            return (None, None, None)
 
         # Save optimized point for next iteration.
         self._state_meta["_predict_slogdetcovs_p0"] = p_opt

### 2018-03-14 gmmmml/visualize.py

@@ -37,10 +37,10 @@ class VisualizationHandler(object):
         self.axes = np.array(self.axes).flatten()
         self._display = True
 
-        self.axes[0].scatter(y.T[0], y.T[1], facecolor="k", alpha=0.5)
+        self.axes[0].scatter(y.T[0], y.T[1], facecolor="k", s=1, alpha=0.5)
 
-        self.axes[0].set_xlabel("Data X")
-        self.axes[0].set_ylabel("Data Y")
+        self.axes[0].set_xlabel(r"$x_0$")
+        self.axes[0].set_ylabel(r"$x_1$")
 
         self.axes[1].set_xlabel("E-M iteration")
         self.axes[1].set_ylabel("I_actual")
@@ -169,28 +169,28 @@ class VisualizationHandler(object):
                 # Show 2 standard deviations
                 width, height = 2 * 2 * np.sqrt(vals)
                 ellip = Ellipse(xy=mean, width=width, height=height, angle=theta,
-                    facecolor="r", alpha=0.5)
+                    facecolor="r", alpha=0.1)
 
                 self._model.append(self.axes[0].add_artist(ellip))
-                self._model.append(self.axes[0].scatter([mean[0]], [mean[1]], facecolor="r"))
+                self._model.append(self.axes[0].scatter([mean[0]], [mean[1]], facecolor="r", s=1))
 
             
             K = params["weight"].size
             slogdet_cov = np.sum(np.log(np.linalg.det(params["cov"])))
             log_mean_det_cov = np.log(np.mean(np.linalg.det(params["cov"])))
 
-            self.axes[3].scatter([K], [slogdet_cov], facecolor="k")
+            self.axes[3].scatter([K], [slogdet_cov], facecolor="k", s=1)
 
-            self.axes[4].scatter([K], [log_mean_det_cov], facecolor="k", alpha=0.5)
+            self.axes[4].scatter([K], [log_mean_det_cov], facecolor="k", alpha=0.5, s=1)
 
             sum_log_weights = np.sum(np.log(params["weight"]))
-            self.axes[5].scatter([K], [sum_log_weights], facecolor="k", alpha=0.5)
+            self.axes[5].scatter([K], [sum_log_weights], facecolor="k", alpha=0.5, s=1)
 
 
         elif kind == "expectation":
             self.axes[1].scatter(
                 [self._expectation_iter], [params["message_length"]],
-                facecolor="k")
+                facecolor="k", s=1)
[…]

### 2018-03-14 sandbox_mixture_search.py

@@ -3,18 +3,18 @@ import numpy as np
 from gmmmml import (mixture_search as mixture, visualize, utils)
 import  scipy.optimize as op
 
-np.random.seed(42)
+np.random.seed(11)
 
 
 for i in range(10):
 
-    y, labels, target, kwds = utils.generate_data(K=100)
+    y, labels, target, kwds = utils.generate_data(center_box=(-100, 100))
 
     visualization_handler = visualize.VisualizationHandler(
         y, target=target, figure_path="tmp/")
 
     search_model = mixture.GaussianMixture()
-    search_model.kmeans_search(y, kwds["centers"] + 10, 
+    search_model.kmeans_search(y, K_max=kwds["centers"] + 25, 
         visualization_handler=visualization_handler)
 
 

### 2018-03-14 .ci/build.sh
build paper setup

@@ -1,0 +1,7 @@
+#!/bin/bash -x
+
+echo "%%% This file is generated by the Makefile." > vc.tex
+git log -1 --date=short --format="format:\\newcommand{\\githash}{%h}\\newcommand{\\gitdate}{%ad}\\newcommand{\\gitauthor}{%an}" >> vc.tex
+
+# Build the paper
+tectonic article/ms.tex

### 2018-03-14 .ci/install.sh

@@ -1,0 +1,16 @@
+#!/bin/bash -x
+
+# SSL workaround
+mkdir -p ~/.config/Tectonic
+cp .ci/tectonic.config.toml ~/.config/Tectonic/config.toml
+
+# Conda install
+wget https://repo.continuum.io/miniconda/Miniconda3-latest-Linux-x86_64.sh -O miniconda.sh;
+bash miniconda.sh -b -p $HOME/miniconda
+export PATH="$HOME/miniconda/bin:$PATH"
+hash -r
+conda config --set always_yes yes --set changeps1 no
+conda update -q conda
+conda info -a
+conda install -c conda-forge -c pkgw-forge tectonic
+tectonic --help

### 2018-03-14 .ci/upload.sh

@@ -1,0 +1,11 @@
+#!/bin/bash -x
+
+# Push to GitHub
+if [ -n "$GITHUB_API_KEY" ]; then
+  cd $TRAVIS_BUILD_DIR
+  git checkout --orphan pdf
+  git rm -rf .
+  git add -f ms.pdf
+  git -c user.name='travis' -c user.email='travis' commit -m "building the paper"
+  git push -q -f https://andycasey:$GITHUB_API_KEY@github.com/andycasey/gmmmml pdf
+fi

### 2018-03-14 setup.py
update setup.py

@@ -19,7 +19,7 @@ def read(filename):
 # Get the version information.
 here = path.abspath(path.dirname(__file__))
 vre = re_compile("__version__ = \"(.*?)\"")
-version = vre.findall(read(path.join(here, "snob", "__init__.py")))[0]
+version = vre.findall(read(path.join(here, "gmmmml", "__init__.py")))[0]
 
 setup(
     name="gmmmml",
@@ -33,10 +33,7 @@ setup(
         "Intended Audience :: Science/Research",
         "License :: OSI Approved :: MIT License",
         "Operating System :: OS Independent",
-        "Programming Language :: Python :: 2.7",
-        "Programming Language :: Python :: 3.6",
-        "Topic :: Scientific/Engineering :: Astronomy",
-        "Topic :: Scientific/Engineering :: Physics"
+        "Programming Language :: Python :: 3.6"
     ],
     packages=find_packages(exclude=["tests"]),
     install_requires=["numpy", "scipy", "matplotlib", "sklearn", "six"],

### 2018-03-14 .ci/upload.sh
fix upload script

@@ -5,7 +5,7 @@ if [ -n "$GITHUB_API_KEY" ]; then
   cd $TRAVIS_BUILD_DIR
   git checkout --orphan pdf
   git rm -rf .
-  git add -f ms.pdf
+  git add -f article/ms.pdf
   git -c user.name='travis' -c user.email='travis' commit -m "building the paper"
   git push -q -f https://andycasey:$GITHUB_API_KEY@github.com/andycasey/gmmmml pdf
 fi

### 2018-03-14 gmmmml/mixture_search.py
in progress changes to approx max log-like; update mml prediction routine

@@ -75,7 +75,7 @@ def _total_parameters(K, D):
     :returns:
         The total number of model parameters, :math:`Q`.
     """
-    return (0.5 * D * (D + 3) * K) + (K - 1)
+    return (0.5 * D * (D + 3) * K) + K - 1
 
 
 def _generator_for_approximate_log_likelihood_improvement(K, log_likelihood,
@@ -265,98 +265,81 @@ def _approximate_bound_sum_log_determinate_covariances(target_K,
     return all_approx_slogdet_covs[keep]
 
 
-def _approximate_bounds_of_slogdet_cov_for_added_mixture(determinates):
-
-    determinates = np.atleast_1d(determinates)
-
-    #determinates = np.linalg.det(covariance_matrices)
-    current_K = determinates.size
-
-    # Two extremes: either the biggest one splits in ~half, or the smallest
-    # one will split in ~half
-
-    idx_min, idx_max = (np.argmin(determinates), np.argmax(determinates))
-
-    mask_min = np.ones(current_K, dtype=bool)
-    mask_max = np.ones(current_K, dtype=bool)
-
-    mask_min[idx_min] = False
-    mask_max[idx_max] = False
-
-    min_dets = np.hstack([determinates[mask_min], 
-        0.5 * determinates[idx_min], 0.5 * determinates[idx_min]])
-    max_dets = np.hstack([determinates[mask_max],
-        0.5 * determinates[idx_max], 0.5 * determinates[idx_max]])
-    min_slogdets = np.sum(np.log(min_dets))
-    max_slogdets = np.sum(np.log(max_dets))
-
-    return (min_slogdets, max_slogdets), (min_dets, max_dets)
 
 
+def _mixture_message_length(K, N, D, log_likelihood, slogdetcov, weights=None, 
+    yerr=0.001):
+    """
+    Estimate the message length of a gaussian mixture model.
 
+    :param K:
+        The number of mixtures. This can be an array.
 
+    :param N:
+        The number of data points.
+    
+    :param D:
+        The dimensionality of the data.
 
+    :param log_likelihood:
+        The estimated sum of the log likelihood of the future mixtures.
 
-def _approximate_slogdets(K, covariance_matrices, covariance_type="full"):
-    r"""
-    Approximate the sum of the log of the determinates of the covariance 
-    matrices for a mixture of :math:`K` gaussians based on the current estimate
-    of the covariance matrices.
+    :param slogdetcov:
[…]

### 2018-03-14 gmmmml/visualize.py

@@ -67,7 +67,7 @@ class VisualizationHandler(object):
                 
             target_kwds = dict(facecolor=self._color_target, s=50, alpha=0.5)
 
-            self.axes[1].scatter([K_target], [target["I"]], **target_kwds)
+            #self.axes[1].scatter([K_target], [target["I"]], **target_kwds)
 
             self.axes[5].scatter(
                 [K_target], [np.sum(np.log(target["weight"]))], **target_kwds)
@@ -274,12 +274,21 @@ class VisualizationHandler(object):
             K = params["K"]
             p_I = params["p_I"]
 
-            self._predict_message_lengths.extend([
-                self.axes[1].plot(K, p_I, c='b', alpha=0.5)[0]
-            ])
+            #self._predict_message_lengths.extend([
+            #    self.axes[1].plot(K, p_I, c='b', alpha=0.5)[0]
+            #])
 
             self.axes[1].relim()
             self.axes[1].autoscale_view()
+
+
+        elif kind == "show_mml":
+
+            K = params["K"]
+            I = params["I"]
+            I_parts = params["I_parts"]
+
+            self.axes[1].plot(K, I, c='g')
 
 
         else:

### 2018-03-14 gmmmml/mixture_search.py
show bounds for sum log w

@@ -167,28 +167,45 @@ def _approximate_log_likelihood_improvement(y, mu, cov, weight,
 
 
 
-def _approximate_sum_log_weights(target_K):
+def _bound_sum_log_weights(K, N):
     r"""
-    Return an approximate expectation of the function:
+    Return the analytical bounds of the function:
 
     .. math:
 
         \sum_{k=1}^{K}\log{w_k}
 
     Where :math:`K` is the number of mixtures, and :math:`w` is a multinomial
-    distribution. The approximating function is:
+    distribution. The bounded function for when :math:`w` are uniformly
+    distributed is:
 
     .. math:
 
-        \sum_{k=1}^{K}\log{w_k} \approx -K\log{K}
+        \sum_{k=1}^{K}\log{w_k} \lteq -K\log{K}
 
-    :param target_K:
+    and in the other extreme case, all the weight would be locked up in one
+    mixture, with the remaining :math:`w` values encapsulating the minimum
+    (physically realistic) weight of one data point. In that extreme case,
+    the bound becomes:
+
+    .. math:
+
+        \sum_{k=1}^{K}\log{w_k} \gteq log{(N + 1 - K)} - K\log{N}
+
+    :param K:
         The number of target Gaussian mixtures.
+
+    :param N:
+        The number of data points.
+
+    :returns:
+        The lower and upper bound on :math:`\sum_{k=1}^{K}\log{w_k}`.
     """
 
-    # TODO: Return a variance on this expectation, or approximate the constant
-    #       from previous values.
-    return -target_K * np.log(target_K)
+    upper = -K * np.log(K)
+    lower = np.log(N + 1 - K) - K * np.log(N)
+
+    return (lower, upper)
 
 
 
@@ -582,7 +599,7 @@ class GaussianMixture(object):
 
         target_K = np.atleast_1d(target_K)
 
-        p_slw, p_slw_err, p_slw_max = self._predict_slogweights(target_K)
+        p_slw, p_slw_err, p_slw_min, p_slw_max = self._predict_slogweights(target_K, N)
 
         # Predict log-likelihoods.
         p_ll, p_ll_err = self._predict_log_likelihoods(target_K)
@@ -621,7 +638,8 @@ class GaussianMixture(object):
         if visualization_handler is not None:
 
             visualization_handler.emit("predict_slw",
[…]

### 2018-03-14 gmmmml/visualize.py

@@ -215,6 +215,7 @@ class VisualizationHandler(object):
             p_slw = params["p_slw"]
             p_slw_err = params["p_slw_err"]
             p_slw_max = params["p_slw_max"]
+            p_slw_min = params["p_slw_min"]
 
             self._predict_slw.extend([
                 self.axes[5].plot(K, p_slw, 
@@ -222,8 +223,8 @@ class VisualizationHandler(object):
                 self.axes[5].fill_between(
                     K, p_slw_err[0] + p_slw, p_slw_err[1] + p_slw, 
                     facecolor=self._color_prediction, alpha=0.5, zorder=-1),
-                self.axes[5].plot(K, p_slw_max,
-                    c=self._color_prediction, linestyle="--", zorder=-1)[0]
+                self.axes[5].fill_between(K, p_slw_min, p_slw_max,
+                    facecolor="#666666", alpha=0.5, zorder=-1)
             ])
 
 

### 2018-03-14 gmmmml/mixture_search.py
minor tests

@@ -662,6 +662,8 @@ class GaussianMixture(object):
             I, I_parts = _mixture_message_length(self._state_K, N, D,
                 self._state_slog_likelihoods, slogdetcovs)
 
+            #I, I_parts = _mixture_message_length(target_K, N, D, p_ll, p_slw)
+
             visualization_handler.emit("show_mml", 
                 dict(K=self._state_K, I=I, I_parts=I_parts))
 

### 2018-03-14 sandbox_mixture_search.py

@@ -8,7 +8,19 @@ np.random.seed(11)
 
 for i in range(10):
 
-    y, labels, target, kwds = utils.generate_data(center_box=(-100, 100))
+    j = 0
+    while True:
+        print(i, j)
+        j += 1
+        try:
+            
+            y, labels, target, kwds = utils.generate_data(K=30, D=10, center_box=(-1000, 1000))
+
+        except:
+            continue
+
+        else:
+          break
 
     visualization_handler = visualize.VisualizationHandler(
         y, target=target, figure_path="tmp/")

### 2018-03-14 gmmmml/mixture_search.py
wip

@@ -78,95 +78,6 @@ def _total_parameters(K, D):
     return (0.5 * D * (D + 3) * K) + K - 1
 
 
-def _generator_for_approximate_log_likelihood_improvement(K, log_likelihood,
-    normalization_factor, *log_likelihoods_of_sequentially_increasing_mixtures):
-    
-    x = np.arange(K, K + len(log_likelihoods_of_sequentially_increasing_mixtures))
-    ll = np.hstack([
-        log_likelihood, 
-        log_likelihoods_of_sequentially_increasing_mixtures])
-    y = np.diff(ll) / normalization_factor
-
-    functions = {
-        1: lambda x, *p: p[0] / np.exp(x),
-        2: lambda x, *p: p[0] / np.exp(p[1] * x),
-        3: lambda x, *p: p[0] / np.exp(p[1] * x) + p[2]
-    }
-
-    foo = False
-    if x.size > 3:
-        #x = x[1:]
-        #y = y[1:]
-        foo = True
-
-    p_opt_ = []
-    for d in np.arange(1, 1 + x.size)[::-1]:
-
-        cost_function = functions.get(d, functions[3])
-        p0 = np.ones(d)
-
-        try:
-            p_opt, p_cov = op.curve_fit(cost_function, x, y, p0=p0)
-        except:
-            assert d > 1
-            continue
-        else:
-            p_opt_.append(p_opt)
-            break
-
-    p_opt = p_opt_.pop(0)
-    d = p_opt.size
-    cost_function = functions.get(d, functions[3])
-
-    generating_function = lambda target_K: log_likelihood \
-        + cost_function(np.arange(1, target_K), *p_opt).sum() * normalization_factor
-
-    if False and foo and x.size > 10:
-        import matplotlib.pyplot as plt
-        fig, ax = plt.subplots()
-        ax.scatter(np.arange(ll.size) + 1, ll)
-        ax.plot(np.arange(ll.size) + 1, [generating_function(_) for _ in np.arange(ll.size) + 1], c='r')
-        raise a
-
-    return generating_function
-
-
-
-def _approximate_log_likelihood_improvement(y, mu, cov, weight, 
-    log_likelihood, *log_likelihoods_of_sequentially_increasing_mixtures):
-    """
-    Return a function that will approximate the log-likelihood value for a
-    given target number of mixtures.
-    """
-
-    # Evaluate the current mixture.
-    K = weight.size
-    evaluate_f1 = np.sum(weight * np.vstack(
[…]

### 2018-03-14 gmmmml/visualize.py

@@ -246,6 +246,29 @@ class VisualizationHandler(object):
             #    ])
 
 
+        elif kind == "predict_ll2":
+
+            K = params["K"]
+            p_ll = params["p_ll"]
+
+            self.axes[6].plot(K, p_ll, c="g", alpha=0.5, zorder=-1)
+
+        elif kind == "predict_ll3":
+
+            K = params["K"]
+            p_ll = params["p_ll"]
+
+            self.axes[6].plot(K, p_ll, c="r", alpha=0.25, zorder=0)
+
+
+        elif kind == "predict_ll4":
+
+            K = params["K"]
+            p_ll = params["p_ll"]
+            
+            self.axes[6].plot(K, p_ll, c="m", alpha=0.25, zorder=0)
+
+
         elif kind == "predict_slogdetcov":
 
             self._update_previous_predict_slogdetcovs()
@@ -255,7 +278,6 @@ class VisualizationHandler(object):
             p_slogdetcovs_pos_err = params["p_slogdetcovs_pos_err"]
             p_slogdetcovs_neg_err = params["p_slogdetcovs_neg_err"]
 
-            """
             self._predict_slogdetcovs.extend([
                 self.axes[3].plot(K, p_slogdetcovs, c=self._color_prediction)[0],
                 self.axes[3].fill_between(K,
@@ -266,7 +288,6 @@ class VisualizationHandler(object):
 
             self.axes[3].relim()
             self.axes[3].autoscale_view()
-            """
 
         elif kind == "predict_message_length":
 

### 2018-03-15 gmmmml/mixture_search.py
(reasonable) analytic predictions of the log likelihood for future mixtures

@@ -446,7 +446,6 @@ class GaussianMixture(object):
             x_squared_norms = row_norms(y, squared=True)
             mu = cluster.k_means_._k_init(y, K, x_squared_norms=x_squared_norms)
             """
-
             
             # estimate covariance matrices.
             cov = _estimate_covariance_matrix_full(y, responsibility, mu)
@@ -459,7 +458,6 @@ class GaussianMixture(object):
 
             except ValueError:
                 logger.exception("Failed to calculate E-step")
-                raise 
                 continue
 
 
@@ -602,10 +600,10 @@ class GaussianMixture(object):
                     p_slw_min=p_slw_min),
                 save=p_ll is None and p_slogdetcovs is None and p_I is None)
 
-            if p_ll is not None:
-                visualization_handler.emit("predict_ll",
-                    dict(K=target_K, p_ll=p_ll, p_ll_err=p_ll_err),
-                    save=p_slogdetcovs is None)
+            #if p_ll is not None:
+            #    visualization_handler.emit("predict_ll",
+            #        dict(K=target_K, p_ll=p_ll, p_ll_err=p_ll_err),
+            #        save=p_slogdetcovs is None)
 
             # Sometimes we don't predict the sum of the log of the determinants
             # of the covariance matrices.
@@ -631,22 +629,23 @@ class GaussianMixture(object):
             p_ll = _approximate_log_likelihood(self._state_K, N, D, 
                 [np.log(ea) for ea in self._state_det_covs])
 
-            visualization_handler.emit("predict_ll2",
-                dict(K=self._state_K, p_ll=p_ll, save=True))
+            is_last_update = not (p_slogdetcovs is not None and p_slw is not None)
+            visualization_handler.emit("predict_ll",
+                dict(K=self._state_K, p_ll=p_ll, save=is_last_update,
+                    plotting_kwds=dict(c="r")))
 
 
-            if p_slogdetcovs is not None and p_slw is not None:
-
-
+            if not is_last_update:
 
                 p_ll = _approximate_log_likelihood(target_K, N, D,
-                    p_slogdetcovs/target_K, p_slw/target_K)
+                    p_slogdetcovs/target_K, np.exp(p_slw/target_K))
 
                 if target_K[0] > 20:
                     assert np.all(np.isfinite(p_ll))
 
-                visualization_handler.emit("predict_ll3",
-                    dict(K=target_K, p_ll=p_ll), save=True)
[…]

### 2018-03-15 gmmmml/visualize.py

@@ -145,9 +145,11 @@ class VisualizationHandler(object):
 
 
 
-    def emit(self, kind, params, save=True):
+    def emit(self, kind, params, save=True, clear_previous=True,
+        plotting_kwds=None):
 
-        
+        plotting_kwds = dict() if plotting_kwds is None else plotting_kwds
+
         if kind == "model":
 
             self._clear_model()
@@ -230,43 +232,26 @@ class VisualizationHandler(object):
 
         elif kind == "predict_ll":
 
-            self._update_previous_predict_lls()
+            if clear_previous:
+                self._update_previous_predict_lls()
 
             K = params["K"]
             p_ll = params["p_ll"]#/self._reference_ll
-            p_ll_err = params["p_ll_err"]#/self._reference_ll
+            p_ll_err = params.get("p_ll_err", None)#/self._reference_ll
 
-            self._predict_ll.extend([
-                self.axes[6].plot(K, p_ll,
-                    c=self._color_prediction, zorder=-1)[0]
-            ])
+            kwds = dict(c=self._color_prediction, zorder=-1)
+            kwds.update(plotting_kwds)
+
+            self._predict_ll.extend(self.axes[6].plot(K, p_ll, **kwds))
+
             #    self.axes[6].fill_between(
             #        K, p_ll_err[0] + p_ll, p_ll_err[1] + p_ll,
             #        facecolor=self._color_prediction, alpha=0.5, zorder=-1)
             #    ])
 
+            self.axes[6].relim()
+            self.axes[6].autoscale_view()
 
-        elif kind == "predict_ll2":
-
-            K = params["K"]
-            p_ll = params["p_ll"]
-
-            self.axes[6].plot(K, p_ll, c="g", alpha=0.5, zorder=-1)
-
-        elif kind == "predict_ll3":
-
-            K = params["K"]
-            p_ll = params["p_ll"]
-
-            self.axes[6].plot(K, p_ll, c="r", alpha=0.25, zorder=0)
-
-
-        elif kind == "predict_ll4":
-
-            K = params["K"]
-            p_ll = params["p_ll"]
-            
-            self.axes[6].plot(K, p_ll, c="m", alpha=0.25, zorder=0)
 
 
         elif kind == "predict_slogdetcov":

### 2018-04-18 gmmmml/gmm.py
add limits etc

@@ -6,6 +6,7 @@ Model data with a mixture of gaussians.
 import logging
 import numpy as np
 import scipy
+from scipy.spatial.distance import pdist
 
 from . import (em, mml)
 
@@ -132,7 +133,7 @@ class GaussianMixture(object):
 
         N, D = y.shape
         K_max = N if K_max is None else K_max
-        K_predict = kwds.pop("K_predict", 25)
+        K_predict = kwds.pop("K_predict", 50)
 
         visualization_handler = kwargs.get("visualization_handler", None)
             
@@ -163,13 +164,15 @@ class GaussianMixture(object):
 
             # Make predictions for past and future mixtures.
             K_target = np.arange(1, weight.size + K_predict)
-            predictions = self._predict_message_length(K_target, N, D, **kwds)
+            predictions = self._predict_message_length(K_target, y, **kwds)
 
             if visualization_handler is not None:
                 visualization_handler.emit("model", 
                     dict(mu=mu, cov=cov, weight=weight, nll=-np.sum(ll), I=I))
                 visualization_handler.emit("prediction", predictions)
 
+            if K >= 8:
+                raise a
         raise a
 
 
@@ -204,7 +207,7 @@ class GaussianMixture(object):
         return None
 
 
-    def _predict_message_length(self, K, N, D, **kwargs):
+    def _predict_message_length(self, K, y, **kwargs):
         """
         Predict the message length of past or future mixtures.
 
@@ -212,12 +215,17 @@ class GaussianMixture(object):
             An array-like object of the :math:`K`-th mixtures to predict the
             message elngths of.
 
-        :param N:
-            The number of data points.
-
-        :param D:
-            The dimensionality of the data points.
+        :param y:
+            The data :math:`y`.
         """
+
+        N, D = y.shape
+
+        mmpwd = self._state_meta.get("min_mean_pairwise_distance", None)
+
+        if mmpwd is None:
+            mmpwd = np.min(pdist(y))/(2.0 * D)
+            self._state_meta["min_mean_pairwise_distance"] = mmpwd
 
         predictions, meta = mml.predict_message_length(K, N, D, 
             previous_states=(
@@ -225,7 +233,8 @@ class GaussianMixture(object):
                 self._state_sum_log_weights,
                 self._state_det_covs,
                 self._state_sum_log_likelihoods),
-            state_meta=self._state_meta)
+            state_meta=self._state_meta,
+            min_mean_pairwise_distance=mmpwd)
 
[…]

### 2018-04-18 gmmmml/mml.py

@@ -281,8 +281,156 @@ def predict_sum_log_weights(K, N, previous_states=None):
         target_upper, fraction, fraction_err)
 
 
+def bounds_of_sum_log_det_covs(K, N, log_max_det, log_min_det, **kwargs):
+    r"""
+    Predict the theoretical bounds on the sum of the log of the determinant of
+    the covariance matrices for future (target) mixtures with :math:`K`
+    components.
 
-def predict_sum_log_det_covs(K, previous_states, draws=100, **kwargs):
+    :param K:
+        The number of target Gaussian mixtures.
+
+    :param N:
+        The number of data points.
+
+    :param log_max_det:
+        The logarithm of the determinant of the covariance matrix for the
+        :math:`K = 1` mixture, which represents the largest amount of 
+        (co-)variance that needs to be captured by the model.
+
+    :param log_min_det:
+        An estimate of the logarithm of the minimum determinant of a covariance
+        matrix for some mixture, as determined by the smallest pair-wise
+        distance between two points.
+
+        # TODO improve docs here.
+
+    """
+
+    # This is the limiting case when the log_max_det must be ~uniformly
+    # distributed over K mixtures. This represents the *lower* bound on the
+    # information required to encode it (or the upper bound on the sum of the
+    # log of the determinant of the covariance matrices).
+    upper_bound = K * log_max_det - K * np.log(K)
+
+    # This is the limiting case when every K-th mixture has a very small
+    # covariance matrix. 
+    lower_bound = K * log_min_det
+
+    return (lower_bound, upper_bound)
+
+
+def predict_sum_log_det_covs(K, N, previous_states, draws=100, **kwargs):
+    r"""
+    Predict the sum of the log of the determinant of covariance matrices for
+    future (target) mixtures with :math:`K` components.
+
+    :param K:
+        The number of target Gaussian mixtures.
+
+    :param N:
+        The number of data points.
+
+    :param previous_states:
+        A two-length tuple containing the previous :math:`K` trials, and the
+        determinant of the covariance matrices for each of those :math:`K`
+        trials.
+
+    :param draws: [optional]
+        The number of error draws to make when predicting the sum of the log
[…]

### 2018-04-18 gmmmml/visualize_new.py

@@ -310,6 +310,16 @@ class VisualizationHandler(object):
                     ax.axvline(params["K"][index], lw=1, linestyle=":", c=color))
 
             _rescale_based_on_data(ax, params["K"], params["p_I"])
+            
+            ylim = ax.get_ylim()
+            self._plot_prediction_items.extend([
+                ax.fill_between(params["K"], params["t_I_lower"], ylim[1],
+                    facecolor="#CCCCCC", zorder=-10),
+                ax.plot(params["K"], params["t_I_lower"],
+                        c="#666666", zorder=-5)[0]
+            ])
+            _rescale_based_on_data(
+                ax, params["K"], [ylim[1], np.min(params["t_I_lower"])])
 
 
         else:
@@ -345,11 +355,32 @@ class VisualizationHandler(object):
 
 
 def _rescale_based_on_data(ax, x, y, y_percent_edge=5):
+    """
+    Re-scale a figure axis based on the given data.
 
-    ax.set_xlim(np.min(x) - 0.5, np.max(x) + 0.5)
-    y_ptp = np.ptp(y)
-    ax.set_ylim(
-        np.min(y) - y_percent_edge/100.0 * y_ptp,
-        np.max(y) + y_percent_edge/100.0 * y_ptp
-    )
+    :param ax:
+        The figure axes.
 
+    :param x:
+        An array-like object of the x-values.
+
+    :param y:
+        An array-like object of the y-values.
+
+    :param y_percent_edge: [optional]
+        The percentage of the peak-to-peak y-range to show either end of the
+        range of `y` values.
+    """
+
+    if np.sum(np.isfinite(np.unique(x))) > 1:
+        ax.set_xlim(np.nanmin(x) - 0.5, np.nanmax(x) + 0.5)
+
+    y_finite = np.array(y)[np.isfinite(y)]
+    if y_finite.size > 1:
+        y_ptp = np.ptp(y_finite)
+        ax.set_ylim(
+            np.min(y_finite) - y_percent_edge/100.0 * y_ptp,
+            np.max(y_finite) + y_percent_edge/100.0 * y_ptp
+        )
+
+    return None

### 2018-04-18 sandbox_gmm.py

@@ -31,7 +31,7 @@ for i in range(10):
 from sklearn.datasets import make_blobs
 
 y, assignments = make_blobs(
-  n_samples=10000, n_features=2, centers=100, random_state=42)
+  n_samples=1000, n_features=2, centers=10, random_state=42)
 #y = np.vstack([y, np.random.uniform(-10, 10, size=(10, y.shape[1]))])
 
 #n_samples=100, n_features=2, centers=3, cluster_std=1.0, center_box=(-10.0, 10.0), shuffle=True, random_state=Non

### 2018-04-18 gmmmml/gmm.py
whoa

@@ -70,6 +70,7 @@ class GaussianMixture(object):
         self._state_weights = []
         self._state_sum_log_weights = []
         self._state_sum_log_likelihoods = []
+        self._state_message_lengths = []
 
         self._state_meta = {}
 
@@ -160,7 +161,7 @@ class GaussianMixture(object):
                 if self.threshold > abs(change):
                     break
 
-            self._record_state(cov, weight, ll)
+            self._record_state(cov, weight, ll, I)
 
             # Make predictions for past and future mixtures.
             K_target = np.arange(1, weight.size + K_predict)
@@ -177,7 +178,7 @@ class GaussianMixture(object):
 
 
 
-    def _record_state(self, cov, weight, log_likelihood):
+    def _record_state(self, cov, weight, log_likelihood, message_length):
         r"""
         Record the state of the model to make better predictions for the
         message lengths of future mixtures.
@@ -190,6 +191,9 @@ class GaussianMixture(object):
 
         :param log_likelihood:
             The log-likelihood of the current mixture.
+
+        :param message_length:
+            The length of the message required to encode the current state.
         """
 
         self._state_K.append(weight.size)
@@ -203,6 +207,9 @@ class GaussianMixture(object):
 
         # Record log-likelihood.
         self._state_sum_log_likelihoods.append(np.sum(log_likelihood))
+
+        # Record message length.
+        self._state_message_lengths.append(message_length)
 
         return None
 
@@ -240,5 +247,14 @@ class GaussianMixture(object):
         # faster (e.g., the optimization functions start from values closer to
         # the true value).
         self._state_meta.update(meta)
+
+        # Say something about K_max.
+        indices = np.where(
+            predictions["t_I_lower"] >= np.min(self._state_message_lengths))[0]
+
+        if any(indices):
+            print("K_true < {0:.0f}".format(1 + K[indices[0]]))
+            
+
 
         return predictions

### 2018-04-18 gmmmml/mml.py

@@ -720,4 +720,5 @@ def predict_message_length(K, N, D, previous_states, yerr=0.001,
         ("t_I_lower", t_I_lower)
     ])
 
+
     return (predictions, update_meta)

### 2018-04-18 sandbox_gmm.py

@@ -31,7 +31,7 @@ for i in range(10):
 from sklearn.datasets import make_blobs
 
 y, assignments = make_blobs(
-  n_samples=1000, n_features=2, centers=10, random_state=42)
+  n_samples=1000, n_features=2, centers=30, random_state=42)
 #y = np.vstack([y, np.random.uniform(-10, 10, size=(10, y.shape[1]))])
 
 #n_samples=100, n_features=2, centers=3, cluster_std=1.0, center_box=(-10.0, 10.0), shuffle=True, random_state=Non

### 2018-04-18 gmmmml/gmm.py
foo

@@ -172,8 +172,7 @@ class GaussianMixture(object):
                     dict(mu=mu, cov=cov, weight=weight, nll=-np.sum(ll), I=I))
                 visualization_handler.emit("prediction", predictions)
 
-            if K >= 8:
-                raise a
+
         raise a
 
 

### 2018-04-18 gmmmml/mml.py

@@ -392,38 +392,6 @@ def predict_sum_log_det_covs(K, N, previous_states, draws=100, **kwargs):
 
     return (p_sldc, (lower_sum_log_det_cov, upper_sum_log_det_cov), {})
 
-    """
-
-    # Predict the log determinants of covariance matrices based on a Gaussian
-    # KDE.
-    kernel = gaussian_kde(np.log(previous_det_cov[-1]))
-
-    # If we are assuming that those determinants are representative, then we
-    # can do one big resample.
-    log_det_covs = kernel.resample(np.sum(K)).flatten()
-
-    if len(previous_det_cov) > 25:
-
-        import matplotlib.pyplot as plt
-        fig, ax = plt.subplots()
-        ax.hist(np.log(previous_det_cov[-1]), normed=True)
-        ax.hist(log_det_covs, edgecolor="b", alpha=0.5, normed=True)
-
-        raise a
-
-    # Calculate the predicted sum of the log of the determinant of the 
-    # covariance matrices for each target K.
-    offset = 0
-    p_sldc = np.zeros(K.size, dtype=float)
-    p_sldc_err = K * np.std(log_det_covs)
-
-    for i, k in enumerate(K):
-        print(i, k, offset)
-        p_sldc[i] = np.sum(log_det_covs[offset:offset + k + 1])
-        offset += k
-
-    return (p_sldc, (p_sldc_err, -p_sldc_err), {})
-    """
 
 
 
@@ -650,7 +618,6 @@ def predict_message_length(K, N, D, previous_states, yerr=0.001,
         K, N, log_max_det, log_min_det)
 
 
-
     p_slogdetcov, p_slogdetcov_err, update_meta = _deprecated_predict_sum_log_det_covs(
         K, N=N, previous_states=(_state_K, _state_det_covs), **state_meta)
 
@@ -685,8 +652,8 @@ def predict_message_length(K, N, D, previous_states, yerr=0.001,
     p_slogdetcov_pos_err, p_slogdetcov_neg_err = p_slogdetcov_err
     p_I_slogdetcov_pos_err = sldc_scalar * p_slogdetcov_pos_err
     p_I_slogdetcov_neg_err = sldc_scalar * p_slogdetcov_neg_err
-    t_I_slogdetcov_lower = sldc_scalar * t_slogdetcov_lower
-    t_I_slogdetcov_upper = sldc_scalar * t_slogdetcov_upper
[…]

### 2018-04-18 gmmmml/visualize.py

@@ -1,4 +1,7 @@
 
+"""
+Visualize the search progress for a gaussian mixture model.
+"""
 
 import os
 import numpy as np
@@ -6,548 +9,404 @@ import matplotlib.pyplot as plt
 
 from matplotlib.patches import Ellipse
 from matplotlib.ticker import MaxNLocator
-import scipy.special
 
-def _total_parameters(K, D):
-    r"""
-    Return the total number of model parameters :math:`Q`, if a full 
-    covariance matrix structure is assumed.
-
-    .. math:
-
-        Q = \frac{K}{2}\left[D(D+3) + 2\right] - 1
-
-
-    :param K:
-        The number of Gaussian mixtures.
-
-    :param D:
-        The dimensionality of the data.
-
-    :returns:
-        The total number of model parameters, :math:`Q`.
-    """
-    return (0.5 * D * (D + 3) * K) + K - 1
-
-
-
-def _mixture_message_length_parts(K, N, D):
-
-    Q = _total_parameters(K, D)
-
-    I_mixtures = K * np.log(2) * (1 - D/2.0) + scipy.special.gammaln(K) \
-        + 0.25 * (2.0 * (K - 1) + K * D * (D + 3)) * np.log(N)
-    I_parameters = 0.5 * np.log(Q * np.pi) - 0.5 * Q * np.log(2 * np.pi)
-
-    return I_mixtures + I_parameters
+from . import mml
 
 
 class VisualizationHandler(object):
 
-    def __init__(self, y, target=None, x_index=0, y_index=1,
-        figure_path=None, **kwargs):
+    def __init__(self, y, data_projection_indices=(0, 1), figure_path=None,
+        target=None, **kwargs):
+        r"""
+        Initialize a visualisation handler to show the search progress and the
+        predictions of future mixtures.
 
-        figure_path = "" if figure_path is None else figure_path
+        :param y:
+            The data :math:`y`.
 
-        self._color_model = "r"
-        self._color_prediction = "b"
-        self._color_target = "g"
+        :param data_projection_indices: [optional]
+            A two-length tuple containing the indices to use when plotting the
+            :math:`x` and :math:`y` projections of the data.
 
-        self._model = []
-        self._expectation_iter = 1
+        :param figure_path: [optional]
+            The local path to store any figures generated during the search.
+
+        :param target: [optional]
+            The target message length components (e.g., for generated data).
[…]

### 2018-04-18 gmmmml/visualize_new.py

@@ -1,386 +1,0 @@
-
-"""
-Visualize the search progress for a gaussian mixture model.
-"""
-
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-
-from matplotlib.patches import Ellipse
-from matplotlib.ticker import MaxNLocator
-
-from . import mml
-
-
-class VisualizationHandler(object):
-
-    def __init__(self, y, data_projection_indices=(0, 1), figure_path=None,
-        target=None, **kwargs):
-        r"""
-        Initialize a visualisation handler to show the search progress and the
-        predictions of future mixtures.
-
-        :param y:
-            The data :math:`y`.
-
-        :param data_projection_indices: [optional]
-            A two-length tuple containing the indices to use when plotting the
-            :math:`x` and :math:`y` projections of the data.
-
-        :param figure_path: [optional]
-            The local path to store any figures generated during the search.
-
-        :param target: [optional]
-            The target message length components (e.g., for generated data).
-        """
-
-        self._y_shape = y.shape
-        self._data_projection_indices = data_projection_indices
-
-        self._figure_path = "" if figure_path is None else figure_path
-        self._figure_iter = 1
-        self._figure_prefix = os.path.join(self._figure_path,
-            "iter_{:.0f}".format(np.random.uniform(0,  10000)))
-
-        self._fig, axes = plt.subplots(1, 5, figsize=(15.6 + 2.8, 2.8))
-            
-        self._init_axes(y, target)
-        self.snapshot()
-
-        # lists for plotting
-        self._plot_model_items = []
-        self._plot_prediction_items = []
-
-
-
-        return None
-
-
-    def _init_axes(self, y, target=None):
-        """
-        Initialize the axes in the figure.
-
-        :param y:
-            The data :math:`y`.
-
-        :param target:
-            The target distribution (e.g.,  in the case of generated data).
-        """
-
-        scatter_data_kwds = dict(facecolor="k", s=1)
-        scatter_progress_kwds = scatter_data_kwds.copy()
-        scatter_progress_kwds.update(dict(s=5))
-
-        x_index, y_index = self._data_projection_indices
-
-        ax = self._ax("data")
-        ax.scatter(y.T[x_index], y.T[y_index], **scatter_data_kwds)
-        ax.set_xlabel(r"$x_{{{0}}}$".format(x_index))
[…]

### 2018-04-18 sandbox_gmm.py

@@ -31,7 +31,7 @@ for i in range(10):
 from sklearn.datasets import make_blobs
 
 y, assignments = make_blobs(
-  n_samples=1000, n_features=2, centers=30, random_state=42)
+  n_samples=1000, n_features=2, centers=500, random_state=4234)
 #y = np.vstack([y, np.random.uniform(-10, 10, size=(10, y.shape[1]))])
 
 #n_samples=100, n_features=2, centers=3, cluster_std=1.0, center_box=(-10.0, 10.0), shuffle=True, random_state=Non

### 2018-04-18 gmmmml/visualize.py
things

@@ -299,7 +299,7 @@ class VisualizationHandler(object):
                     **theory_plot_kwds)[0],
                 ax.plot(
                     params["K"], params["t_I_slogdetcov_upper"],
-                    linestyle=".-", **theory_plot_kwds)[0]
+                    linestyle="-.", **theory_plot_kwds)[0]
                 ])
             _rescale_based_on_data(ax, params["K"], 
                 np.hstack([

### 2018-04-18 sandbox_gmm.py

@@ -1,6 +1,6 @@
 
 import numpy as np
-from gmmmml import (gmm as mixture, visualize_new as visualize, utils)
+from gmmmml import (gmm as mixture, visualize, utils)
 import  scipy.optimize as op
 
 np.random.seed(11)
@@ -31,7 +31,7 @@ for i in range(10):
 from sklearn.datasets import make_blobs
 
 y, assignments = make_blobs(
-  n_samples=1000, n_features=2, centers=500, random_state=4234)
+  n_samples=10000, n_features=2, centers=43, random_state=11)
 #y = np.vstack([y, np.random.uniform(-10, 10, size=(10, y.shape[1]))])
 
 #n_samples=100, n_features=2, centers=3, cluster_std=1.0, center_box=(-10.0, 10.0), shuffle=True, random_state=Non

### 2018-04-19 article/ms.tex
fix upload script
foo
lots of text updates

@@
-%% This document is part of the snob project
-%% Copyright 2017 the authors. All rights reserved.
 \documentclass{elsarticle}
@@
 \usepackage{algpseudocode}
+
 % For revision history
@@
 \newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
+\renewcommand{\vec}[1]{\vect{#1}}
 
@@
 %       \title{Searching for the Origins of the Galaxy with Minimum Message Length mixture modelling}
-        \title{Reconstructing the Origins of the Galaxy with Minimum Message Length mixture modelling and new search techniques}
+%        \title{Reconstructing the Origins of the Galaxy with Minimum Message Length mixture modelling and new search techniques}
+        \title{Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy}
 
 %% Group authors per affiliation:
-\author[moca]{Andy Casey}
-\author[fit]{David L. Dowe}
 \author[fit]{Aldeida Aleti\corref{mycorrespondingauthor}}
@@
 \ead{aldeida.aleti@monash.edu}
+\author[moca,fit]{Andy Casey}
 \author[moca]{John Lattanzio}
+\author[fit]{David L. Dowe}
 \address[moca]{School of Physics and Astronomy, Monash University, Melbourne, Clayton VIC 3800, Australia}
@@
 % https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
-Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history of our Galaxy by identifying the remains of a) ancient star formation events and b) smaller galaxies accreted by our own Galaxy during its youth.
+Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
+of our Galaxy by identifying ancient star formation events. Most stars are 
+born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
+share the same chemical composition of the gas cloud that they form from.
+Stars in clusters often do not remain gravitationally bound, and disperse 
+throughout the disk of our Galaxy within about $10^9\,{\rm yr}$. %citation needed.
+As a consequence, where a star is now does not reflect its formation site, 
+complicating our understanding of galaxy formation and evolution. However, the
+observable chemical composition of a star remains largely the same throughout
+that star's lifetime, providing an immutable marker (a chemical tag, or
+`fingerprint') that carries with it information about where and when star
+formed.
+
+Observations reveal that stars in surviving clusters show very little scatter
+in their chemical composition~\cite{pancino2010chemical}, and that star
+clusters can be identified solely on the basis of stellar chemical
+abundances~\cite{hogg2016kmeans}. Astronomers have precisely measured the
+chemical composition for nearly one million stars in the disk of our galaxy,
+where the natal formation site for most stars is not known~\cite{galah}.
+Indeed, from those data it is not known how many star formation sites we might
+expect to be able to detect, and we do not know the number of stars that we
+might to observe from each cluster (the so-called stellar mass function).
+Theoretical simulations can provide a guide to these quantities (e.g., of the
+order X star forming sites that have total masses less than Y), but the
+predictions have a large variance.
+
+These facts conspire to produce a challenging parameter estimation and model
+selection problem that is analagous to mixture modelling, where the true number 
+of mixtures is unknown. Little work has been done to address this problem so 
+far, apart from the work of \cite{mitschang2014quantitative}, who used the 
+\emph{Manhattan distance} metric to quantify the chemical difference between 
+any two stars, and estimate the level of homogeneity expected within a cluster. 
+This metric calculates the mean weighted absolute difference of abundance 
+levels, $\sigma_C$, between two stars $S^a$ and $S^b$ as follows:
+% Given that $\sigma$ is so widely used to mean standard deviation, this notation 
+% (to follow) seems a bit unfortunate
+$\sigma_C = \sum_{c \in C}w_c|S^a_c-S^b_c|/||C||$, where $C$ is the set of 
+chemical species, $S^a_c$ is the chemical abundance of element $c$ in star $
+S^a$, $w_c$ is the weight given to species $c$, and $||C||$ is the total 
+number of chemical species\footnote{Here we describe the work of Mitschang et
+al. (2014)~\cite{mitschang2014quantitative} using their notation, but we 
+caution that we use similar notation in this work with very different 
+terminology and meaning.}. The Manhattan distance values are high for stars 
+from different clusters, and low for stars from the same cluster. Note that 
+this metric is not particularly robust or versatile; for example it is not 
+invariant even to a simple rotation of the co-ordinate axes.
+% - and nor does it take into account that a difference between two stars of (say) 0.01\% in a rare element might be
+% vastly more significant than the same difference in a vastly more abundant element
+% Retraction.  It does endeavour to take that into account, via the weights, $w_c$, but not by setting them all to 1.
+An important issue is how to set the weights for each chemical species $w_c$, 
+and how they affect the results (Mitschang et al.~\cite{mitschang2014quantitative} 
+used $w_c=1$). Another issue is setting the boundary between clusters. It is 
+not clear what a high and low Manhattan distance is; these values are subjective. 
+In this paper, we express the problem in terms of mixture modelling and infer 
+the number of clusters, their mixing proportions (or relative abundances) and 
+each cluster's statistical distribution using Minimum Message Length (MML) and 
+novel search techniques that are readily applicable to the large data sets
+being accumulated by astronomers worldwide.
+
+% Not clear how many star formation sites, etc...
+
+%and using that as a reflection from the gas cloud that those stars
+%formed. Through successive generations of star formation, each fusing light
+%elements into heavy ones and releasing them into the interstellar medium,
+%the chemical composition of a star-forming gas cloud reflects the lives and
+%deaths of previous generations of stars.
+
+% accreted by our own Galaxy during its youth.
 This is possible because stars form in clusters of 10s to 1000s at a time, all sharing the same composition, being that of the gas cloud from which they formed.
@@
 
-The idea of Galactic Archaeology relies on identifying stars born together based on their composition.
+%The idea of Galactic Archaeology relies on identifying stars born together based on their composition.
 Internal mechanisms change the abundance of the lighter elements, but the heavier ones remain mostly unchanged.
@@
 
-Observations have shown that each star cluster has a uniform and distinct chemical fingerprint~\cite{pancino2010chemical}. Our aim is to identify stars with the same composition, as these
+%Observations have shown that each star cluster has a uniform and distinct chemical fingerprint~\cite{pancino2010chemical}. Our aim is to identify stars with the same composition, as these
 % were
-are widely believed to have been
-almost certainly
-formed at the same time in the same cluster.
+%are widely believed to have been
+%almost certainly
+%formed at the same time in the same cluster.
 We expect this cluster has now dispersed, but the tell-tale chemical tags will be present and will allow us to identify the groups of stars that are the relics of ancient star forming activity.
 
-Little work  has been done to address this problem so far, apart from the work of \cite{mitschang2014quantitative}, who used the \emph{Manhattan distance} metric to  quantify the chemical difference between any two stars, and as a result estimate the level of homogeneity expected within a cluster.
+%Little work  has been done to address this problem so far, apart from the work of \cite{mitschang2014quantitative}, who used the \emph{Manhattan distance} metric to  quantify the chemical difference between any two stars, and as a result estimate the level of homogeneity expected within a cluster.
 This metric calculates the mean weighted absolute difference of abundance levels, $\sigma_C$, between two stars $S^a$ and $S^b$ as follows:
-% Given that $\sigma$ is so widely used to mean standard deviation, this notation (to follow) seems a bit unfortunate
-$\sigma_C = \sum_{c \in C}w_c|S^a_c-S^b_c|/||C||$, where $C$ is the set of chemical species, $S^a_c$ is the chemical abundance of element $c$ in star $S^a$, $w_c$ is the weight given to species $c$, 
+%% Given that $\sigma$ is so widely used to mean standard deviation, this notation (to follow) seems a bit unfortunate
+%$\sigma_C = \sum_{c \in C}w_c|S^a_c-S^b_c|/||C||$, where $C$ is the set of chemical species, $S^a_c$ is the chemical abundance of element $c$ in star $S^a$, $w_c$ is the weight given to species $c$, 
 %which is considered 1 for all elements in the current work, 
-and $||C||$ is the total number of chemical species.
+%and $||C||$ is the total number of chemical species.
 The Manhattan distance values are high for stars from different clusters, and low for stars from the same cluster.
@@
 % Retraction.  It does endeavour to take that into account, via the weights, $w_c$, but not by setting them all to 1.
-An important issue is how to set the weights for each chemical species $w_c$, and how they affect 
-the results (Mitschang et al.~\cite{mitschang2014quantitative} used $w_c=1$).
+%An important issue is how to set the weights for each chemical species $w_c$, and how they affect 
+%the results (Mitschang et al.~\cite{mitschang2014quantitative} used $w_c=1$).
 Another issue is setting the boundary between clusters.
 It is not clear what a high and low Manhattan distance is; at the moment
-these values are largely subjective.
+%these values are largely subjective.
 In this paper, we express the problem in terms of mixture modelling and infer the number of clusters, their mixing proportions (or relative abundances) and each cluster's statistical distribution using Minimum Message Length (MML) and two novel search techniques.
@@
 
-Mixture modelling is composed of three main parts: (i) an estimator of component parameters of a mixture, (ii) an objective functions that scores a mixture, and (iii) a search strategy which identifies the best number of components and their weights.
+\noindent{}Mixture modelling is composed of three main parts:
+\begin{itemize}
+    \item an estimator of component parameters of a mixture,
+    \item an objective functions that scores a mixture, and 
+    \item a search strategy which identifies the best number of components and
+          their relative weights.
+\end{itemize}
 
-Different criteria for estimating component parameters have been used in the literature, with traditional methods using maximum likelihood estimation (MLE)
+Different criteria for estimating component parameters have been used in the 
+literature, with traditional methods using maximum likelihood estimation (MLE)
 % Perhaps because ML could refer to many things (e.g., machine learning), maximum likelihood is often (I think usually) called MLE
 ~\cite{akaike1971determination} or maximum a posteriori probability (MAP)~\cite{shimony1994finding}.
-% I re-write the draft below.
-% Traditional methods, such as ML or MAP, have not been designed to consider the complexity of the model, but only focus on the goodness-of-fit.
+These methods do not consider the complexity of the model, but only focus on
+the goodness-of-fit, and require modification to minimise the risk of
+over-fitting. In this work, we focus on the Bayesian information-theoretic 
+Minimum Message Length (MML) principle~\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a, Wallace05}.
+Unlike MLE or MAP estimators, the MML principle is invariant under non-linar
+data transformations, and can seemlessly incorporate different probability
+distributions.
+
+%have not been designed to consider the complexity of the model, but only focus on the goodness-of-fit.
 They have to be modified with ad-hoc principles in order to ensure that the data is not over-fitted.
@@
 %, and different from MLE, can be used with different probability distributions.  
-In this work, we focus on the Bayesian information-theoretic Minimum Message Length (MML) principle~\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a, Wallace05}.
 %
-The MML principle states that the best explanation of the data is the one that leads to the shortest two-part message~\cite{Wallace05}, where a \textit{message} takes into account both the complexity of the model and its explanatory power (corresponding respectively to the lengths of the first and second parts of the message).
-MML has been shown to perform well on a variety of empirical analyses (see, e.g., \cite{viswanathan1999finding,fitzgibbon2004minimum}, with
-% further examples and
-references to further examples in \cite{Wallace05,dowe2007bayes,Dowe2008a,Dowe2011a}.
-Arguments about the statistical consistency
-(i.e.,~as the number of data points increases the distributions of the estimates become increasingly concentrated near the true value)
-of MML are given in
-\cite{DoweWallace1997a,Dowe2011a}.
-Unlike the rival Bayesian approach MAP, for continuous parameters MML is a statistically invariant (meaning that $1$-to-$1$ transformations of co-ordinates, such as polar $(r, \theta)$ to Cartesian $(x, y)$, leave the answer unchanged) Bayesian approach~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}.
+The classically-described princple of MML is that the best explanation of the
+data is the one that leads to the shortest so-called two-part message~\cite{Wallace05}, 
+where a \textit{message} takes into account both the complexity of the model 
+and its explanatory power. The complexity of the model is described through
+the first part of the message, and the second part of the message describes
+its explanatory power. The \emph{length} of each message part is quantified
+(or measured) using information theory, allowing for a fair evaluation between
+different models of varying complexity or explanatory power. MML has been 
+shown to perform well on a variety of empirical analyses (see, e.g., 
+\cite{viswanathan1999finding,fitzgibbon2004minimum}, with references to 
+further examples in \cite{Wallace05,dowe2007bayes,Dowe2008a,Dowe2011a}).
+
+% ARC: Not sure if this paragraph should be here...
+Arguments about the statistical consistency (i.e.,~as the number of data 
+points increases the distributions of the estimates become increasingly 
+concentrated near the true value) of MML are given in \cite{DoweWallace1997a,Dowe2011a}.
+The MML principle requires that we explicitly specify our prior beliefs on the
+model parameters, providing a Bayesian optimisation approach which can be
+applied across entire classes of models.
+
+%Unlike the rival Bayesian approach MAP, for continuous parameters MML is a statistically invariant (meaning that $1$-to-$1$ transformations of co-ordinates, such as polar $(r, \theta)$ to Cartesian $(x, y)$, leave the answer unchanged) Bayesian approach~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}.
 It allows the modelling of prior beliefs in the calculation of the best model parameters.
@@
 
-Two other well-known methods based on information theory are the Akaike Information Criterion (AIC)~\cite{akaike1974} and Bayesian Information Criterion
-(BIC)~\cite{schwarz1978estimating}, both of which endeavour to temper Maximum Likelihood's tendency to over-fit and under-estimate the noise.
-% Unlike Maximum Likelihood, MML takes into account model complexity - and, unlike MAP, it is invariant under non-linear continuous transformations.
-In MML terms, Maximum Likelihood can be thought of as MML without paying for the first part of the message.
-
-AIC~\cite{akaike1974} chooses a model 
-% that minimizes
-with the motivation of minimizing
-the Kullback-Leibler
-% distance
-divergence
-between
+Two other well-known methods based on information theory are the Akaike 
+Information Criterion (AIC)~\cite{akaike1974} and Bayesian Information Criterion
+(BIC)~\cite{schwarz1978estimating}, both of which endeavour to temper maximum 
+likelihood's tendency to over-fit and under-estimate the noise. Unlike maximum
+likelihood, MML accounts for model complexity, and for this reason maximum
+likelihood can be thought of as MML without including the first part of the
+message that describes the model complexity. AIC~\cite{akaike1974} chooses a 
+model with the motivation of minimizing the Kullback-Leibler divergence 
+between the model and the truth:
 % KL divergence is generally asymmetric
-the model and the truth, and is defined as:
-
 \begin{equation}
-AIC = -2 (\ln ( f(\vec{x} | \vec{\theta} )) + 2 k
+    AIC = -2 \log_{e}f(\vec{x} | \vec{\theta}) + 2 k
 \end{equation}
-where the likelihood $f(\vec{x} | \vec{\theta})$ is the probability of the data $\vec{x}$ given a model parameterised by $\vec{\theta}$,
-and $k$ is the number of free parameters in the model.
-
-BIC~\cite{schwarz1978estimating} uses a constant penalty equal to $(\log N)/2$ for each free parameter in the model,
-where $N$ is the number of data points.
-Minimum Description Length (MDL)~\cite{rissanen1978modeling} was independently introduced in the same year (1978) as BIC with a very similar formula, a decade after MML \cite{WallaceBoulton1968}.
-Both AIC and BIC/MDL associate parameter costs with the number of free parameters and not their values.
-In this case, all models of a particular probability distribution have the same cost, regardless of the respective means, covariance matrices and Fisher information,
-which may hinder accurate inference.
+\noindent{}where the likelihood $f(\vec{x} | \vec{\theta})$ is the probability 
+of the data $\vec{x}$ given a model parameterised by $\vec{\theta}$, and $k$ 
+is the number of free parameters in the model. BIC~\cite{schwarz1978estimating} 
+uses a constant penalty equal to $\frac{1}{2}\log{N}$ for each free parameter in the 
+model, where $N$ is the number of data points. Minimum Description Length (MDL)
+~\cite{rissanen1978modeling} is a similar approach to BIC~\cite{schwarz1978estimating},
+and has close (albeit also distinct) ties to MML \cite{WallaceBoulton1968}.
+AIC and BIC/MDL associate parameter costs (or model complexity) with the number 
+of free parameters, not the values of those parameters or the variance that
+can be explained by those parameters. As a result, when using AIC or BIC/MDL
+formalisms, all models of a particular probability distribution share the same
+cost, regardless of their respective means, covariance matrices, or the
+Fisher information, which could bias inferences.
 
@@
 % Let's not overly annoy the reviewers and readers.
-Examples of MML's promise against rival methods include comparisons with AIC~\cite{dowe2007bayes}, BIC \cite{agusta2002mml}, MDL~\cite{viswanathan1999finding} and MLE~\cite{dowe2007bayes} - and also in statistical econometric time series autoregression \cite{fitzgibbon2004minimum}, machine learning problems \cite{viswanathan1999finding},
-angular data \cite{WallaceDowe1993,DoweOliverWallace1996}
-and using this modelling \cite{WallaceDowe1993} to infer the order in which various parts of a protein fold~\cite{edgoose1998mml}. 
-%
-A further example is the problem of deciding whether or not some 1-dimensional data generated uniformly in the interval $[ 0, 1 ]$ has a gap in it.
+%Examples of MML's promise against rival methods include comparisons with AIC~\cite{dowe2007bayes}, BIC \cite{agusta2002mml}, MDL~\cite{viswanathan1999finding} and MLE~\cite{dowe2007bayes} - and also in statistical econometric time series autoregression \cite{fitzgibbon2004minimum}, machine learning problems \cite{viswanathan1999finding},
+%angular data \cite{WallaceDowe1993,DoweOliverWallace1996}
+%and using this modelling \cite{WallaceDowe1993} to infer the order in which various parts of a protein fold~\cite{edgoose1998mml}.
+
+%A further example is the problem of deciding whether or not some 1-dimensional data generated uniformly in the interval $[ 0, 1 ]$ has a gap in it.
 Curiously, even when there is clearly no gap, AIC infers that there is a gap, whereas MML gives the right answer \cite{dowe2007bayes}.
 
-Very importantly, the MML message structure enables us to robustly deal with missing data \cite{WallaceDowe1994b,WallaceDowe1997,WallaceDowe2000}.
+%Very importantly, the MML message structure enables us to robustly deal with missing data \cite{WallaceDowe1994b,WallaceDowe1997,WallaceDowe2000}.
 Other methods often replace missing values with the mean - in turn, reducing the true standard deviation - or various other imputed methods.
@@
 
-The parameters of a mixture model can be estimated using Expectation-Maximisation (EM)~\cite{dempster1977maximum}.
-The EM algorithm has been used with MML to infer Gaussian mixtures under different simplifying assumptions, such as assuming that the covariance matrices are diagonal~\cite{WallaceDowe1994b, WallaceDowe1997a, WallaceDowe2000} (where it is sometimes called an `adjust cycle') and \cite{oliver1996unsupervised}, or approximating the probabilities of mixture parameters~\cite{figueiredo2002unsupervised,roberts1998bayesian}.
+The parameters of a mixture model can be estimated using Expectation-Maximisation 
+(EM)~\cite{dempster1977maximum}. The EM algorithm has been used with MML to 
+infer Gaussian mixtures under different simplifying assumptions, such as 
+assuming that the covariance matrices are diagonal~\cite{WallaceDowe1994b, WallaceDowe1997a, WallaceDowe2000} 
+(where it is sometimes called an `adjust cycle') and \cite{oliver1996unsupervised},
+or approximating the probabilities of mixture parameters~\cite{figueiredo2002unsupervised,roberts1998bayesian}. 
+The search algorithm employed in the majority of these approaches is based on 
+running EM iteratively for different numbers of components~\cite{WallaceDowe1994b, 
+WallaceDowe1997a, WallaceDowe2000, oliver1996unsupervised,roberts1998bayesian,
+biernacki2000assessing}. The search method proposed by Figueiredo and Jain
+~\cite{figueiredo2002unsupervised}, on the other hand, starts with a large 
+number of components, and iteratively deletes components. A mirror extension 
+of this method was proposed by Kasarapu and Allison~\cite{kasarapu2015minimum}, 
+which starts with one components, and selectively splits, deletes, or merges 
+components in an attempt to minimise the message length. While this method 
+removes the overhead introduced by the search algorithm of Figueiredo and Jain
+~\cite{figueiredo2002unsupervised}, the search remains greedy, and can be 
+prone to premature convergence in suboptimal solutions. The search methods 
+introduced in this paper, incorporate a perturbation step which encourages 
+explorations of new local optima.
 
-The search algorithm employed in the majority of these approaches is based on running EM iteratively for different numbers of components~\cite{WallaceDowe1994b, WallaceDowe1997a, WallaceDowe2000, oliver1996unsupervised,roberts1998bayesian,biernacki2000assessing}.
-The search method proposed by Figueiredo and Jain~\cite{figueiredo2002unsupervised}, on the other hand, starts with a large number of components, and iteratively deletes components.
-An extension of this method was proposed by Kasarapu and Allison~\cite{kasarapu2015minimum}, which starts with one components, and selectively splits, deletes, or merges components in an attempt to minimise the message length.
-While this method removes the overhead introduces by the search algorithm of Figueiredo and Jain~\cite{figueiredo2002unsupervised}, the search remains greedy, and can be prone to premature convergence in suboptimal solutions.
-The search methods introduced in this paper, incorporate a perturbation step which encourages explorations of new local optima.
-
-{\color{blue}References that we/you seem to be missing include
-\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a}
-% these three above are probably the three most-cited MML papers
-and
-WallaceDowe1993,WallaceDowe1994b,DoweOliverWallace1996,
+%{\color{blue}References that we/you seem to be missing include
+%\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a}
+%% these three above are probably the three most-cited MML papers
+%and
+%WallaceDowe1993,WallaceDowe1994b,DoweOliverWallace1996,
 % these three works above are the earliest MML works on angular data and mixture modelling with angular data
-DoweWallace1997
-% the above is a work about MML statistical consistency
-WallaceDowe1997,WallaceDowe2000
+%DoweWallace1997
+%% the above is a work about MML statistical consistency
+%WallaceDowe1997,WallaceDowe2000
 % these two papers above are MML mixture modelling papers
-Dowe2008a,Dowe2011a
+%Dowe2008a,Dowe2011a
 % these two papers above are MML survey pieces
 
-Things that could also be added include
-ConwaySloane1988
+%Things that could also be added include
+%ConwaySloane1988
 % the above is a book on lattices and lattice constants
-EdwardsDowe1998
+%EdwardsDowe1998
 % this was (and still is) a work on single latent factors in mixture modelling with application to astro' data
-McLachlanPeel2000
+%McLachlanPeel2000
 % this is quite a popular book on finite mixture models
-}
-\section{Minimum Message Length (MML)}
-A complex model is not an optimal one, unless its complexity is justifiable by the added explanatory power.
-There is an established connection between simplicity and truth, as beautifully elucidated by the Occam's Razor principle.
+%}
+
+\section{Minimum Message Length (MML) for mixture modelling}
+A complex model is not an optimal one, unless its complexity is justifiable by 
+the added explanatory power. There is an established connection between 
+simplicity and truth, as beautifully elucidated by the Occam's Razor principle. 
 MML is a statistical technique that formalizes and quantifies this principle.
 
-The basic idea is to consider the transmission of a set of data and a model that describes it.
+%The basic idea is to consider the transmission of a set of data and a model that describes it.
 Clearly one can make increasingly complex models to obtain better fits -- think of epicycles in the solar system.
@@
 
-Traditional curve fitting methods would give a polynomial of degree N-1, which is perhaps the most precise, but it is also the most complex, and the one that most models the noise, by spuriously {\it over-fitting non-existing patterns.\/}
+%Traditional curve fitting methods would give a polynomial of degree N-1, which is perhaps the most precise, but it is also the most complex, and the one that most models the noise, by spuriously {\it over-fitting non-existing patterns.\/}
 
-Over-fitting can result in a model that may be far from the truth.
+%Over-fitting can result in a model that may be far from the truth.
 MML will only consider a more complex model if the encoding of the complex model and the data is more efficient  than the less complex model, i.e.
@@
 
-The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace05}, 
-where a \textit{message} takes into account both the complexity of the model, and its explanatory power.
+%The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace05}, 
+%where a \textit{message} takes into account both the complexity of the model, and its explanatory power.
 
-In essence, MML infers the most concise way of encoding the model, $\theta$, and the data, $x$, given the model, $x|\theta$.
+%In essence, MML infers the most concise way of encoding the model, $\theta$, and the data, $x$, given the model, $x|\theta$.
 A complex model will produce a longer first part of the message than a simple model, since more quantities must be stated.
@@
 
-The \textit{message} must encode two parts: the model, and the data.
-The encoding of the message is based on Shannon's information theory.
-According to Shannon, the information we gain from an event $e$ is $h(e) = - \log_2 p(e)$, where $p(e)$ is the probability of that event.
+The \textit{message} must encode two parts: the model, and the data given the
+model. The encoding of the message is based on Shannon's information theory~\cite{shannon}. 
+The information gained from an event $e$ occurring, where $p(e)$ is the
+probability of that event, is $h(e) = -\log_{2}{p(e)}$. The information content
+is largest for improbable outcomes, and smallest for outcomes that we are 
+almost certain about. In other words, an outcome that has a probability close
+to unity has zero information content because nothing new is learned from it,
+whereas rarer events convey a much higher information content.
+
+%According to Shannon, the information we gain from an event $e$ is $h(e) = - \log_2 p(e)$, where $p(e)$ is the probability of that event.
 The information content is biggest for the improbable outcomes, and smallest for outcomes that we are almost certain about.
@@
 
-Formally, the length of the first part of the message is $- \log_2 h(\vec{\theta})$, and that of the second part is $- \log_2 f(\vec{x}|\vec{\theta})$, where $h(\theta)$ is the prior probability of model $\theta$, and $f(x|\theta)$ is the likelihood of data $x$ assuming a model parameterised by $\vec{\theta}$.
+Formally, the length of the first part of the message is
+    $-\log_{2}h(\vec{\theta})$, 
+and that of the second part is 
+    $-\log_{2}f(\vec{x}|\vec{\theta})$, 
+where $h(\vec{\theta})$ is the prior probability of model described with 
+parameters $\vec{\theta}$, and 
+    $f(\vec{x}|\vec{\theta})$ 
+is the likelihood of data $\vec{x}$ given a model parameterised by $\vec{\theta}$. 
 The total message length is the sum of the two parts, i.e. 
-$- \log_2 h(\vec{\theta}) - \log_2f(\vec{x} | \vec{\theta})$, which is minimised precisely when $h(\vec{\theta}) f(\vec{x}| \vec{\theta})$ is maximised.
-The multiplication of the prior and the likelihood is proportional to the posterior {\em probability} $g(\theta|x)$.
-An important but often misunderstood subtle point is that we are quantising (or rounding off in) parameter space, and so $h(\theta)$ corresponds to a {\em probability} and {\em not} a density.
-Thus, the MML estimate can in some sense be thought of as maximising $h(\theta)f(x|\theta)$ and in turn (in some sense) equal to the posterior mode.
-This is similar in spirit to the (Bayesian) Maximum A Posteriori (MAP) posterior mode, but
-(partly because both terms $- \log_2h(\vec{\theta})$ and $\log_2 f(\vec{x}|\vec{\theta})$ here correspond to probabilities)
-with the benefit of statistical invariance under re-parameterisation
-- for further discussion, see, e.g., ~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes} or immediately below.
+    $-\log_{2} h(\vec{\theta}) - \log_{2}f(\vec{x} | \vec{\theta})$, 
+which is minimised precisely when $h(\vec{\theta}) f(\vec{x}|\vec{\theta})$
+is maximised. The multiplication of the prior and the likelihood is 
+proportional to the posterior {\em probability} $g(\vec{\theta}|\vec{x})$. 
+An important but subtle point is that, within the MML framework, we are 
+quantising (or rounding off in) parameter space, and so 
+$h(\vec{\theta})$ corresponds to a {\em probability} and {\em not} 
+a density. Thus, the MML estimate can in some sense be thought of as maximising 
+    $h(\vec{\theta})f(\vec{x}|\vec{\theta})$ 
+and in turn (in some sense) equal to the posterior mode. This is similar in 
+spirit to the (Bayesian) Maximum A Posteriori (MAP) posterior mode, partly 
+because both terms 
+    $-\log_{2}h(\vec{\theta})$
+and 
+    $\log_{2} f(\vec{x}|\vec{\theta})$ 
+here correspond to probabilities, but with the benefit of statistical 
+invariance under re-parameterisation (e.g., ~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}).
 
-\section{MML for Mixture Modelling}
-MML is used to identify groups (also known as ``classes'') of stars with the  same composition. Mixture modelling enables the partitioning of the data into overlapping groups.
+\subsection{Component message lengths}
 
-In MML we model this by constructing a two-part message, with the first part describing the mixture model (or hypothesis, $\theta$) and the second part representing the encoded data ($x$) given the hypothesis.
-The MML \textit{message} contains the following components:
+Mixture modelling allows for the partitioning of data into overlapping groups.
+The MML principle requires that we encode everything required to reconstruct 
+the message, including our prior beliefs on the model parameters. Consider 
+that we have $N$ data points each with $D$ dimensions which are to be modelled 
+by a finite number of $K$ Gaussian mixtures. Within the context of using the 
+MML principle for mixture modelling, the \emph{message} would have to 
+specifically encode:
 
 \begin{itemize}
-\item[\textbf{1a)}] A statement of the number of groups (or classes, or Gaussian mixture components), $K$.
+    \item[\textbf{(1)}] \textbf{The model}
+    \item[\textbf{a.}]  The number of mixture components (or groups) $K$.
+    \item[\textbf{b.}]  The relative weight $w_k$ (or contribution, or mixing 
+                        proportion) of the amount(s) of data in each group.
+
+%    The relative abundances (or relative weights or mixing proportions) of the amount(s) of data in each group.
+
+    \item[\textbf{c.}]  The parameters of each component, specifically the mean
+                        $\vecmean_k$ and covariance matrix $\veccov_k$ for all 
+                        $K$ Gaussian components.
+
+    \item[\textbf{(2)}] \textbf{The data given the model} -- the likelihood --
+                        where the probability density function $f$ for a
+                        multivariate normal distribution with $D$ dimensions
+                        is given by
+                        \begin{equation}
+                            f(\vec{x}|\vec{\theta}) = \frac{1}{\sqrt{(2\pi)^{D}|\veccov|}}\exp{\left[-\frac{1}{2}(\vec{x} - \vecmean)^\intercal\veccov^{-1}(\vec{x}-\vecmean)\right]}
+                        \end{equation}
+                        \noindent{}and for a mixture of $K$ gaussian mixtures
+                        this becomes
+                        \begin{equation}
+                            f(\vec{x}|\vec{\theta}) = \sum_{k=1}^{K}\weight_{k}f(\vec{x}|\vec{\theta}_{k}) \quad .
+                        \end{equation}
+
+%\item[\textbf{2)}] The data given the model and its parameters.
+% , whose cost is the well-known statistical log-likelihood. In the case of a Normal distribution, this % would appear to a first approximation to equal 
+%   $$-\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}).$$
+%To a first approximation, this would appear to be
+%$ \min_{k} -\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) \rm{,}$
+%minimising (over cluster indices $k$) $ -\log (p_k(x)) $ (which is the cost of encoding the particular cluster, $k$) plus the negative log-likelihood given cluster $k$'s parameters.
+%However, as per \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}\cite[sec. 6.8]{Wallace05}, the cost can be reduced to
+%$ -\log ( \Sigma_{k=1}^{K} p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) $
+\end{itemize}
+
+%A statement of the number of groups (or classes, or Gaussian mixture components), $K$.
 Each class is assumed to have the same prior probability, equal to $2^{-K}$.
@@
 $K$.
-\item[\textbf{1b)}] The relative abundances (or relative weights or mixing proportions) of the amount(s) of data in each group.
-Only $(K - 1)$ weights need be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
-The length of this part is equal (within a small constant) to the negative log of the multiplication of the prior probabilities of each group ($h(\vec{p})$) divided by the square root of the expected
-Fisher information ($\sqrt{F(\vec{p_k})}$). 
-        The Fisher information matrix contains the expected second-order partial derivatives of the log-likelihood function, and quantifies how sharply the likelihood function peaks, and as a result how precisely the parameters should be stated.
 
-    In the case of uniform prior probabilities $h(\vec{p}) = (K-1)!$, and the Fisher information $F(\vec{p}) = N^{K-1} / (\prod_{k=1}^{K} p_k)$.
-It follows that the length of this part is equal to $- \log( h(\vec{p}) \times (1 / \sqrt{{\kappa_K}^K F(\vec{p})}) ) = - \log((K-1)!) + (K/2) \log(\kappa_K) + ((K-1)/2) \log(N) - (1/2) \sum_{k=1}^{K} \log(p_k)$, where $\kappa_K$ is a lattice constant (varying between $1/12$ and $1/(2 \pi e)$).
+A common question arising from listing the components above is that we have
+expliticly encoded the model parameters, but we have not encoded `the model'
+in the sense that we have not described how these encoded parameters relate 
+to each other, nor we do not spell out `\texttt{gaussian}'. This is implicit,
+and justified if we are evaluating and comparing models within the same 
+class (e.g., gaussian mixture models), and is analogous to the probability
+being written as $p(\vec{x}|\vec{\theta})$ rather than what is implied:
+$p(\vec{x}|\vec{\theta},\vec{\mathcal{M}})$, the probability of the data given 
+the model parameters $\vec{\theta}$ and class of model $\vec{\mathcal{M}}$.
+Whatever the message length to encode to class of model $\vec{\mathcal{M}}$,
+it would be a constant added to the message length. Here we provide the 
+message length of each component above in full, and the prior beliefs on those 
+components, before describing our full objective function.
 
-\item[\textbf{1c)}]  The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian components.
+\subsubsection{The number of mixture components (or groups) $K$}
 
-\item[\textbf{2)}] The data given the model and its parameters.
-% , whose cost is the well-known statistical log-likelihood. In the case of a Normal distribution, this % would appear to a first approximation to equal 
-%   $$-\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}).$$
-To a first approximation, this would appear to be
-$ \min_{k} -\log (p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) \rm{,}$
-minimising (over cluster indices $k$) $ -\log (p_k(x)) $ (which is the cost of encoding the particular cluster, $k$) plus the negative log-likelihood given cluster $k$'s parameters.
-However, as per \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}\cite[sec. 6.8]{Wallace05}, the cost can be reduced to
-$ -\log ( \Sigma_{k=1}^{K} p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) $
-\end{itemize}
+We assume a prior on the number of mixtures $K$ of $\prior{K} \propto 2^{-K}$
+\cite[e.g., see p279, sec. 6.8.2, of ][]{Wallace05} such that the length of the 
+optimal lossless message to encode $K$ is 
+\begin{equation}
+    I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant} \quad .
+\end{equation}
+This prior expresses our belief that, in general, there ought to be fewer 
+mixtures.
 
-Parts 1a, 1b and 1c encode the parameters from the clustering process, and part 2 encodes the likelihood (the data given the hypothesis), which represents the goodness-of-fit.
+\subsubsection{The relative weight $w_k$ of each mixture}
 
-The estimation of the multinomial parameters from part 1b is done by quantifying the trade-off between stating the parameters imprecisely, so as to have higher prior probability in the relevant uncertainty region and keep the first part of the message short, and the goodness of fit to the data, which is typically better if we state the parameter estimates more precisely.
+We assume a uniform prior on the individual weights $\weight_{k}$. The weights
+$\weights$ are drawn from a multinomial distribution, such that only $K - 1$
+weights need to be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
 
-Quantitatively, this is done by minimising the sum of the lengths of the three components (1a, 1b, 1c and 2 above).
+The length of this component is equal (within a small constant) to the 
+negative log of the product of the prior probabilities of each group 
+($\prior{w}$) divided by the square root of the expected Fisher information
+($\sqrt{F(\vec{w_k})}$), where the Fisher information matrix contains the
+\emph{expected} second-order partial derivatives of the log-likelihood
+function. This describes how sharply the likelihood function peaks, and as a
+result it quantifies how precisely the parameters should be stated. This
+estimation of the multinomial parameters quantifies the trade-off between
+stating the parameters $\vec{\weights}$ imprecisely, so as to place higher
+prior probability in the relevant uncertainty region and to keep the first
+part of the message short, and the goodness-of-fit to the data, which is
+typically better if we state the estimates of the parameters more precisely.
+
+\noindent{}In the case of uniform prior probabilities,
+\begin{equation}
+    \prior{w} = (K - 1)! \quad ,
+\end{equation}
+\noindent{}and the Fisher information is
+\begin{equation}
+F(\vec{w}) = \frac{N^{K - 1}}{\prod_{k=1}^{K} w_k} \quad ,
+\end{equation}
+\noindent{}which gives the message length of this part to be
+\begin{eqnarray}
+I(\vec{w}) &=& -\log\left(\frac{\prior{\vec{w}}}{\sqrt{{\kappa_K}^K F(\vec{w})}}\right)  \nonumber \\
+I(\vec{w}) &=& -\log{(K - 1)!} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k} \nonumber \\
+I(\vec{w}) &=& -\log{\Gamma(K)} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k}
+\end{eqnarray}
+\noindent{}where $\kappa_K$ is a lattice constant that varies between $1/12$
+and $1/(2\pi e)$ \cite{BoultonWallace1969}.
+
+\subsubsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
+               $K$ Gaussian components}
+test
+
+%Parts 1a, 1b and 1c encode the parameters from the clustering process, and part 2 encodes the likelihood (the data given the hypothesis), which represents the goodness-of-fit.
+
+%The estimation of the multinomial parameters from part 1b is done by quantifying the trade-off between stating the parameters imprecisely, so as to have higher prior probability in the relevant uncertainty region and keep the first part of the message short, and the goodness of fit to the data, which is typically better if we state the parameter estimates more precisely.
+
+%Quantitatively, this is done by minimising the sum of the lengths of the three components (1a, 1b, 1c and 2 above).
 If MML finds that the data for a specific element do not affect the selection of groups, then it will {\it ignore that element.\/} This could tell us something important - such as that there is a large random component for that element or that there are significant observational errors ruining the usefulness of that measurement.
 These are significant insights into the data that are, importantly, quantified by the MML method.
-
-\subsection{Message Length}
-
-The MML principle requires that we encode everything required to reconstruct the message, including our prior beliefs on the model parameters.
-Consider that we have $N$ data points each with $D$ dimensions which are to be modelled by a finite number of Gaussian mixtures.
-Therefore the message for a finite number of Gaussian mixtures must encode:
 
@@
 % \end{enumerate}
-
-We provide the message length of each component in turn before outlining our
-full objective function. We assume a prior on the number of mixtures $K$ of
-% $\prior{K} \propto 2^{-K}$
-$\prior{K} = 2^{-K}$
-\cite[p279, sec. 6.8.2, (a)]{Wallace05} such that the length of the 
-optimal lossless message to encode $K$ is 
-\begin{equation}
-    I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant}.
-\end{equation}
-
-We assume a uniform prior on the individual weights $\weight_{k}$. The weights
-$\weights$ are drawn from a multinomial distribution such that they can be 
-optimally encoded in a message with length~\todo{\cite{someone}}:
-\begin{equation}
-  I(\weights) 
-    = \frac{K - 1}{2}\log{N} 
-    - \frac{1}{2}\sum_{k=1}^{K}\log\weight_k 
-    - \log{\Gamma{\left(K\right)}} \quad .
-\end{equation}
 

### 2018-04-19 article/ms.tex
ermgerd somuch text

@@
 \newcommand{\alda}[1]{\nbb{Aldeida}{#1}}
+\newcommand{\andy}[1]{\nbb{Andy}{#1}}
 
@@
 \def\infinity{\rotatebox{90}{8}}
-\def\veccov{\vect{C}}
+\def\cov{C}
+\def\veccov{\vect{\cov}}
 \def\vecmean{\vect{\mu}}
@@
 \begin{equation}
-    AIC = -2 \log_{e}f(\vec{x} | \vec{\theta}) + 2 k
-\end{equation}
-\noindent{}where the likelihood $f(\vec{x} | \vec{\theta})$ is the probability 
-of the data $\vec{x}$ given a model parameterised by $\vec{\theta}$, and $k$ 
+    AIC = -2 \log_{e}f(\vec{\data} | \vec{\theta}) + 2 k
+\end{equation}
+\noindent{}where the likelihood $f(\vec{\data} | \vec{\theta})$ is the probability 
+of the data $\vec{\data}$ given a model parameterised by $\vec{\theta}$, and $k$ 
 is the number of free parameters in the model. BIC~\cite{schwarz1978estimating} 
@@
 The information gained from an event $e$ occurring, where $p(e)$ is the
-probability of that event, is $h(e) = -\log_{2}{p(e)}$. The information content
+probability of that event, is $I(e) = -\log_{2}{p(e)}$. The information content
 is largest for improbable outcomes, and smallest for outcomes that we are 
@@
 Formally, the length of the first part of the message is
-    $-\log_{2}h(\vec{\theta})$, 
+    $-\log_{2}\prior{\vec{\theta}}$, 
 and that of the second part is 
-    $-\log_{2}f(\vec{x}|\vec{\theta})$, 
-where $h(\vec{\theta})$ is the prior probability of model described with 
+    $-\log_{2}f(\vec{\data}|\vec{\theta})$, 
+where $\prior{\vec{\theta}}$ is the prior probability of model described with 
 parameters $\vec{\theta}$, and 
-    $f(\vec{x}|\vec{\theta})$ 
-is the likelihood of data $\vec{x}$ given a model parameterised by $\vec{\theta}$. 
+    $f(\vec{\data}|\vec{\theta})$ 
+is the likelihood of data $\vec{\data}$ given a model parameterised by $\vec{\theta}$. 
 The total message length is the sum of the two parts, i.e. 
-    $-\log_{2} h(\vec{\theta}) - \log_{2}f(\vec{x} | \vec{\theta})$, 
-which is minimised precisely when $h(\vec{\theta}) f(\vec{x}|\vec{\theta})$
+    $-\log_{2} \prior{\vec{\theta}} - \log_{2}f(\vec{\data} | \vec{\theta})$, 
+which is minimised precisely when $\prior{\vec{\theta}} f(\vec{\data}|\vec{\theta})$
 is maximised. The multiplication of the prior and the likelihood is 
-proportional to the posterior {\em probability} $g(\vec{\theta}|\vec{x})$. 
+proportional to the posterior {\em probability} $g(\vec{\theta}|\vec{\data})$. 
 An important but subtle point is that, within the MML framework, we are 
 quantising (or rounding off in) parameter space, and so 
-$h(\vec{\theta})$ corresponds to a {\em probability} and {\em not} 
+$\prior{\vec{\theta}}$ corresponds to a {\em probability} and {\em not} 
 a density. Thus, the MML estimate can in some sense be thought of as maximising 
-    $h(\vec{\theta})f(\vec{x}|\vec{\theta})$ 
+    $\prior{\vec{\theta}}f(\vec{\data}|\vec{\theta})$ 
 and in turn (in some sense) equal to the posterior mode. This is similar in 
@@
 because both terms 
-    $-\log_{2}h(\vec{\theta})$
+    $-\log_{2}\prior{\vec{\theta}}$
 and 
-    $\log_{2} f(\vec{x}|\vec{\theta})$ 
+    $\log_{2} f(\vec{\data}|\vec{\theta})$ 
 here correspond to probabilities, but with the benefit of statistical 
@@
                         \begin{equation}
-                            f(\vec{x}|\vec{\theta}) = \frac{1}{\sqrt{(2\pi)^{D}|\veccov|}}\exp{\left[-\frac{1}{2}(\vec{x} - \vecmean)^\intercal\veccov^{-1}(\vec{x}-\vecmean)\right]}
+                            f(\vec{\data}|\vec{\theta}) = \frac{1}{\sqrt{(2\pi)^{D}|\veccov|}}\exp{\left[-\frac{1}{2}(\vec{\data} - \vecmean)^\intercal\veccov^{-1}(\vec{\data}-\vecmean)\right]}
                         \end{equation}
@@
                         \begin{equation}
-                            f(\vec{x}|\vec{\theta}) = \sum_{k=1}^{K}\weight_{k}f(\vec{x}|\vec{\theta}_{k}) \quad .
+                            f(\vec{\data}|\vec{\theta}) = \sum_{k=1}^{K}\weight_{k}f(\vec{\data}|\vec{\theta}_{k}) \quad .
+                        \end{equation}
+                        \noindent{}Such that the log-likelihood $\log\likelihood(\vec{\data}|\vec{\theta})$ is
+                        \begin{equation}
+                            \log\likelihood(\vec{\data}|\vec{\theta}) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\weight_{k}f(\vec{\data}_{n}|\vecmean_{k},\veccov_{k}) \quad .
                         \end{equation}
@@
 
-A common question arising from listing the components above is that we have
-expliticly encoded the model parameters, but we have not encoded `the model'
+Calculating the length of the message can be a non-trivial task, especially
+if the model is reasonably complex. This makes the MML principle intractable
+(or uncomputable) in most cases, and forces us to make approximations when 
+calculating the message length. Using a Taylor expansion, a generalised scheme
+can be derived to estimate the parameter vector $\vec{\theta}$ that minimises
+the message length expression\cite{WallaceFreeman1987}:
+
+\begin{equation}
+    I(\vec{\theta}, \vec{\data}) = \frac{Q}{2}\log{\kappa(Q)} - \log\left(\frac{\prior{\vec{\theta}}}{\sqrt{|\fisher{\vec{\theta}}|}}\right) - \log\likelihood\left(\vec{\data}|\vec{\theta}\right) + \frac{Q}{2} \quad .
+    \label{eq:taylor}
+\end{equation}
+
+Here $\prior{\vec{\theta}}$ is the prior, $Q$ is the number of free parameters
+in the model, $\kappa(Q)$ is a function to approximate the lattice quantisation
+constant for $Q$ free parameters\cite{ConwaySloane1984}, and $|\fisher{\vec{\theta}}|$
+is the determinant of the \emph{expected} Fisher information matrix (the second
+order partial derivatives of the negative log-likelihood function $-\log\likelihood\left(\vec{\data}|\vec{\theta}\right)$).
+A common that arises after listing the components above is that although we
+have expliticly encoded the model parameters, we have not encoded `the model'
 in the sense that we have not described how these encoded parameters relate 
-to each other, nor we do not spell out `\texttt{gaussian}'. This is implicit,
-and justified if we are evaluating and comparing models within the same 
-class (e.g., gaussian mixture models), and is analogous to the probability
-being written as $p(\vec{x}|\vec{\theta})$ rather than what is implied:
-$p(\vec{x}|\vec{\theta},\vec{\mathcal{M}})$, the probability of the data given 
-the model parameters $\vec{\theta}$ and class of model $\vec{\mathcal{M}}$.
-Whatever the message length to encode to class of model $\vec{\mathcal{M}}$,
-it would be a constant added to the message length. Here we provide the 
-message length of each component above in full, and the prior beliefs on those 
-components, before describing our full objective function.
+to each other. In a sense, we do not describe to the receiver that these
+encoded parameters refer to a mixture of Gaussians, or what we mean when we
+refer to the term `Gaussian'. This information is implicit, and justified
+if we are evaluating and comparing models within the same class (e.g., gaussian 
+mixture models). For example, it is analogous to writing the probability of
+some data given a model as $p(\vec{\data}|\vec{\theta})$ rather than what is 
+actually implied: $p(\vec{\data}|\vec{\theta},\vec{\mathcal{M}})$, the 
+probability of the data given the model parameters $\vec{\theta}$ and class of 
+model $\vec{\mathcal{M}}$. Whatever the message length required to encode to 
+class of model $\vec{\mathcal{M}}$, under consideration, it would be a constant 
+added to the message length regardless of the number of components in our mixture.
+
+In the following subsections we provide expressions for the message length of
+each component listed above, and our prior beliefs, before describing our
+full objective function.
 
@@
 
-We assume a prior on the number of mixtures $K$ of $\prior{K} \propto 2^{-K}$
-\cite[e.g., see p279, sec. 6.8.2, of ][]{Wallace05} such that the length of the 
+We assume a prior on the number of components $K$ of $\prior{K} \propto 2^{-K}$
+\cite[][p279, sec. 6.8.2]{Wallace05} such that the length of the 
 optimal lossless message to encode $K$ is 
@@
 \end{equation}
-This prior expresses our belief that, in general, there ought to be fewer 
-mixtures.
-
-\subsubsection{The relative weight $w_k$ of each mixture}
+This  expresses our prior belief that there ought to be fewer components.
+
+\subsubsection{The relative weight $w_k$ of each component}
 
@@
 $\weights$ are drawn from a multinomial distribution, such that only $K - 1$
-weights need to be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
-
-The length of this component is equal (within a small constant) to the 
-negative log of the product of the prior probabilities of each group 
-($\prior{w}$) divided by the square root of the expected Fisher information
-($\sqrt{F(\vec{w_k})}$), where the Fisher information matrix contains the
-\emph{expected} second-order partial derivatives of the log-likelihood
-function. This describes how sharply the likelihood function peaks, and as a
-result it quantifies how precisely the parameters should be stated. This
-estimation of the multinomial parameters quantifies the trade-off between
-stating the parameters $\vec{\weights}$ imprecisely, so as to place higher
-prior probability in the relevant uncertainty region and to keep the first
-part of the message short, and the goodness-of-fit to the data, which is
+weights need to be encoded because $\sum_{k=1}^{K}\weight_k = 1$. Following
+from Equation \ref{eq:taylor}, here we describe how our estimate of the message 
+length of the multinomial parameter $\vec{\weight}$ quantifies the trade-off 
+between stating the parameters $\vec{\weights}$ imprecisely, so as to keep the 
+first part of the message short and place higher prior probability in the 
+relevant uncertainty region, and the goodness-of-fit to the data, which is 
 typically better if we state the estimates of the parameters more precisely. 
-
-\noindent{}In the case of uniform prior probabilities,
-\begin{equation}
-    \prior{w} = (K - 1)! \quad ,
-\end{equation}
-\noindent{}and the Fisher information is
-\begin{equation}
-F(\vec{w}) = \frac{N^{K - 1}}{\prod_{k=1}^{K} w_k} \quad ,
+By balancing this trade-off, the length of this component is equal (within a 
+small constant) to the negative log of the product of the prior probabilities 
+of each group ($\prior{\vec{\weight}}$) divided by the square root of the 
+expected Fisher information ($\sqrt{\fisher{\vec{\weight_k}}}$), where the 
+Fisher information matrix contains the \emph{expected} second-order partial 
+derivatives of the log-likelihood function. This describes how sharply the 
+likelihood function peaks, and as a result it quantifies how precisely the 
+parameters should be stated. In the case of uniform prior probabilities,
+\begin{equation}
+    \prior{\vec{\weight}} = (K - 1)! \quad ,
+\end{equation}
+\noindent{}the Fisher information is
+\begin{equation}
+\fisher{\vec{w}} = \frac{N^{K - 1}}{\prod_{k=1}^{K} w_k} \quad ,
 \end{equation}
@@
 \begin{eqnarray}
-I(\vec{w}) &=& -\log\left(\frac{\prior{\vec{w}}}{\sqrt{{\kappa_K}^K F(\vec{w})}}\right)  \nonumber \\
+I(\vec{w}) &=& -\log\left(\frac{\prior{\vec{w}}}{\sqrt{{\kappa_K}^K \fisher{\vec{w}}}}\right)  \nonumber \\
 I(\vec{w}) &=& -\log{(K - 1)!} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k} \nonumber \\
@@
                $K$ Gaussian components}
-test
-
-%Parts 1a, 1b and 1c encode the parameters from the clustering process, and part 2 encodes the likelihood (the data given the hypothesis), which represents the goodness-of-fit.
-
-%The estimation of the multinomial parameters from part 1b is done by quantifying the trade-off between stating the parameters imprecisely, so as to have higher prior probability in the relevant uncertainty region and keep the first part of the message short, and the goodness of fit to the data, which is typically better if we state the parameter estimates more precisely.
-
-%Quantitatively, this is done by minimising the sum of the lengths of the three components (1a, 1b, 1c and 2 above).
-If MML finds that the data for a specific element do not affect the selection of groups, then it will {\it ignore that element.\/} This could tell us something important - such as that there is a large random component for that element or that there are significant observational errors ruining the usefulness of that measurement.
-These are significant insights into the data that are, importantly, quantified by the MML method.
-
-% \begin{enumerate}
-% k has been used previously in describing AIC and BIC.
-% If k is the number of parameters, then it will be neither (the number of clusters) nor an index running from 1 to (the number of clusters).
-% \item
-% \begin{itemize}
-% \item (1a) ~ The number of Gaussian mixture components, $K$.
-% \item (1b) ~ The relative weights $\weights$ of the $K - 1$ Gaussian components. Only
-%       $K - 1$ weights need be encoded because $\sum_{k=1}^{K}\weight_k = 1$.
-% \item (1c) ~ The component parameters $\vecmean$ and $\veccov$ for all $K$ Gaussian
-%       components.
-% \end{itemize}
-% \item (2) ~ The data, given the model parameters.
-%  \item The lattice quantisation constant $\kappa(Q)$ for the number of model       parameters $Q$.
-% \end{enumerate}
 
@@
     $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
-However,  the MML principle \todo{demands} proper priors. In practice we make 
+However,  the MML principle requires proper priors. In practice we make 
 this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
@@
 We adopt a conjugate inverted Wishart prior for the covariance matrix of an
-individual mixture
-- e.g.,
-%[]
-\cite[Section 5.2.3]{Schafer_1997}:
+individual mixture \cite[Section 5.2.3]{Schafer_1997}:
 
@@
 determinant of the expected Fisher information matrix for $\vecmean_k$ and
-$\veccov_k$ in order to encode the full message length. We approximate the
-determinant of the Fisher information $\detfisher{{\vecmean_k, \veccov_k}}$ 
+$\veccov_k$ in order to encode the full message. In practice the determinant
+of the Fisher information is non-trivial to evaluate, and as such we
+approximate the determinant of the Fisher information
+ $\detfisher{{\vecmean_k, \veccov_k}}$ 
 as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
@@
 we take the second derivative of the log-likelihood function 
-$\likelihood(\data|\vecmean,\veccov)$, which yields:
+$\likelihood(\data|\vectheta)$, which yields:
 \begin{equation}
@@
     + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
-  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
-    -\frac{KD}{2}\log{2} \nonumber \\
-  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
+%  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
+    %-\frac{KD}{2}\log{2} \nonumber \\
+%  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
   I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
@@
 
+\subsubsection{The data given the model}
 If we assume that the data have homoskedastic noise properties, then the 
@@
 
-%  Here
-
- Each $k$-th mixture is a $D$-dimensional Gaussian that contributes a relative weight $\weight_k$ such that $\sum_{k=1}^{K}\weight_k = 1$.
-We will assume that the data have homoscedastic noise properties.
-
-The data have the same error value, $y_{err}$, in all $D$ dimensions, for all $N$ observations.
-
-The message length for the data is given by:
-
-Here we introduce the individual components that contribute to the message length.
-
-Combining these expressions, the complete message length for $K$ mixtures is given by
-
-\begin{eqnarray}
-I_K & = & K\log{2} % I_k
-    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
-    +\mathcal{L}(\data|\vectheta) \nonumber \\
-    & -& DN\log{y_{err}} % Likelihood
-   +  \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right]\nonumber \\% I_t
-    &-& \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
-\end{eqnarray}
-
-\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
-$\mathcal{L}$ is the log-likelihood of a multivariate Gaussian distribution.
+\subsection{Objective function}
+
+Combining the expressions for the message length (or information content) of
+each component, we arrive at our objective function $I$ to minimise across 
+all model parameters (including the number of mixtures, $K$):
+
+\begin{equation}
+\todo{I = 0}
+\end{equation}
+
+%\begin{eqnarray}
+%I_K & = & K\log{2} % I_k
+%    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
+%    +\mathcal{L}(\data|\vectheta) \nonumber \\
+%    & -& DN\log{y_{err}} % Likelihood
+%   +  \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right]\nonumber \\% I_t
+%    &-& \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
+%%\end{eqnarray}
+
+%\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
+%$\mathcal{L}$ is the log-likelihood of a multivariate Gaussian distribution.
+
+While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, where 
+for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
+\vecmean, \veccov\}$, and we can meaningfully compare models with different numbers of components or parameter
+values, we require a search strategy in order to efficiently find the global
+solution: the minimum message length over a class of finite mixture models
+(where $N > K$). Before we introduce and compare search strategies,
+let us first make some comments and quantitative descriptions on how the
+total message length $I$ will vary with the model parameters. Specifically, we
+will re-arrange Equation \ref{eq:objective_function} to be a sum of three 
+smaller functions, and we will derive limits and predictions for those 
+quantites. In an ideal scenario, those limits and predictions would be able
+to provide us with an estimate of the message length for a Gaussian mixture
+before computing the mixture (i.e., before performing expectation-maximization.
+We begin by re-casting Equation \ref{eq:objective_function} as
+\begin{eqnarray}
+I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
+                          + I_{2}(K, N, \veccov)
+                          + I_{3}(\vec{\psi}) \quad ,
+\end{eqnarray}
+\noindent{}where:
+\begin{eqnarray}
+    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
+          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
+    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad {\textrm and},\\
+    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\sigma_{\datum}} \quad .
+\end{eqnarray}
+The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
+message that we introduced earlier to describe message length. Rather we are
+simply re-arranging the objective function because, as we will show, there are
+useful theoretical bounds and predictions that we can make for each part of
+this re-arranged expression.
+
+In Eq. \ref{eq:I1} it can be seen that for any finite mixture, $I_1$ can be
+almost directly computed just from knowing the number of mixture components
+$K$ and the properties of the data: the number of data points $N$ and its
+dimensionality $D$. The only ambiguous term is $\sum_{k=1}^{K}\log\weight_k$,
+which will depend on the data. However, for any finite mixture of $K$
+components we can consider the two extremes of $\sum_{k=1}^{K}\log\weight_k$:
+
+\begin{enumerate}
+    \item The weights are distributed uniformly such that $w_k = 1/K$ and
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
+          \end{equation}
+    \item The weights are distributed such that $K - 1$ components each 
+          describe just one data point $w_{1..K-1} = 1/N$, and the remaining 
+          component describes all other data points $w_{K} = 1 - (K-1)/N$
+          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
+          \end{equation}
+          \noindent{}and by making use of the identity 
+            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$
+          we can simplify this expression to:
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
+          \end{equation}
+\end{enumerate}
+
+\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
+derived the bounds
+\begin{equation}
+-K\log{K} \leq \sum_{k=1}^{K}\log\weight_k \leq - K\log{K} + \log(N + 1 - K)
+\end{equation}
+\noindent{}which gives us analytic expressions for the upper and lower value 
+of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture.
+
+Developing similar theoretical bounds for $I_2$ is not straightforward, because
+the determinants of the covariance matrices of Gaussian components will not
+-- to our knowledge -- strictly follow some expected distribution, in the way 
+that the weights $\weights$ follow a multinomial distribution. Similarly, the
+\emph{scale} of $|\veccov_k|$ will depend entirely on the distribution of the
+data. In that sense, we cannot a priori state with any certainty about what 
+values $|\veccov_k|$ might take (or $\log{|\veccov_k|}$) without knowing
+something about the data first. Even demanding that the covariance matrices
+$\veccov_k$ are positive semi-definite only informs us that $\log{|\veccov_k|}$
+is finite. However, there are quantitative predictions we can make on the 
+bounds of $\sum_{k=1}^{K}\log{|\veccov_k|}$ if something is known about the
+data. Specifically if we consider the determinant of the sample covariance 
+matrix, which we denote $|\cov_{s}|$, then we can state with confidence that
+the sample covariance matrix describes the maximum amount of variance that 
+could be captured by the components in the mixture. For example, a limiting
+case of $\sum_{k=1}^{K}\log{|\veccov_k|}$ can be described when the determinant
+of the sample covariance matrix must be equally (uniformly) captured by $K$ 
+components in a mixture. In other words, the maximum variance in the data 
+must be equally spread between the components in the mixture. Uniformly 
+describing the variance in the data is the limiting case because any 
+non-uniformity would lead to smaller $|\veccov|$ for one mixture (and larger 
+for another), and that asymmetry would lead to a higher $I_2$ because 
+$\log(x - \epsilon) + \log(x + \epsilon) < 2\log{x}$, and because of the 
+$-(1 + \frac{D}{2})$ term that enters multiplicatively in the calculation of 
+$I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
+the sample covariance matrix.
+
+It can be seen that for some finite mixture, the total message length depends 
+on the number of components $K$, the number of data points $N$ and its 
+dimensionality $D$, and the parameters of that mixture: $\vec{\weight}$, 
+$\vecmean$, and $\veccov$. That is to say that we can re-arrange Equation
+\ref{eq:objective_function} to separate our objective function into three
+smaller functions, and use information theory to determine limits or 
+predictions for those quantities. Specifically,
+
+Re-arranging Equation \ref{eq:objective_function},
+we can describe our objective function as being the sum of three functions:
+$I_{g}(N, D, K, \vec{\weight})$
+
+given mixture of finite
+components, the total message length depends on the
+
+Before we introduce and compare search strategies,
+here we will
 

### 2018-04-19 article/ms.tex
wip

@@
 $I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
-the sample covariance matrix.
+the sample covariance matrix $|\cov_{s}|$.
+
+The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
+possible variance that could be captured by any single component.
 

### 2018-04-20 article/ms.tex
foo

@@
 
-While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, where 
-for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
-\vecmean, \veccov\}$, and we can meaningfully compare models with different numbers of components or parameter
-values, we require a search strategy in order to efficiently find the global
-solution: the minimum message length over a class of finite mixture models
-(where $N > K$). Before we introduce and compare search strategies,
-let us first make some comments and quantitative descriptions on how the
-total message length $I$ will vary with the model parameters. Specifically, we
-will re-arrange Equation \ref{eq:objective_function} to be a sum of three 
-smaller functions, and we will derive limits and predictions for those 
-quantites. In an ideal scenario, those limits and predictions would be able
-to provide us with an estimate of the message length for a Gaussian mixture
-before computing the mixture (i.e., before performing expectation-maximization.
-We begin by re-casting Equation \ref{eq:objective_function} as
+While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
+where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
+\vecmean, \veccov\}$, and we can meaningfully compare models with different 
+numbers of components or parameter values, we require a search strategy in 
+order to efficiently find the global solution: the minimum message length 
+over a class of finite mixture models (where $N > K$). Before we introduce and 
+compare search strategies, let us first make some comments and quantitative 
+descriptions on how the message length $I$ will vary with the model parameters. 
+Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
+sum of three smaller functions, and we will derive limits and predictions for 
+those quantites. In an ideal scenario, those limits and predictions would be 
+able to provide us with an estimate of the message length for a Gaussian 
+mixture before computing the mixture (i.e., before performing 
+expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
+as
 \begin{eqnarray}
@@
 \begin{equation}
--K\log{K} \leq \sum_{k=1}^{K}\log\weight_k \leq - K\log{K} + \log(N + 1 - K)
+-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad - K\log{K} + \log(N + 1 - K)
 \end{equation}
 \noindent{}which gives us analytic expressions for the upper and lower value 
-of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture.
+of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
+bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.
 
@@
 data. Specifically if we consider the determinant of the sample covariance 
-matrix, which we denote $|\cov_{s}|$, then we can state with confidence that
+matrix, which we denote $|\cov_{u}|$ (where the $u$ subscript denotes `upper',
+for the upper limit on the covariance matrix), then we can state with confidence that
 the sample covariance matrix describes the maximum amount of variance that 
@@
 $I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
-the sample covariance matrix $|\cov_{s}|$.
+the sample covariance matrix $|\cov_{u}|$:
+
+\begin{eqnarray}
+    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{\frac{|\cov_u|}{K}} \nonumber \\
+    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\left(K\log{|\cov_u|} - K\log{K}\right) 
+\end{eqnarray}
 
 The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
-possible variance that could be captured by any single component.
+possible variance that could be captured by any single component. In theory
+this approaches zero, but in practice it will be set by the minimum pairwise
+distance in $D$ dimensions. Specifically, if we consider two data points to be
+captured by a component in the mixture, an those two data points are the
+closest two data points (in high dimensional space) of any data points, then
+the variance captured by that Gaussian component will represent the \emph{minimum}
+possible variance that can be captured by any single component.  
+In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
+in $D$ dimensions, average it over the dimensions, and approximate the minimum
+determinant of the covariance matrix $|\cov_l|$ as
+\begin{equation}
+    |\cov_{l}| \approx \prod_{d=1}^{D}\frac{d_{min}^2}{4} \quad ,
+\end{equation}
+\noindent{}which approximately represents the variance between two data points,
+but ignores correlations and assumes only a mean distance in $D$ dimensions
+(for comuptational cost). With these approximations,
+\begin{eqnarray}
+    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\frac{d_{min}^2}{4}}  \nonumber \\
+    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
+\end{eqnarray}
+\noindent{}and the upper limit on $I_2$ is:
+\begin{eqnarray}
+    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
+    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
+\end{eqnarray}
+
+Here our bounds on $I_2$ are only dependent on $K$, $D$, the minimum euclidean
+pair-wise distance $d_{min}$, and the determinant of the sample covariance
+matrix $|\cov_u|$. In practice it is the lower bound on $I_{2}$ that is 
+arguably more useful for search and optimisation, but one could conceive
+situations where there is great utility in also having upper bounds on $I_1$
+and $I_2$.
+
+There is no finite upper bound on $I_{3}$ because the negative log likelihood
+$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
+data. A lower limit, however, does exist.
+
+\andy{here}
 

### 2018-04-20 gmmmml/em.py
updates

@@ -490,6 +490,6 @@ def maximization(y, mu, cov, weight, responsibility, parent_responsibility=1,
     assert np.all(np.isfinite(new_mu))
     assert np.all(np.isfinite(new_cov))
     assert np.all(np.isfinite(new_weight))
-    
+
 
     return state 

### 2018-04-20 gmmmml/gmm.py

@@ -6,7 +6,7 @@ Model data with a mixture of gaussians.
 import logging
 import numpy as np
 import scipy
-from scipy.spatial.distance import pdist
+from scipy.spatial import distance
 
 from . import (em, mml)
 
@@ -227,11 +227,12 @@ class GaussianMixture(object):
 
         N, D = y.shape
 
-        mmpwd = self._state_meta.get("min_mean_pairwise_distance", None)
+        pwdm = self._state_meta.get("pairwise_distance_matrix", None)
 
-        if mmpwd is None:
-            mmpwd = np.min(pdist(y))/(2.0 * D)
-            self._state_meta["min_mean_pairwise_distance"] = mmpwd
+        if pwdm is None:
+
+            raise a
+            self._state_meta["pairwise_distance_matrix"] = mmpwd
 
         predictions, meta = mml.predict_message_length(K, N, D, 
             previous_states=(

### 2018-04-20 gmmmml/mml.py

@@ -473,7 +473,7 @@ def _deprecated_predict_sum_log_det_covs(K, previous_states, draws=100, **kwargs
 
 
 def predict_negative_log_likelihood(K, N, D, uniformity_fraction,
-    previous_states, **kwargs):
+    previous_states, min_mean_pairwise_distance, **kwargs):
     r"""
     Predict the negative log likelihood for future (target) mixtures with
     :math:`K` components.
@@ -571,6 +571,10 @@ def predict_negative_log_likelihood(K, N, D, uniformity_fraction,
     nll_lower_bound = nll.copy()
     nll += 0.5 * N * D * pred_chisqs
 
+    #nll_lower_bound = N * np.log(N) + (K - 1 - N) * np.log(N - K + 1) \
+    #                + N * K * D * np.log(min_mean_pairwise_distance/2.0) \
+    #                + 0.5 * N * D * (1 + np.log(2 * np.pi))
+
     return (nll, nll_lower_bound)
 
 
@@ -622,7 +626,7 @@ def predict_message_length(K, N, D, previous_states, yerr=0.001,
         K, N=N, previous_states=(_state_K, _state_det_covs), **state_meta)
 
     p_nll, t_nll_lower = predict_negative_log_likelihood(
-        K, N, D, uniformity_fraction, previous_states=(
+        K, N, D, uniformity_fraction, min_mean_pairwise_distance=min_mean_pairwise_distance, previous_states=(
             _state_K, 
             _state_det_covs,
             _state_sum_log_likelihoods

### 2018-04-20 article/ms.tex
maath

@@
 \begin{equation}
-    |\cov_{l}| \approx \prod_{d=1}^{D}\frac{d_{min}^2}{4} \quad ,
+    |\cov_{l}| \approx \prod_{d=1}^{D}\left(\frac{d_{min}}{2}\right)^2 \quad ,
 \end{equation}
@@
 \begin{eqnarray}
-    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\frac{d_{min}^2}{4}}  \nonumber \\
+    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\left(\frac{d_{min}}{2}\right)^2}  \nonumber \\
     \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
@@
 
-There is no finite upper bound on $I_{3}$ because the negative log likelihood
+There is no finite upper bound on $I_{3}$ because the negative log-likelihood
 $-\log\likelihood$ will approach infinity for an increasingly poor fit to the
-data. A lower limit, however, does exist.
+data. A useful lower limit, however, does exist. The negative log-likelihood
+in Eq. \ref{eq:log-likelihood} can be expanded as,
+\begin{equation}
+    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
+\end{equation}
+\noindent{}and we can define
+\begin{equation}
+\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
+\end{equation}
+For the sake of placing a lower bound on $I_{3}$, let us assume that each data
+point will be `mostly' explained by one of the components in the 
+mixture\footnote{We model the data with partial assignment, but here we merely 
+seek to place a lower bound on the negative log-likelihood.}, but we do not 
+necessarily know which component, or what the parameters of that component are. 
+In that situation, we can state that the component that \emph{does} explain
+the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
+whereas the other components that do not explain the data will have 
+$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
+this simplification we can write an approximation for the negative log-likelihood
+that does not require knowledge about which data points are associated to what
+component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
+will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
+approximate lower bound is:
+\begin{eqnarray}
+    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
+    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad .
+\nonumber
+    \label{eq:logl-lower-bound}
+\end{eqnarray}
+\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
+for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
+to the smallest realistic values, where those mixtures encapsulate only two
+data points and $\todo{w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N}$:
+
+\begin{eqnarray}
+-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& -N\left[(K-1)\left(\frac{1}{N}\log\frac{1}{N}\right) + \left(1 - \frac{K-1}{N}\right)\log\left(1 - \frac{K-1}{N}\right)\right] \nonumber \\
+-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& (K - 1)\log{N} + (K - 1 - N)\log\left(1-\frac{K-1}{N}\right) \nonumber \\
+-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& N\log{N} + (K - 1 - N)\log\left(N-K+1\right) \quad .
+\end{eqnarray}
+
+The remaining term in Eq. \ref{eq:logl-lower-bound}, $\frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log|\vec{\cov}_k|$, requires some  consideration.
 
 \andy{here}
-
-It can be seen that for some finite mixture, the total message length depends 
-on the number of components $K$, the number of data points $N$ and its 
-dimensionality $D$, and the parameters of that mixture: $\vec{\weight}$, 
-$\vecmean$, and $\veccov$. That is to say that we can re-arrange Equation
-\ref{eq:objective_function} to separate our objective function into three
-smaller functions, and use information theory to determine limits or 
-predictions for those quantities. Specifically,
-
-Re-arranging Equation \ref{eq:objective_function},
-we can describe our objective function as being the sum of three functions:
-$I_{g}(N, D, K, \vec{\weight})$
-
-given mixture of finite
-components, the total message length depends on the
-
-Before we introduce and compare search strategies,
-here we will
 

### 2018-04-20 gmmmml/gmm.py
stuff

@@ -35,12 +35,30 @@ class GaussianMixture(object):
     :param max_em_iterations: [optional]
         The maximum number of iterations to run per expectation-maximization
         loop (default: ``10000``).
+
+    :param information_bound: [optional]
+        Whether or not to use a strict calculation for the lower bound on the
+        predictions for the sum of the log of the determinant of the covariance
+        matrices.
+
+        If 'strict' is employed, then the pair-wise distances between all points
+        will be calculated, and this matrix will be used when predicting the
+        entropy of future gaussian mixtures. This can be computationally
+        expensive for large data sets.
+
+        If 'loose' is employed, then no pair-wise distances will be calculated
+        and the logarithm of the determinant of the covariance matrices will
+        be approximated by a Gaussian kernel density estimator. This approach
+        is computationally efficient and reasonaly accurate, but may result in
+        poor or uncertain predictions when the variance of components in the 
+        mixture vary by several orders of magnitude.
     """
 
     parameter_names = ("mean", "covariance", "weight")
 
     def __init__(self, covariance_type="full", covariance_regularization=0, 
-        threshold=1e-3, max_em_iterations=1000, **kwargs):
+        threshold=1e-3, max_em_iterations=1000, information_bound="strict",
+        **kwargs):
 
         available = ("full", )
         covariance_type = covariance_type.strip().lower()
@@ -59,10 +77,18 @@ class GaussianMixture(object):
         if 1 > max_em_iterations:
             raise ValueError("max_em_iterations must be a positive integer")
 
+        available = ("strict", "loose")
+        information_bound = information_bound.strip().lower()
+        if information_bound not in available:
+            raise ValueError("information bound type '{}' is invalid. Must be "\
+                             "one of: {}".format(
+                                information_bound, ", ".join(available)))
+
         self._threshold = threshold
         self._max_em_iterations = max_em_iterations
         self._covariance_type = covariance_type
         self._covariance_regularization = covariance_regularization
+        self._information_bound = information_bound
 
         # Lists to record states for predictive purposes.
         self._state_K = []
@@ -231,7 +257,9 @@ class GaussianMixture(object):
 
         if pwdm is None:
 
-            raise a
+            pwdm = distance.dist(y, y)
+
+
             self._state_meta["pairwise_distance_matrix"] = mmpwd
 
         predictions, meta = mml.predict_message_length(K, N, D, 

### 2018-09-10 gmmmml/utils.py
add concentration

@@ -2,9 +2,146 @@
 import numpy as np
 from sklearn import datasets
 
+from sklearn.neighbors import NearestNeighbors
+
+def concentration(y, K_max=None):
+    """
+    Calculate the maximum concentration as a function of the number of data
+    points. Specifically, calculate the minimum of
+
+    .. math:
+
+        w_{k}\log{|C_k|}
+
+    for any subset of the N data points that exist. That is, find the :math:`M` 
+    nearest points in order to calculate the minimum sample covariance among 
+    :math:`M` points, and do this for increasing :math:`M` values. Then for
+    each number of :math:`K` components, calculate the minimum mixture among a
+    set o
+    """
+
+    y = np.atleast_2d(y)
+    N, D = y.shape
+
+    if K_max is None or K_max > N:
+        K_max = N
+
+    knn = NearestNeighbors().fit(y)
+    distances, indices = knn.kneighbors(y, n_neighbors=2)
+    
+    # Successively build up the minimum concentration.
+    K = np.arange(1, 1 + K_max, dtype=int)
+    minimum_concentration = np.zeros(K.size, dtype=float)
+    maximum_concentration = np.zeros(K.size, dtype=float)
+    minimum_concentration[0] = maximum_concentration[0] \
+                             = np.sum(np.log(np.var(y, axis=0)))
+    
+    minimum_sum_log_dets = np.zeros(K.size)
+    minimum_sum_log_dets[0] = minimum_concentration[0]
+
+    for i, k in enumerate(K[1:]):
+
+        # Find the closest and smallest k-1 mixtures
+        closest_indices = np.argsort(distances.T[1])
+
+        assigned = np.zeros(N, dtype=bool)
+        components, concentration = (0, 0)
+        min_sum_log_det = 0
+
+        for index in closest_indices:
+            if np.any(assigned[indices[index, :2]]): continue
+            concentration += 2.0/N * np.sum(np.log(np.var(y[indices[index, :2]],
+                                                          axis=0)))
+            min_sum_log_det += np.sum(np.log(np.var(y[indices[index, :2]], axis=0)))
+
+            assigned[indices[index, :2]] = True
+            components += 1
+
+            if components == k - 1:
+                break
+
+        sum_log_det = np.sum(np.log(np.var(y[~assigned], axis=0)))
+        weight = (N - (k-1) * 2)/float(N)
+        minimum_concentration[i] = concentration + weight * sum_log_det
[…]

### 2018-09-10 gmmmml/mixture_search.py
in progress fixes

@@ -275,6 +275,7 @@ def _mixture_message_length_parts(K, N, D):
     return I_mixtures + I_parameters
 
 
+
 def _mixture_message_length(K, N, D, log_likelihood, slogdetcov, weights=None, 
     yerr=0.001):
     """
@@ -462,11 +463,6 @@ class GaussianMixture(object):
 
         for K in range(1, K_max):
 
-            #model = cluster.KMeans(n_clusters=K)
-            #model.fit(y)
-            #mu = model.cluster_centers_
-            #labels = model.labels_
-
             # Assign everything to the closest thing.            
             mu = _initialize_with_kmeans_pp(y, K)
             labels = np.argmin(scipy.spatial.distance.cdist(mu, y), axis=0)
@@ -603,6 +599,10 @@ class GaussianMixture(object):
         _, I_parts = _mixture_message_length(K, N, D, _z, _z)
         I_other = I_parts["I_mixtures"] + I_parts["I_parameters"]
 
+        # TODO: Move this calculation to one place so we use the same calculation
+        # for actual as we do for predictions.
+
+
         I_slw_c = (0.25 * D * (D + 3) - 0.5)
         I_sldc_c = -0.5 * (D + 2)
             
@@ -612,7 +612,6 @@ class GaussianMixture(object):
 
         nll_strict_bound, nll_relaxed_bound \
             = self._predict_lower_bounds_on_negative_log_likelihood(K, N, D)
-
 
         yerr = 0.001
         I_strict_lower_bound = I_other \
@@ -643,8 +642,9 @@ class GaussianMixture(object):
                 upper=I_other + I_slw_c * slw_upper))
 
             visualization_handler.emit("predict_I_slw",
-                dict(K=K, p_slw=I_other + I_slw_c * p_slw, 
-                    p_slw_err=I_other + I_slw_c * p_slw_err))
+                dict(K=K, 
+                     p_I_slw=I_other + I_slw_c * p_slw, 
+                     p_I_slw_err=I_slw_c * p_slw_err))
 
 
             # Show the information contributions from the sum of the log of the
@@ -658,10 +658,10 @@ class GaussianMixture(object):
             if p_sldc is not None \
             and np.all(np.isfinite(np.hstack([p_sldc, p_sldc_pos_err]))) \
             and np.sum((p_sldc + p_sldc_pos_err) <= (K * np.max(log_det_covs))) >= (K.size - 2):
[…]

### 2018-09-10 gmmmml/visualize.py

@@ -4,8 +4,9 @@ Visualize the search progress for a gaussian mixture model.
 """
 
 import os
-import numpy as np
+import logging as logger
 import matplotlib.pyplot as plt
+import numpy as np
 
 from matplotlib.patches import Ellipse
 from matplotlib.ticker import MaxNLocator
@@ -38,6 +39,13 @@ class VisualizationHandler(object):
         self._y_shape = y.shape
         self._data_projection_indices = data_projection_indices
 
+        self.__plot_bounds_kwds = dict(c="k", lw=2)
+        self.__plot_prediction_kwds = dict(c="tab:blue", lw=2)
+        self.__fill_between_prediction_kwds = dict(facecolor="tab:blue",
+                                                   alpha=0.5, edgecolor=None,
+                                                   zorder=-1)
+
+
         self._figure_path = "" if figure_path is None else figure_path
         self._figure_iter = 1
         self._figure_prefix = os.path.join(self._figure_path,
@@ -50,7 +58,8 @@ class VisualizationHandler(object):
 
         # lists for plotting
         self._plot_model_items = []
-        self._plot_prediction_items = []
+        self._plot_prediction_items = {}
+
 
         return None
 
@@ -194,7 +203,7 @@ class VisualizationHandler(object):
             cov_mask[y_index, y_index] = True
 
             for k in range(K):
-                mu = params["mu"][k][[x_index, y_index]]
+                mu = params["mean"][k][[x_index, y_index]]
                 cov = params["cov"][k][cov_mask].reshape((2, 2))
 
                 vals, vecs = np.linalg.eigh(cov)
@@ -238,6 +247,142 @@ class VisualizationHandler(object):
                 scatter.set_facecolor("k")
                 scatter.set_sizes(30 * np.ones(len(data)))
                 scatter.set_zorder(100)
+
+
+
+
+        elif kind == "maximization":
+
+            self._clear_items(self._plot_model_items)
+
+            ax = self._ax("data")
+            x_index, y_index = self._data_projection_indices
+
+            N, D = self._y_shape
+            K = params["weight"].size
+
+            cov_mask = np.zeros((D, D), dtype=bool)
+            cov_mask[y_index, x_index] = True
+            cov_mask[x_index, y_index] = True
+            cov_mask[x_index, x_index] = True
+            cov_mask[y_index, y_index] = True
+
+
+            for k in range(K):
[…]

### 2018-09-10 sandbox_mixture_search.py

@@ -16,7 +16,7 @@ for i in range(10):
         print(i, j)
         j += 1
         try:
-            y, labels, target, kwds = utils.generate_data(K=30, D=10, center_box=(-1000, 1000))
+            y, labels, target, kwds = utils.generate_data(K=30, D=5, center_box=(-100, 100))
 
         except:
           logging.exception("failed")
@@ -34,7 +34,7 @@ for i in range(10):
     search_model.kmeans_search(y, K_max=kwds["centers"] + 25, 
         visualization_handler=visualization_handler)
 
-
+    """
     K = np.array(search_model._state_K)
 
     ke = int(np.ceil(K.size**0.5))
@@ -52,7 +52,7 @@ for i in range(10):
             verticalalignment="top")
 
     fig.subplots_adjust(wspace=0, hspace=0, top=1.0, bottom=0.0, left=0, right=1)
-
+    """
     
     raise a
 

### 2018-09-10 gmmmml/mixture_search.py
wip

@@ -464,18 +464,7 @@ class GaussianMixture(object):
         for K in range(1, K_max):
 
             # Assign everything to the closest thing.            
-            mu = _initialize_with_kmeans_pp(y, K)
-            labels = np.argmin(scipy.spatial.distance.cdist(mu, y), axis=0)
-
-            # generate repsonsibilities.
-            responsibility = np.zeros((K, N))
-            responsibility[labels, np.arange(N)] = 1.0
-
-            # estimate covariance matrices.
-            cov = _estimate_covariance_matrix_full(y, responsibility, mu)
-
-            # If this is K = 1, then use this as the bound limit for sumlogdetcov
-            weight = responsibility.sum(axis=1)/N
+            mu, cov, weight, responsibility = _initialize_with_kmeans_pp(y, K)
 
             # Do one E-M step.
             try:
@@ -1129,13 +1118,32 @@ def _parameters_per_mixture(D, covariance_type):
         raise ValueError("unknown covariance type '{}'".format(covariance_type))
 
 
-def _initialize_with_kmeans_pp(y, K, random_state=None):
+def _initialize_with_kmeans_pp(y, K, random_state=None, **kwargs):
 
     random_state = check_random_state(random_state)
     squared_norms = row_norms(y, squared=True)
-    return cluster.k_means_._k_init(y, K, x_squared_norms=squared_norms,
+    mu = cluster.k_means_._k_init(y, K, x_squared_norms=squared_norms,
         random_state=random_state)
 
+    labels = np.argmin(scipy.spatial.distance.cdist(mu, y), axis=0)
+
+    # generate repsonsibilities.
+    N, D = y.shape
+    responsibility = np.zeros((K, N))
+    responsibility[labels, np.arange(N)] = 1.0
+
+    # estimate covariance matrices.
+    cov = _estimate_covariance_matrix_full(y, responsibility, mu)
+
+    # If this is K = 1, then use this as the bound limit for sumlogdetcov
+    weight = responsibility.sum(axis=1)/N
+
+    handler = kwargs.get("visualization_handler", None)
+    if handler is not None:
+        handler.emit("actual_I_slw",
+                     dict(K=weight.size, ))
+
+    return (mu, cov, weight, responsibility)
 
 
 

### 2018-09-10 gmmmml/visualize.py

@@ -41,6 +41,8 @@ class VisualizationHandler(object):
 
         self.__plot_bounds_kwds = dict(c="k", lw=2)
         self.__plot_prediction_kwds = dict(c="tab:blue", lw=2)
+        self.__fill_between_bounds_kwds = dict(facecolor="k", alpha=0.3,
+                                               zorder=-1)
         self.__fill_between_prediction_kwds = dict(facecolor="tab:blue",
                                                    alpha=0.5, edgecolor=None,
                                                    zorder=-1)
@@ -300,6 +302,8 @@ class VisualizationHandler(object):
 
             self._plot_prediction_items[kind].extend([
                 ax.plot(K, lower, **self.__plot_bounds_kwds)[0],
+                ax.plot(K, upper, **self.__plot_bounds_kwds)[0],
+                ax.fill_between(K, lower, upper, **self.__fill_between_bounds_kwds),
             ])
 
         elif kind == "predict_I_slw":
@@ -326,7 +330,7 @@ class VisualizationHandler(object):
 
             K, I_slw = params["K"], params["I_slw"]
 
-            self._slogw_data.append([K, I_slw])
+            self._slogw_data.append([K, I_slw[0]])
             scat, data = (self._scatter_slogw_data, self._slogw_data)
 
             data = np.array(data)

### 2018-09-10 gmmmml/mixture_search.py
better predictions for sum log weights

@@ -23,6 +23,7 @@ from sklearn.utils.extmath import row_norms
 
 from collections import defaultdict
 
+from . import mml
 
 logger = logging.getLogger(__name__)
 
@@ -71,178 +72,7 @@ def _total_parameters(K, D):
     return (0.5 * D * (D + 3) * K) + K - 1
 
 
-def _bound_sum_log_weights(K, N):
-    r"""
-    Return the analytical bounds of the function:
 
-    .. math:
-
-        \sum_{k=1}^{K}\log{w_k}
-
-    Where :math:`K` is the number of mixtures, and :math:`w` is a multinomial
-    distribution. The bounded function for when :math:`w` are uniformly
-    distributed is:
-
-    .. math:
-
-        \sum_{k=1}^{K}\log{w_k} \lteq -K\log{K}
-
-    and in the other extreme case, all the weight would be locked up in one
-    mixture, with the remaining :math:`w` values encapsulating the minimum
-    (physically realistic) weight of one data point. In that extreme case,
-    the bound becomes:
-
-    .. math:
-
-        \sum_{k=1}^{K}\log{w_k} \gteq log{(N + 1 - K)} - K\log{N}
-
-    :param K:
-        The number of target Gaussian mixtures.
-
-    :param N:
-        The number of data points.
-
-    :returns:
-        The lower and upper bound on :math:`\sum_{k=1}^{K}\log{w_k}`.
-    """
-
-    upper = -K * np.log(K)
-    lower = np.log(N + 1 - K) - K * np.log(N)
-
-    return (lower, upper)
-
-
-'''
-def _approximate_log_likelihood(N, D, K, logdetcovs, weights, actuals):
-    """
-    Calculate a first-order approximation of the log-likelihood for the
-    :math:`K`-th mixture.
-
-    :param K:
-        The number of target Gaussian mixtures.
-
-    :param N:
-        The number of data points.
-
-    :param D:
-        The dimensionality of the data points.
-
-    :param logdetcovs:
-        The natural logarithm of the determinant of the covariance matrices of
-        the :math:`K`-th mixtures. If :math:`K` is like a list, then the
-        length of `logdetcovs` must match the size of `K`.
-
-    :param weights:
-        The weights of the :math:`K`-th mixtures. 
-
-    """
-
-    K = np.atleast_1d(K)
-    log_likelihoods = np.zeros(K.size)
-
-    assert len(K) == len(logdetcovs)
-    assert len(K) == len(weights)
-
[…]

### 2018-09-10 gmmmml/mml.py

@@ -1,145 +1,27 @@
-
-"""
-Functions to calculate information of quantities.
-"""
-
-import logging
 import numpy as np
-import warnings
-from collections import OrderedDict
-from scipy import optimize as op, interpolate
-from scipy.special import gammaln
-from scipy.stats import gaussian_kde
+import scipy.optimize as op
 
-from .utils import aggregate
+# TODO: move to utils?
 
-def number_of_gmm_parameters(K, D):
-    r"""
-    Return the total number of model parameters :math:`Q`, if a full 
-    covariance matrix structure is assumed, for a gaussian mixture model of
-    :math:`K` mixtures in :math:`D` dimensions.
+def _group_over(x, y, function):
 
-    .. math:
+    x = np.atleast_1d(x)
+    y = np.atleast_1d(y)
 
-        Q = \frac{K}{2}\left[D(D+3) + 2\right] - 1
+    x_unique = np.sort(np.unique(x))
+    y_unique = np.nan * np.ones_like(x_unique)
 
-    :param K:
-        The number of Gaussian mixtures.
+    for i, xi in enumerate(x_unique):
+        match = (x == xi)
+        y_unique[i] = function(y[match])
 
-    :param D:
-        The dimensionality of the data.
-
-    :returns:
-        The total number of model parameters, :math:`Q`.
-    """
-
-    return (0.5 * D * (D + 3) * K) + K - 1
+    return (x_unique, y_unique)
 
 
-def _gmm_parameter_message_length(K, N, D):
-    """
-    Retrun the message length required to encode some of the easier parameters
-    of the mixture.
-
-    # TODO: Update docs.
-
-    :param K:
-        The number of Gaussian components in the mixture.
-
-    :param N:
-        The number of data points.
-
-    :param D:
-        The dimensionality of the data.
-    """
-
-    # TODO: update docs.
-    #print("update docs for _gmm_parameter_message_length")
-
-    Q = number_of_gmm_parameters(K, D)
-
-    # Calculate information required to encode the mixture parameters,
-    # regardless of the covariance matrices and the weights, etc.
-    I_mixtures = K * np.log(2) * (1 - D/2.0) + gammaln(K) \
-               + 0.25 * (2.0 * (K - 1) + K * D * (D + 3)) * np.log(N)
-    I_parameters = 0.5 * np.log(Q * np.pi) - 0.5 * Q * np.log(2 * np.pi)
-
[…]

### 2018-09-10 gmmmml/visualize.py

@@ -17,7 +17,7 @@ from . import mml
 class VisualizationHandler(object):
 
     def __init__(self, y, data_projection_indices=(0, 1), figure_path=None,
-        target=None, **kwargs):
+        target=None, colours=None, **kwargs):
         r"""
         Initialize a visualisation handler to show the search progress and the
         predictions of future mixtures.
@@ -34,33 +34,44 @@ class VisualizationHandler(object):
 
         :param target: [optional]
             The target message length components (e.g., for generated data).
+
+        :param colours: [optional]
+            A dictionary containing the colours for these items:
+
+                data
+                predictions
+                bounds
+
         """
 
         self._y_shape = y.shape
         self._data_projection_indices = data_projection_indices
 
-        self.__plot_bounds_kwds = dict(c="k", lw=2)
-        self.__plot_prediction_kwds = dict(c="tab:blue", lw=2)
-        self.__fill_between_bounds_kwds = dict(facecolor="k", alpha=0.3,
-                                               zorder=-1)
-        self.__fill_between_prediction_kwds = dict(facecolor="tab:blue",
-                                                   alpha=0.5, edgecolor=None,
-                                                   zorder=-1)
+        if colours is None:
+            colours = dict(bounds="#666666",
+                           predictions="tab:blue",
+                           model="tab:blue",
+                           data="#000000")
+        self._colours = colours
 
 
         self._figure_path = "" if figure_path is None else figure_path
+        if not os.path.exists(os.path.dirname(self._figure_path)):
+            os.makedirs(os.path.dirname(self._figure_path), exist_ok=True)
+
         self._figure_iter = 1
         self._figure_prefix = os.path.join(self._figure_path,
             "iter_{:.0f}".format(np.random.uniform(0,  10000)))
 
         self._fig, axes = plt.subplots(1, 5, figsize=(15.6 + 2.8, 2.8))
+
+        # lists for plotting
+        self._plot_items = {}
+
             
         self._init_axes(y, target)
         self.snapshot()
 
-        # lists for plotting
-        self._plot_model_items = []
-        self._plot_prediction_items = {}
 
 
         return None
[…]

### 2018-09-10 gmmmml/mixture_search.py
use a GP for slogdetcov predictions

@@ -241,6 +241,7 @@ class GaussianMixture(object):
         # Lists to record states for predictive purposes.
         self._state_K = []
         self._state_det_covs = []
+        self._state_sum_log_det_covs = []
         self._state_weights = []
         self._state_slog_weights = []
         self._state_slog_likelihoods = []
@@ -303,6 +304,9 @@ class GaussianMixture(object):
             handler.emit("actual_I_weights", 
                          dict(K=weight.size, I=message_length["I_weights"]))
 
+            handler.emit("actual_I_slogdetcovs", 
+                         dict(K=weight.size, I=message_length["I_slogdetcovs"]))
+
 
         return (R, ll, message_length)
 
@@ -338,7 +342,8 @@ class GaussianMixture(object):
         kwds = {**self._em_kwds, **kwargs}
         handler = kwds.get("visualization_handler", None)
 
-        for K in range(1, K_max):
+        #for K in range(1, K_max):
+        for K in np.random.choice(np.arange(1, 1 + K_max), K_max, replace=False):
 
             # Assign everything to the closest thing.            
             means, covs, weights, responsibilities \
@@ -454,11 +459,19 @@ class GaussianMixture(object):
 
         K = np.atleast_1d(K)
 
+        # Sum of the log of the weights.
         I_sum_log_weights, I_sum_log_weights_var, \
         I_sum_log_weights_lower, I_sum_log_weights_upper \
             = mml.predict_information_of_sum_log_weights(
-                K, N, D,
-                data=(self._state_K, self._state_slog_weights))
+                K, N, D, data=(self._state_K, self._state_slog_weights))
+
+
+        # Sum of the log of the determinant of the covariance matrices.
+        I_sum_log_det_covs, I_sum_log_det_covs_var \
+            = mml.predict_information_of_sum_log_det_covs(
+                K, D, data=(self._state_K, self._state_sum_log_det_covs))
+
+
 
         handler = kwargs["visualization_handler"]
         if handler is not None:
@@ -466,6 +479,8 @@ class GaussianMixture(object):
                 K=K, I=I_sum_log_weights, I_var=I_sum_log_weights_var,
                 I_lower=I_sum_log_weights_lower, I_upper=I_sum_log_weights_upper))
 
+            handler.emit("predict_I_slogdetcovs", dict(
+                K=K, I=I_sum_log_det_covs, I_var=I_sum_log_det_covs_var))
 
 
         """
[…]

### 2018-09-10 gmmmml/mml.py

@@ -1,6 +1,9 @@
 import numpy as np
 import scipy.optimize as op
 
+import george
+from george import (kernels, modeling)
+
 # TODO: move to utils?
 
 def _group_over(x, y, function):
@@ -43,9 +46,8 @@ def _bounds_of_sum_log_weights(K, N):
     the bound becomes:
 
     .. math:
-
-
-
+    
+        \sum_{k=1}^{K}\log{w_k} \gteq (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
 
     :param K:
         The number of target Gaussian mixtures.
@@ -56,10 +58,8 @@ def _bounds_of_sum_log_weights(K, N):
     :returns:
         The lower and upper bound on :math:`\sum_{k=1}^{K}\log{w_k}`.
     """
-
     upper = -K * np.log(K)
     lower = (K - 1) * np.log(2) - K * np.log(N) + np.log(N - 2 * K + 2)
-
     return (lower, upper)
 
 
@@ -72,7 +72,7 @@ def information_bounds_of_sum_log_weights(K, N, D):
 
     .. math:
 
-        I_w \gteq C\log{(N + 1 - K)} - K\log{N}
+        I_w \gteq C\left[(K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}\right]
 
     and the upper bound is given when the weights are uniformly distributed
 
@@ -101,20 +101,34 @@ def information_bounds_of_sum_log_weights(K, N, D):
         A two-length tuple with the lower and upper information bounds on the
         sum of the log of the mixing weights, in units of nats.
     """
-
     return information_of_sum_log_weights(_bounds_of_sum_log_weights(K, N), D)
 
 
 def information_of_sum_log_weights(sum_log_weights, D):
     r"""
-    Return the information content on the sum of the log of the weights.
+    Return the information content on the sum of the log of the weights:
 
-    # TODO:
+    .. math:
+
+        I_w = C\sum_{k=1}^{K}\log{w_k}
+
+    where 
+
+    .. math:
+
+        C = \frac{D(D+3)}{4} - \frac{1}{2}
+
+    :param sum_log_weights:
+        The sum of the log of the weights for a Gaussian mixture.
+
+    :param D:
+        The dimensionality of the data.
+
+    :returns:
+        The information content of the sum of the log of the weights, in units
+        of nats.
     """
-
[…]

### 2018-09-10 gmmmml/visualize.py

@@ -109,11 +109,13 @@ class VisualizationHandler(object):
         ax.yaxis.set_major_locator(MaxNLocator(5))
         self._scatter_actual_sum_log_weights = ax.scatter([np.nan], [np.nan])
 
-        ax = self._ax("slogdet")
+        ax = self._ax("sum_log_det_covs")
         ax.set_xlabel(r"$K$")
         ax.set_ylabel(r"$-\frac{(D+2)}{2}\sum\log{|C_k|}$ $[{\rm nats}]$")
         ax.xaxis.set_major_locator(MaxNLocator(5))
         ax.yaxis.set_major_locator(MaxNLocator(5))
+        self._actual_sum_log_det_covs = []
+        self._scatter_actual_sum_log_det_covs = ax.scatter([np.nan], [np.nan])
 
         ax = self._ax("nll")
         ax.set_xlabel(r"$K$")
@@ -148,7 +150,7 @@ class VisualizationHandler(object):
         """
 
         axes = np.array(self._fig.axes).flatten()
-        index = ["data", "sum_log_weights", "slogdet", "nll", "I"].index(descriptor)
+        index = ["data", "sum_log_weights", "sum_log_det_covs", "nll", "I"].index(descriptor)
         return axes[index]
 
 
@@ -254,7 +256,7 @@ class VisualizationHandler(object):
 
             K, I = params["K"], params["I"]
 
-            self._actual_sum_log_weights.append([K, I])
+            self._actual_sum_log_weights.append(np.hstack([K, I]))
 
             scat, data = (self._scatter_actual_sum_log_weights, self._actual_sum_log_weights)
 
@@ -263,6 +265,43 @@ class VisualizationHandler(object):
             scat.set_facecolor(self._colours["data"])
             scat.set_sizes(30 * np.ones(len(data)))
             scat.set_zorder(100)
+
+            #_rescale_based_on_data(self._ax("sum_log_weights"), *data.T)
+            
+
+
+        elif kind == "actual_I_slogdetcovs":
+
+
+            K, I = params["K"], params["I"]
+
+            self._actual_sum_log_det_covs.append(np.hstack([K, I]))
+
+            scat, data = (self._scatter_actual_sum_log_det_covs, self._actual_sum_log_det_covs)
+
+            data = np.array(data)
+            scat.set_offsets(data)
+            scat.set_facecolor(self._colours["data"])
[…]

### 2018-09-10 article/ms.tex
correct bounds on slogw

@@
     \item The weights are distributed such that $K - 1$ components each 
-          describe just one data point $w_{1..K-1} = 1/N$, and the remaining 
-          component describes all other data points $w_{K} = 1 - (K-1)/N$
+          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
+          component describes all other data points $w_{K} = 1 - (K-1)\frac{2}{N}$
           as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
-          \end{equation}
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
+          %\end{equation}
+          \begin{equation}
+            \sum_{k=1}^{K}\log\weight_k = (K - 1)\log\frac{2}{N} + \log\left(1 - \frac{2(K - 1)}{N}\right)
+          \end{equation}
+
           \noindent{}and by making use of the identity 
@@
           \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
-          \end{equation}
+            \sum_{k=1}^{K}\log\weight_k = (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
+          \end{equation}
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
+          %\end{equation}
 \end{enumerate}
@@
 \begin{equation}
--K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad - K\log{K} + \log(N + 1 - K)
+-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
 \end{equation}

### 2018-09-10 gmmmml/mixture_search.py
minor

@@ -465,11 +465,13 @@ class GaussianMixture(object):
             = mml.predict_information_of_sum_log_weights(
                 K, N, D, data=(self._state_K, self._state_slog_weights))
 
-
         # Sum of the log of the determinant of the covariance matrices.
         I_sum_log_det_covs, I_sum_log_det_covs_var \
             = mml.predict_information_of_sum_log_det_covs(
                 K, D, data=(self._state_K, self._state_sum_log_det_covs))
+
+        # Negative log-likelihood.
+
 
 
 

### 2018-09-10 gmmmml/mml.py

@@ -262,9 +262,11 @@ def predict_information_of_sum_log_det_covs(K, D, data):
     x, y = _group_over(data[0], data[1], np.mean)
     _, yerr = _group_over(data[0], data[1], np.std)
 
+    yerr = np.clip(yerr, 0, np.inf)
+
     kernel = np.var(y) * kernels.LinearKernel(order=1, log_gamma2=1) \
            + np.var(y) * kernels.ExpSquaredKernel(1)
-        
+
     gp = george.GP(kernel=kernel, 
                    white_noise=np.log(np.std(y)), fit_white_noise=True)
                    
@@ -277,18 +279,11 @@ def predict_information_of_sum_log_det_covs(K, D, data):
         gp.set_parameter_vector(p)
         return -gp.grad_log_likelihood(y, quiet=True)
 
-
-    gp.compute(x.astype(float), yerr=np.ones_like(x))
-
-    #print("Initial \log{{L}} = {:.2f}".format(gp.log_likelihood(y)))
-    #print("Initial \grad\log{{L}} = {}".format(gp.grad_log_likelihood(y)))
+    gp.compute(x.astype(float), yerr=yerr)
 
     p0 = gp.get_parameter_vector()
 
-    gp_result = op.minimize(nll, p0, jac=grad_nll, method="L-BFGS-B", tol=1e-8)
-
-    #print("Final \log{{L}} = {:.2f}".format(gp.log_likelihood(y)))
-    #print("Final \grad\log{{L}} = {}".format(gp.grad_log_likelihood(y)))
+    gp_result = op.minimize(nll, p0, jac=grad_nll, method="L-BFGS-B")
 
     pred, pred_var = gp.predict(y, K, return_var=True)
 

### 2018-09-10 article/ms.tex
merge

--- 2018-04-20 article/ms.tex
@@
           &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
-    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad {\textrm and},\\
-    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\sigma_{\datum}} \quad .
+    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad \textrm{and},\\
+    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\mathcal{E}} \quad .
 \end{eqnarray}
@@
     \item The weights are distributed such that $K - 1$ components each 
-          describe just one data point $w_{1..K-1} = 1/N$, and the remaining 
-          component describes all other data points $w_{K} = 1 - (K-1)/N$
+          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
+          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
           as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
           \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
+                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
+                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
           \end{equation}
           \noindent{}and by making use of the identity 
-            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$
-          we can simplify this expression to:
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
+            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
           \end{equation}
@@
 derived the bounds
-\begin{equation}
--K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad - K\log{K} + \log(N + 1 - K)
-\end{equation}
+\begin{eqnarray}
+\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
+\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
+\end{eqnarray}
 \noindent{}which gives us analytic expressions for the upper and lower value 
@@
 to the smallest realistic values, where those mixtures encapsulate only two
-data points and $\todo{w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N}$:
-
-\begin{eqnarray}
--N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& -N\left[(K-1)\left(\frac{1}{N}\log\frac{1}{N}\right) + \left(1 - \frac{K-1}{N}\right)\log\left(1 - \frac{K-1}{N}\right)\right] \nonumber \\
--N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& (K - 1)\log{N} + (K - 1 - N)\log\left(1-\frac{K-1}{N}\right) \nonumber \\
--N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& N\log{N} + (K - 1 - N)\log\left(N-K+1\right) \quad .
-\end{eqnarray}
+data points and $w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N$ such that
+%\begin{eqnarray}
+%	\sum_{k=1}^{K}\weight_{k}\log\weight_k = \left[(K-1)\left(\frac{2}{N}\log{\frac{2}{N}}\right) + \left(1-\frac{2(K-1)}{N}\right)\log{\left(1-\frac{2(K-1)}{N}\right)}\right] \nonumber
+%\end{eqnarray}`
+%\noindent{}and
+\begin{eqnarray}
+	-N\sum_{k=1}^{K}\weight_{k}\log\weight_k \geq 2(1-K)\log{2} + N\log{N} - (N-2K+2)\log{\left(N - 2K + 2\right)}
+\end{eqnarray}
+%  log(a-b) = log(a) + log(1 - b/a)
+% log(1 - b/a) = log(a-b) - log(a)
+% log(1 - b/a) = log(N - 2K + 2) - log(N)
+
+% 2(1-K)log(2) - 2(1-K)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
+% 2(1-K)log(2) - 2log(N) + 2Klog(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
+% 2(1-K)log(2) + (2K - 2)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - 2 - N)log(N)
+% 2(1-K)log(2) + Nlog(N) + (2K - N - 2)log(N - 2K + 2)
 

### 2018-09-10 gmmmml/mixture_search.py
gp predictions for all

@@ -240,6 +240,7 @@ class GaussianMixture(object):
 
         # Lists to record states for predictive purposes.
         self._state_K = []
+        self._state_I = []
         self._state_det_covs = []
         self._state_sum_log_det_covs = []
         self._state_weights = []
@@ -296,7 +297,7 @@ class GaussianMixture(object):
         R, ll, message_length = _expectation(y, mu, cov, weight, **kwds)
 
         # Record state for predictions.
-        self._record_state_for_predictions(cov, weight, ll)
+        self._record_state_for_predictions(cov, weight, ll, message_length)
 
         # Do visualization stuff.
         handler = kwargs.get("visualization_handler", None)
@@ -306,6 +307,12 @@ class GaussianMixture(object):
 
             handler.emit("actual_I_slogdetcovs", 
                          dict(K=weight.size, I=message_length["I_slogdetcovs"]))
+
+            handler.emit("actual_I_data", 
+                         dict(K=weight.size, I=-np.sum(ll)))
+
+            handler.emit("actual_I",
+                dict(K=weight.size, I=np.sum(np.hstack(message_length.values()))))
 
 
         return (R, ll, message_length)
@@ -342,9 +349,10 @@ class GaussianMixture(object):
         kwds = {**self._em_kwds, **kwargs}
         handler = kwds.get("visualization_handler", None)
 
-        #for K in range(1, K_max):
-        for K in np.random.choice(np.arange(1, 1 + K_max), K_max, replace=False):
+        #for K in np.random.choice(np.arange(1, 1 + K_max), K_max, replace=False):
 
+        for K in range(1, K_max):
+        
             # Assign everything to the closest thing.            
             means, covs, weights, responsibilities \
                 = self.initialize(y, K, **kwargs)
@@ -466,14 +474,18 @@ class GaussianMixture(object):
                 K, N, D, data=(self._state_K, self._state_slog_weights))
 
         # Sum of the log of the determinant of the covariance matrices.
-        I_sum_log_det_covs, I_sum_log_det_covs_var \
+        I_sum_log_det_covs, I_sum_log_det_covs_var, I_sum_log_det_covs_lower \
             = mml.predict_information_of_sum_log_det_covs(
-                K, D, data=(self._state_K, self._state_sum_log_det_covs))
+                K, D, data=(self._state_K, self._state_det_covs))
 
         # Negative log-likelihood.
+        I_data, I_data_var \
[…]

### 2018-09-10 gmmmml/mml.py

@@ -218,8 +218,8 @@ def information_of_sum_log_det_covs(sum_log_det_covs, D):
     :param D:
         The dimensionality of the data.
 
-    :returns:
 
+    :returns:
     """
     return -0.5 * (D + 2) * sum_log_det_covs
 
@@ -242,32 +242,41 @@ def predict_information_of_sum_log_det_covs(K, D, data):
 
     :param data:
         A two-length tuple containing (1) the number of components in previously
-        trialled Gaussian mixtures, and the sum of the log of the determinant of
-        their covariance matrices.
+        trialled Gaussian mixtures, and the determinant of the
+        covariance matrices in each mixture.
 
     :returns:
-        A two-length tuple containing:
+        A three-length tuple containing:
 
             (1) the predicted information content of the sum of the log of the
                 determinant of the covariance matrices for each :math:`K`
                 mixture;
 
             (2) the variance on the predicted information content on the sum of
-                the log of the determinant of the covariance matrices.
+                the log of the determinant of the covariance matrices
+
+            (3) the lower bound on the information content on the sum of the
+                log of the determinant of the covariance matrices. This bound
+                is not a theoretical bound: it is a lower bound based on the
+                smallest determinant of the covariance matrices.
     """
 
     if data is None:
         raise NotImplementedError("cannot predict this theoretically")
 
-    x, y = _group_over(data[0], data[1], np.mean)
-    _, yerr = _group_over(data[0], data[1], np.std)
+    slogdetcovs = np.array([np.sum(dc) for dc in data[1]])
+    x, y = _group_over(data[0], slogdetcovs, np.mean)
+    _, yerr = _group_over(data[0], slogdetcovs, np.std)
 
-    yerr = np.clip(yerr, 0, np.inf)
+    yerr = np.clip(yerr, 1, np.inf)
 
-    kernel = np.var(y) * kernels.LinearKernel(order=1, log_gamma2=1) \
-           + np.var(y) * kernels.ExpSquaredKernel(1)
+    kernel = np.var(y) * kernels.ExpSquaredKernel(1) \
+           + np.var(y) * kernels.LinearKernel(log_gamma2=0, order=1)
+
[…]

### 2018-09-10 gmmmml/visualize.py

@@ -117,17 +117,21 @@ class VisualizationHandler(object):
         self._actual_sum_log_det_covs = []
         self._scatter_actual_sum_log_det_covs = ax.scatter([np.nan], [np.nan])
 
-        ax = self._ax("nll")
+        ax = self._ax("negative_sum_log_likelihood")
         ax.set_xlabel(r"$K$")
         ax.set_ylabel(r"$-\sum\log{\mathcal{L}(y\|\theta)}$ $[{\rm nats}]$")
         ax.xaxis.set_major_locator(MaxNLocator(5))
         ax.yaxis.set_major_locator(MaxNLocator(5))
+        self._actual_negative_log_likelihoods = []
+        self._scatter_actual_negative_log_likelihoods = ax.scatter([np.nan], [np.nan])
 
+        self._actual_message_lengths = []
         ax = self._ax("I")
         ax.set_xlabel(r"$K$")
         ax.set_ylabel(r"$I$ $[{\rm nats}]$")
         ax.xaxis.set_major_locator(MaxNLocator(5))
         ax.yaxis.set_major_locator(MaxNLocator(5))
+        self._scatter_actual_message_lengths = ax.scatter([np.nan], [np.nan])
 
         if target is not None:
             print("no target implemented  yet")
@@ -150,7 +154,8 @@ class VisualizationHandler(object):
         """
 
         axes = np.array(self._fig.axes).flatten()
-        index = ["data", "sum_log_weights", "sum_log_det_covs", "nll", "I"].index(descriptor)
+        index = ["data", "sum_log_weights", "sum_log_det_covs", 
+                 "negative_sum_log_likelihood", "I"].index(descriptor)
         return axes[index]
 
 
@@ -294,6 +299,49 @@ class VisualizationHandler(object):
             self._clear_items(self._plot_items[ax])
             
             K, I, I_var = (params["K"], params["I"], params["I_var"])
+            I_lower = params["I_lower"]
+
+            prediction_colour = self._colours["predictions"]
+
+            self._plot_items[ax].extend([
+                ax.plot(K, I, lw=2, c=prediction_colour)[0],
+                ax.fill_between(K, I - np.sqrt(I_var), I + np.sqrt(I_var),
+                                facecolor=prediction_colour, alpha=0.3, zorder=-1),
+            ])
+            """
+                ax.fill_between(K, 
+                                I_lower, 
+                                np.max(I + np.sqrt(I_var)) * np.ones_like(I_lower),
+                                facecolor=self._colours["bounds"], alpha=0.3,
[…]

### 2018-09-10 sandbox_mixture_search.py

@@ -3,7 +3,7 @@ import numpy as np
 from gmmmml import (mixture_search as mixture, visualize, utils)
 import  scipy.optimize as op
 
-np.random.seed(11)
+np.random.seed(43)
 
 # Seed 1100 is a fun one: K = 103, D = 5, N = 4249
 
@@ -16,7 +16,7 @@ for i in range(10):
         print(i, j)
         j += 1
         try:
-            y, labels, target, kwds = utils.generate_data(K=30, D=5, center_box=(-100, 100))
+            y, labels, target, kwds = utils.generate_data(K=40, D=10, center_box=(-10, 10))
 
         except:
           logging.exception("failed")

### 2018-09-10 gmmmml/mixture_search.py
visualisation, etc

@@ -166,7 +166,9 @@ def _mixture_message_length(K, N, D, log_likelihood, slogdetcov, weights=None,
         + 0.25 * (2.0 * (K - 1) + K * D * (D + 3)) * np.log(N)
     I_parameters = 0.5 * np.log(Q * np.pi) - 0.5 * Q * np.log(2 * np.pi)
     
-    I_data = -log_likelihood - D * N * np.log(yerr)
+    I_data = -log_likelihood 
+    # TODO: this requires a thinko
+    #- D * N * np.log(yerr)
     I_slogdetcovs = -0.5 * (D + 2) * slogdetcov
     I_weights = (0.25 * D * (D + 3) - 0.5) * slogw
 
@@ -212,7 +214,7 @@ class GaussianMixture(object):
 
     def __init__(self, covariance_type="full", covariance_regularization=0, 
         threshold=1e-5, max_em_iterations=10000, visualization_handler=None,
-        **kwargs):
+        predict=25, **kwargs):
 
         available = ("full", )
         covariance_type = covariance_type.strip().lower()
@@ -247,9 +249,6 @@ class GaussianMixture(object):
         self._state_slog_weights = []
         self._state_slog_likelihoods = []
 
-        self._state_predictions_K = []
-        self._state_predictions_slog_det_covs = []
-        self._state_predictions_slog_likelihoods = []
         self._state_meta = {}
 
         return None
@@ -349,7 +348,6 @@ class GaussianMixture(object):
         kwds = {**self._em_kwds, **kwargs}
         handler = kwds.get("visualization_handler", None)
 
-        #for K in np.random.choice(np.arange(1, 1 + K_max), K_max, replace=False):
 
         for K in range(1, K_max):
         
@@ -467,6 +465,9 @@ class GaussianMixture(object):
 
         K = np.atleast_1d(K)
 
+        # Constant terms.
+        I_other = mml.information_of_mixture_constants(K, N, D)
+
         # Sum of the log of the weights.
         I_sum_log_weights, I_sum_log_weights_var, \
         I_sum_log_weights_lower, I_sum_log_weights_upper \
@@ -474,7 +475,8 @@ class GaussianMixture(object):
                 K, N, D, data=(self._state_K, self._state_slog_weights))
 
         # Sum of the log of the determinant of the covariance matrices.
-        I_sum_log_det_covs, I_sum_log_det_covs_var, I_sum_log_det_covs_lower \
+        I_sum_log_det_covs, I_sum_log_det_covs_var, \
[…]

### 2018-09-10 gmmmml/mml.py

@@ -3,6 +3,7 @@ import scipy.optimize as op
 
 import george
 from george import (kernels, modeling)
+from scipy.special import gammaln
 
 # TODO: move to utils?
 
@@ -22,6 +23,67 @@ def _group_over(x, y, function):
 
 
 
+
+def gmm_number_of_parameters(K, D):
+    r"""
+    Return the total number of model parameters :math:`Q`, if a full 
+    covariance matrix structure is assumed.
+
+    .. math:
+
+        Q = \frac{K}{2}\left[D(D+3) + 2\right] - 1
+
+
+    :param K:
+        The number of Gaussian mixtures.
+
+    :param D:
+        The dimensionality of the data.
+
+    :returns:
+        The total number of model parameters, :math:`Q`.
+    """
+    return (0.5 * D * (D + 3) * K) + K - 1
+
+
+
+def information_of_mixture_constants(K, N, D):
+    r"""
+    Return the message length, or information, to encode the additional parameters
+    for a Gaussian mixture model. Specifically this returns the sum of the
+    information to encode the number of parameters and the constant terms
+    related to the mixture itself:
+
+    .. math:
+
+        I_{mixture} = (1 - \frac{D}{2})K\log{2} + \Gamma\log{K} 
+                    + \frac{1}{4}(2(K - 1) + KD(D+3))\log{N}
+
+        I_{parameters} = \frac{1}{2}\log{Q\pi} - \frac{Q}{2}\log{2\pi}
+
+    where :math:`Q` is the total number of model parameters
+
+    .. math:
+
+        Q = \frac{KD(D + 3)}{2} + K - 1
+
+    :param K:
+        The number of components in the target Gaussian mixture.
+
+    :param N:
+        The number of data points.
+
+    :param D:
+        The dimensionality of the data.
+    """
+
+    Q = gmm_number_of_parameters(K, D)
+
+    I_mixtures = K * np.log(2) * (1 - D/2.0) + gammaln(K) \
+        + 0.25 * (2.0 * (K - 1) + K * D * (D + 3)) * np.log(N)
+    I_parameters = 0.5 * np.log(Q * np.pi) - 0.5 * Q * np.log(2 * np.pi)
+
+    return I_mixtures + I_parameters
 
 
 def _bounds_of_sum_log_weights(K, N):
[…]

### 2018-09-10 gmmmml/visualize.py

@@ -5,6 +5,7 @@ Visualize the search progress for a gaussian mixture model.
 
 import os
 import logging as logger
+import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 
@@ -13,6 +14,15 @@ from matplotlib.ticker import MaxNLocator
 
 from . import mml
 
+mpl_style = {
+    "font.size": 12.0,
+    "text.usetex": True,
+    "text.latex.preamble": r"\usepackage{amsmath}",
+    "text.latex.preview": True,
+    "axes.unicode_minus": False
+}
+
+matplotlib.style.use(mpl_style)
 
 class VisualizationHandler(object):
 
@@ -100,18 +110,18 @@ class VisualizationHandler(object):
         ax.set_ylabel(r"$x_{{{0}}}$".format(y_index))
         ax.xaxis.set_major_locator(MaxNLocator(3))
         ax.yaxis.set_major_locator(MaxNLocator(3))
-        
+         
         self._actual_sum_log_weights = []
         ax = self._ax("sum_log_weights")
         ax.set_xlabel(r"$K$")
-        ax.set_ylabel(r"$\left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum\log{w_k}$ $[{\rm nats}]$")
+        ax.set_ylabel(r"$\left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum\log{w_k} \textrm{ / nats}$")
         ax.xaxis.set_major_locator(MaxNLocator(5))
         ax.yaxis.set_major_locator(MaxNLocator(5))
         self._scatter_actual_sum_log_weights = ax.scatter([np.nan], [np.nan])
 
         ax = self._ax("sum_log_det_covs")
         ax.set_xlabel(r"$K$")
-        ax.set_ylabel(r"$-\frac{(D+2)}{2}\sum\log{|C_k|}$ $[{\rm nats}]$")
+        ax.set_ylabel(r"$-\frac{(D+2)}{2}\sum\log{|C_k|} \textrm{ / nats}$")
         ax.xaxis.set_major_locator(MaxNLocator(5))
         ax.yaxis.set_major_locator(MaxNLocator(5))
         self._actual_sum_log_det_covs = []
@@ -119,24 +129,20 @@ class VisualizationHandler(object):
 
         ax = self._ax("negative_sum_log_likelihood")
[…]

### 2018-09-11 gmmmml/mixture_search.py
lower bounds and things

@@ -243,10 +243,9 @@ class GaussianMixture(object):
         # Lists to record states for predictive purposes.
         self._state_K = []
         self._state_I = []
-        self._state_det_covs = []
-        self._state_sum_log_det_covs = []
         self._state_weights = []
         self._state_slog_weights = []
+        self._state_det_covs = []
         self._state_slog_likelihoods = []
 
         self._state_meta = {}
@@ -311,7 +310,8 @@ class GaussianMixture(object):
                          dict(K=weight.size, I=-np.sum(ll)))
 
             handler.emit("actual_I",
-                dict(K=weight.size, I=np.sum(np.hstack(message_length.values()))))
+                dict(K=weight.size, I=np.sum(np.hstack(message_length.values()))),
+                snapshot=kwds.get("__snapshot_every_em_step", False))
 
 
         return (R, ll, message_length)
@@ -323,7 +323,7 @@ class GaussianMixture(object):
         means, covs, weights = _maximization(y, means, covs, weights, 
                                              responsibilities, **kwds)
 
-        # Make predictions?
+        # TODO: Make predictions?
 
         # Do visualization stuff.
         handler = kwargs.get("visualization_handler", None)
@@ -331,7 +331,6 @@ class GaussianMixture(object):
             handler.emit("maximization", dict(means=means,
                                               covs=covs,
                                               weights=weights))
-
 
         return (means, covs, weights)
 
@@ -351,26 +350,33 @@ class GaussianMixture(object):
 
         for K in range(1, K_max):
         
-            # Assign everything to the closest thing.            
-            means, covs, weights, responsibilities \
-                = self.initialize(y, K, **kwargs)
+            # Assign everything to the closest thing.
+            means, covs, weights, responsibilities = self.initialize(y, K, **kwargs)
 
-            # Do one E-M step.
-            try:
-                responsibilities, ll, message_length \
-                    = self.expectation(y, means, covs, weights, **kwargs)
+            prev_I = np.inf
+            for i in range(self.max_em_iterations):
 
+                means, covs, weights = self.maximization(y, means, covs, weights,
+                                                         responsibilities, **kwargs)
 
-            except ValueError:
[…]

### 2018-09-11 gmmmml/mml.py

@@ -330,28 +330,24 @@ def predict_information_of_sum_log_det_covs(K, D, data):
     x, y = _group_over(data[0], slogdetcovs, np.mean)
     _, yerr = _group_over(data[0], slogdetcovs, np.std)
 
+    # TODO
+    y = y*x
+
     yerr = np.clip(yerr, 1, np.inf)
 
     var_y = np.var(y)
     var_y = var_y if (np.isfinite(var_y) and var_y > 0) else 1
 
-    kernel = var_y * kernels.ExpSquaredKernel(1) #\
-   #+ np.var(y) * kernels.LinearKernel(log_gamma2=0, order=1)
+    kernel = var_y * kernels.ExpSquaredKernel(1)
+    #       + var_y * kernels.LocalGaussianKernel(location=0, log_width=0)
+    #+ np.var(y) * kernels.LinearKernel(log_gamma2=0, order=1)
 
     #(y) * kernels.LocalGaussianKernel(location=0, log_width=0) \
-
-    class MeanModel(modeling.Model):
-
-        parameter_names = ("a", "b", "c")
-
-        def get_value(self, k):
-            k = k.flatten()
-            return self.a/(k - self.b) +  self.c
 
     white_noise = np.log(np.sqrt(var_y))
 
     gp = george.GP(kernel=kernel, 
-                   mean=MeanModel(a=1, b=0, c=np.mean(y)), fit_mean=True,
+                   mean=np.mean(y), fit_mean=True,
                    white_noise=white_noise, fit_white_noise=True)
                   
     gp.compute(x.astype(float), yerr=yerr)
@@ -373,6 +369,10 @@ def predict_information_of_sum_log_det_covs(K, D, data):
 
     pred, pred_var = gp.predict(y, K, return_var=True)
 
+    # TODO
+    pred /= K
+    pred_var /= K**2
+
     I = information_of_sum_log_det_covs(pred, D)
     I_var = information_of_sum_log_det_covs(np.sqrt(pred_var), D)**2
 
@@ -389,7 +389,7 @@ def predict_information_of_sum_log_det_covs(K, D, data):
     return (I, I_var, I_lower, I_upper)
 
 
-def predict_negative_sum_log_likelihood(K, data):
+def predict_negative_sum_log_likelihood(K, N, D, data):
     r"""
     Predict the negative sum of the log likelihood for a Gaussian mixture model
     with :math:`K` components.
[…]

### 2018-09-11 gmmmml/visualize.py

@@ -73,7 +73,8 @@ class VisualizationHandler(object):
         self._figure_prefix = os.path.join(self._figure_path,
             "iter_{:.0f}".format(np.random.uniform(0,  10000)))
 
-        self._fig, axes = plt.subplots(1, 5, figsize=(15.6 + 2.8, 2.8))
+
+        self._fig, axes = plt.subplots(1, 5, figsize=(15.5, 3))
 
         # lists for plotting
         self._plot_items = {}
@@ -378,7 +379,10 @@ class VisualizationHandler(object):
             self._clear_items(self._plot_items[ax])
             
             K, I, I_var = (params["K"], params["I"], params["I_var"])
+            I_lower = params["I_lower"]
+            I_upper = np.nan
 
+            bound_colour = self._colours["bounds"]
             prediction_colour = self._colours["predictions"]
 
             self._plot_items[ax].extend([
@@ -386,6 +390,41 @@ class VisualizationHandler(object):
                 ax.fill_between(K, I - np.sqrt(I_var), I + np.sqrt(I_var),
                                 facecolor=prediction_colour, alpha=0.3, zorder=-1),
             ])
+
+            v = np.vstack([
+                np.vstack([K, I_lower]).T,
+                np.array(self._actual_negative_log_likelihoods)
+            ])
+
+            _rescale_based_on_data(ax, *v.T)
+
+
+            # Adjust K so that it extends the region we want.
+            if K.size > 1:
+                Ks = np.hstack([K[0] - 0.5, K[1:-1], K[-1] + 0.5])
+            else:
+                Ks = K
+
+            set_upper = not np.all(np.isfinite(I_upper))
+
+            if set_upper:
+                plt.draw()
+                ylim = ax.get_ylim()
+                I_upper = np.max(ylim) * np.ones_like(I_lower)
+
+            self._plot_items[ax].extend([
+                ax.plot(K, I_lower, c=bound_colour, lw=1)[0],
+                ax.plot(K, I_upper, c=bound_colour, lw=1)[0],
+                ax.fill_between(Ks, 
+                                I_lower, 
+                                I_upper,
+                                facecolor=bound_colour, alpha=0.3,
+                                zorder=-1)
+            ])
+
+            if set_upper:
+                ax.set_ylim(ylim)
+
 
 
         elif kind == "actual_I":
@@ -410,7 +449,10 @@ class VisualizationHandler(object):
             self._clear_items(self._plot_items[ax])
             
[…]

### 2018-09-11 sandbox_mixture_search.py

@@ -3,7 +3,7 @@ import numpy as np
 from gmmmml import (mixture_search as mixture, visualize, utils)
 import  scipy.optimize as op
 
-np.random.seed(43)
+np.random.seed(100)
 
 # Seed 1100 is a fun one: K = 103, D = 5, N = 4249
 
@@ -16,7 +16,7 @@ for i in range(10):
         print(i, j)
         j += 1
         try:
-            y, labels, target, kwds = utils.generate_data(K=40, D=10, center_box=(-10, 10))
+            y, labels, target, kwds = utils.generate_data(N=10000, K=50, D=10, center_box=(-50, 50))
 
         except:
           logging.exception("failed")
@@ -30,7 +30,7 @@ for i in range(10):
     visualization_handler = visualize.VisualizationHandler(
         y, target=target, figure_path="tmp/")
 
-    search_model = mixture.GaussianMixture()
+    search_model = mixture.GaussianMixture(threshold=1e-1)
     search_model.kmeans_search(y, K_max=kwds["centers"] + 25, 
         visualization_handler=visualization_handler)
 

### 2018-09-11 gmmmml/mixture_search.py
this is the demo

@@ -15,6 +15,7 @@ import scipy
 import scipy.misc
 import scipy.stats
 import scipy.optimize as op
+import tqdm
 import warnings
 import os
 from sklearn import cluster
@@ -335,6 +336,99 @@ class GaussianMixture(object):
         return (means, covs, weights)
 
 
+    def _search_log_jumper(self, y, **kwargs):
+
+        y = np.atleast_1d(y)
+
+        N, D = y.shape
+        kwds = {**self._em_kwds, **kwargs}
+
+        handler = kwds.get("visualization_handler", None)
+
+        # Initial guesses.
+        K_inits = np.logspace(0, np.log10(N), kwds.get("K_init", 10)).astype(int)
+
+        for K in K_inits[1:]:
+
+            # Assign everything to the closest thing.
+            means, covs, weights, responsibilities = self.initialize(y, K, **kwds)
+
+            # Run one E-M step.
+            try:                       
+                responsibilities, ll, message_length \
+                    = self.expectation(y, means, covs, weights, **kwds)
+
+            except ValueError:
+                logger.exception("Failed to calculate E-step")
+                # Break to adjust the delta_K
+                break
+
+            means, covs, weights = self.maximization(y, means, covs, weights,
+                                                     responsibilities, **kwds)
+
+        # Predict message lengths.
+        K_predict = np.arange(1, K)
+
+        K_skip = []
+
+        while True:
+
+            I, I_var, I_lower = self._predict_message_length(K_predict, N, D, **kwds)
+
+            # TODO: Check I_lower
+            #if np.any(I_lower > I):
+            #    keep = I > I_lower
+            #    K_predict, I, I_var, I_lower = K_predict[keep], I[keep], I_var[keep], I_lower[keep]
+
+            idx = np.argsort(I)
+
+            for K in K_predict[idx]:
+                if K in K_skip or K in self._state_K:
+                    continue
+
+                kbest = np.array(self._state_K)[np.argmin(self._state_I)]
+                print(f"Running {K}: best so far is K = {kbest}")
+
+                try:
+                    means, covs, weights, responsibilities = self.initialize(y, K, **kwds)
+
+                    # Run one E-M step.
+                    responsibilities, ll, message_length \
+                        = self.expectation(y, means, covs, weights, **kwds)
+
+                except ValueError:
+                    logger.exception("Failed to calculate E-step")
+                    # Break to adjust the delta_K
+                    K_skip.append(K)
+                    continue
+
+
+                prev_I = np.inf
[…]

### 2018-09-11 gmmmml/mml.py

@@ -327,18 +327,21 @@ def predict_information_of_sum_log_det_covs(K, D, data):
         raise NotImplementedError("cannot predict this theoretically")
 
     slogdetcovs = np.array([np.sum(np.log(dc)) for dc in data[1]])
-    x, y = _group_over(data[0], slogdetcovs, np.mean)
+    x, y = _group_over(data[0], slogdetcovs, np.max)
     _, yerr = _group_over(data[0], slogdetcovs, np.std)
 
+    yerr = np.ones_like(y)
+
     # TODO
-    y = y*x
+    #y = y*x
 
     yerr = np.clip(yerr, 1, np.inf)
+    yerr[~np.isfinite(yerr)] = 1
 
     var_y = np.var(y)
     var_y = var_y if (np.isfinite(var_y) and var_y > 0) else 1
 
-    kernel = var_y * kernels.ExpSquaredKernel(1)
+    kernel = var_y * kernels.Matern32Kernel(1)#ExpSquaredKernel(1)
     #       + var_y * kernels.LocalGaussianKernel(location=0, log_width=0)
     #+ np.var(y) * kernels.LinearKernel(log_gamma2=0, order=1)
 
@@ -370,19 +373,19 @@ def predict_information_of_sum_log_det_covs(K, D, data):
     pred, pred_var = gp.predict(y, K, return_var=True)
 
     # TODO
-    pred /= K
-    pred_var /= K**2
+    #pred /= K
+    #pred_var /= K**2
 
     I = information_of_sum_log_det_covs(pred, D)
     I_var = information_of_sum_log_det_covs(np.sqrt(pred_var), D)**2
 
     # Calculate the lower bound based on the data we have.
-    max_log_det_cov = np.log(np.max(np.hstack(data[1])))
-    min_log_det_cov = np.log(np.min(np.hstack(data[1])))
+    max_log_det_cov = np.max(np.log(np.hstack(data[1])))
+    min_log_det_cov = np.min(np.log(np.hstack(data[1])))
 
 
-    I_lower = information_of_sum_log_det_covs(K * max_log_det_cov, D)
-    I_upper = information_of_sum_log_det_covs(K * min_log_det_cov, D)
+    I_lower = information_of_sum_log_det_covs(K * min_log_det_cov, D)
+    I_upper = information_of_sum_log_det_covs(K * max_log_det_cov, D)
 
     # Calculate upper bound based on the two-point autocorrelation function
 
[…]

### 2018-09-11 gmmmml/visualize.py

@@ -441,8 +441,6 @@ class VisualizationHandler(object):
             scat.set_sizes(30 * np.ones(len(data)))
             scat.set_zorder(100)
 
-            print(f"{kind} {K} {I}")
-
         elif kind == "predict_I":
 
             ax = self._ax("I")

### 2018-09-11 sandbox_mixture_search.py

@@ -16,7 +16,7 @@ for i in range(10):
         print(i, j)
         j += 1
         try:
-            y, labels, target, kwds = utils.generate_data(N=10000, K=50, D=10, center_box=(-50, 50))
+            y, labels, target, kwds = utils.generate_data(N=10000, K=100, D=5, center_box=(-50, 50))
 
         except:
           logging.exception("failed")
@@ -27,12 +27,15 @@ for i in range(10):
         else:
           break
 
-    visualization_handler = visualize.VisualizationHandler(
-        y, target=target, figure_path="tmp/")
+    handler = visualize.VisualizationHandler(
+        y, target=target, figure_path="tmp/jumper")
 
-    search_model = mixture.GaussianMixture(threshold=1e-1)
+    search_model = mixture.GaussianMixture(threshold=1e-1, max_em_iterations=1)
+    search_model._search_log_jumper(y, visualization_handler=handler)
+
+    raise a
     search_model.kmeans_search(y, K_max=kwds["centers"] + 25, 
-        visualization_handler=visualization_handler)
+        visualization_handler=handler)
 
     """
     K = np.array(search_model._state_K)

### 2018-10-02 gmmmml/gmm.py
codes

@@ -362,7 +362,8 @@ class GaussianMixture(object):
         logger.info(f"Best mixture has K = {K_best} and I = {I_best:.0f}")
 
         state, R, ll, I = self._results[K_best]
-        meta = dict(strategy=strategy_class.__name__, t_search=time() - t_init)
+        meta = dict(strategy=search_strategy_class.__name__, 
+                    t_search=time() - t_init)
 
         # Set the state attribuets.
         self.means_, self.covs_, self.weights_ = state

### 2018-10-02 gmmmml/policies/initialisation.py

@@ -44,6 +44,7 @@ class DefaultInitialisationPolicy(BaseInitialisationPolicy):
 
             except ValueError:
                 logger.warn(f"Failed to initialise at K = {K}")
+                raise
                 break
 
 

### 2018-10-02 gmmmml/utils.py

@@ -1,7 +1,8 @@
 
 import numpy as np
-from sklearn import datasets
+from scipy.special import beta
 
+from sklearn import datasets
 from sklearn.neighbors import NearestNeighbors
 
 
@@ -282,10 +283,116 @@ def aggregate(x, y, function):
 
 
 
+def generate_data(N, D, K=None, dirichlet_concentration=1, isotropy=10, psi=0.05,
+                  random_seed=None, **kwargs):
+    r"""
+    Generate data from a mixture of multivariate Gaussian components with full
+    rank covariance matrices with non-zero off-diagonal terms.
+
+    :param N:
+        The number of samples to draw.
+
+    :param D:
+        The dimensionality of the data.
+
+    :param K: [optional]
+        The number of components to model. If ``None`` is given then a random
+        number will be drawn from a log-uniform (base 10) distribution between
+        :math:`K = 1` and :math:`K = \floor{N/2}`.
+
+    :param dirichlet_concentration: [optional]
+        A metric representing the uniformity of the relative weights of the
+        mixture. This is the concentration parameter for a symmetric Dirichlet
+        distribution to draw the weights of the components. The concentration
+        parameter must be greater than zero. A concentration parameter much
+        less than one indicates most of the data will be drawn from one
+        component. A concentration parameter much greater than one indicates
+        that the support will be distributed evenly between the components.
+
+    #  TODO
+    """
+
+    np.random.seed(random_seed)
+
+    N, D = (int(N), int(D))
+
+    if K is None:
+        K = int(np.round(10**np.random.uniform(0, np.log10(N/2.0))))
+
+    if K >= N/2.0:
+        raise ValueError(f"K must be less than N/2 ({2*K} > {N})")
+
+    if 0 >= dirichlet_concentration:
+        raise ValueError("dirichlet_concentration must be positive")
+
+    # Generate the weights.
+    min_weight = 2.0/N
+    max_weight = 1 - min_weight * (K - 1)
+
+    alphas = np.ones(K) * dirichlet_concentration
+    for iteration in range(kwargs.pop("dirichlet_draws", 100000)):
+        weights = np.random.dirichlet(alphas).flatten()
+
+        # Check the weights.
+        if np.min(weights) >= min_weight and np.max(weights) <= max_weight:
+            break
+
+    else:
[…]

### 2018-10-02 article/ms.tex

@@
 %\usepackage{boldsymbol}
+\usepackage{minted}
 \usepackage{color,colortbl} % for \todo command
@@
 % For revision history
-\IfFileExists{vc.tex}{\input{vc.tex}}{
+\newcommand{\vcpath}{vc.tex}
+
+\IfFileExists{\vcpath}{\input{\vcpath}}{
+	\newcommand{\giturl}{UNKNOWN}
+	\newcommand{\gitslug}{UNKNOWN}
 	\newcommand{\githash}{UNKNOWN}
-    \newcommand{\giturl}{UNKNOWN}}
+	\newcommand{\gitdate}{UNKNOWN}
+	\newcommand{\gitauthor}{UNKNOWN}
+}
 
@@
 \newcommand{\Cspaceno}{$\mathcal{C}$-space}
+
 % Common terms
@@
 
+\newcommand{\bashcommand}[1]{\noindent\texttt{#1}\\}
+
 % Affiliation(s)
@@
 \ead{aldeida.aleti@monash.edu}
-\author[moca,fit]{Andy Casey}
-\author[moca]{John Lattanzio}
+\author[moca,fit]{Andrew R. Casey}
+\author[moca]{John C. Lattanzio}
 \author[fit]{David L. Dowe}
@@
         \begin{keyword}
-            Search\sep Minimum Message Length\sep MML\sep mixture models\sep Galactic Archaeology
+            Search\sep Minimum Message Length\sep MML\sep gaussian mixture models\sep Galactic Archaeology
         \end{keyword}
@@
 Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
-of our Galaxy by identifying ancient star formation events. Most stars are 
+of our galaxy by identifying ancient star formation events. Most stars are 
 born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
@@
 Stars in clusters often do not remain gravitationally bound, and disperse 
-throughout the disk of our Galaxy within about $10^9\,{\rm yr}$. %citation needed.
+throughout the disk of our galaxy within about $10^9\,{\rm yr}$. %citation needed.
 As a consequence, where a star is now does not reflect its formation site, 
@@
 Here $\prior{\vec{\theta}}$ is the prior, $Q$ is the number of free parameters
-in the model, $\kappa(Q)$ is a function to approximate the lattice quantisation
+in the model,
+\begin{eqnarray}
+	Q = \frac{KD(D+3)}{2} + K - 1 \quad ,
+\end{eqnarray}
+\noindent{}$\kappa(Q)$ is a function to approximate the lattice quantisation
 constant for $Q$ free parameters\cite{ConwaySloane1984}, and $|\fisher{\vec{\theta}}|$
@@
 
-In the following subsections we provide expressions for the message length of
-each component listed above, and our prior beliefs, before describing our
-full objective function.
-
-\subsubsection{The number of mixture components (or groups) $K$}
-
-We assume a prior on the number of components $K$ of $\prior{K} \propto 2^{-K}$
-\cite[][p279, sec. 6.8.2]{Wallace05} such that the length of the 
-optimal lossless message to encode $K$ is 
-\begin{equation}
-    I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant} \quad .
-\end{equation}
-This  expresses our prior belief that there ought to be fewer components.
-
-\subsubsection{The relative weight $w_k$ of each component}
-
-We assume a uniform prior on the individual weights $\weight_{k}$. The weights
-$\weights$ are drawn from a multinomial distribution, such that only $K - 1$
-weights need to be encoded because $\sum_{k=1}^{K}\weight_k = 1$. Following
-from Equation \ref{eq:taylor}, here we describe how our estimate of the message 
-length of the multinomial parameter $\vec{\weight}$ quantifies the trade-off 
-between stating the parameters $\vec{\weights}$ imprecisely, so as to keep the 
-first part of the message short and place higher prior probability in the 
-relevant uncertainty region, and the goodness-of-fit to the data, which is 
-typically better if we state the estimates of the parameters more precisely. 
-By balancing this trade-off, the length of this component is equal (within a 
-small constant) to the negative log of the product of the prior probabilities 
-of each group ($\prior{\vec{\weight}}$) divided by the square root of the 
-expected Fisher information ($\sqrt{\fisher{\vec{\weight_k}}}$), where the 
-Fisher information matrix contains the \emph{expected} second-order partial 
-derivatives of the log-likelihood function. This describes how sharply the 
-likelihood function peaks, and as a result it quantifies how precisely the 
-parameters should be stated. In the case of uniform prior probabilities,
-\begin{equation}
-    \prior{\vec{\weight}} = (K - 1)! \quad ,
-\end{equation}
-\noindent{}the Fisher information is
-\begin{equation}
-\fisher{\vec{w}} = \frac{N^{K - 1}}{\prod_{k=1}^{K} w_k} \quad ,
-\end{equation}
-\noindent{}which gives the message length of this part to be
-\begin{eqnarray}
-I(\vec{w}) &=& -\log\left(\frac{\prior{\vec{w}}}{\sqrt{{\kappa_K}^K \fisher{\vec{w}}}}\right)  \nonumber \\
-I(\vec{w}) &=& -\log{(K - 1)!} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k} \nonumber \\
-I(\vec{w}) &=& -\log{\Gamma(K)} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k}
-\end{eqnarray}
-\noindent{}where $\kappa_K$ is a lattice constant that varies between $1/12$
-and $1/(2\pi e)$ \cite{BoultonWallace1969}.
-
-\subsubsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
-               $K$ Gaussian components}
-
-In order to properly encode the means $\vecmean$ and covariance matrices
-$\veccov$ for all $K$ mixtures, we must encode both our prior belief on 
-those parameters and the determinant of the expected Fisher information 
-matrix. For the $k$-th mixture the joint message length is
-\begin{eqnarray}
-  I(\vecmean_k,\veccov_k) &=& -\log{\left(\frac{\prior{{\vecmean_k,\veccov_k}}}{\sqrt{\detfisher{{\vecmean_k,\veccov_k}}}}\right)} \nonumber \\ 
-  I(\vecmean_k,\veccov_k) &=& -\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\log{\detfisher{{\vecmean_k,\veccov_k}}}
-\end{eqnarray}
-
-\noindent{}and for all mixtures:
-\begin{equation}
-  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \quad .
-  \label{eq:I_component_params}
-\end{equation}
-
-We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
-that 
-    $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
-However,  the MML principle requires proper priors. In practice we make 
-this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
-we do so without actually specifying the value of the bounds because they add 
-only a constant value to the total message length\todo{\cite{someone}}.
-We adopt a conjugate inverted Wishart prior for the covariance matrix of an
-individual mixture \cite[Section 5.2.3]{Schafer_1997}:
-
-\begin{equation}
-  \prior{{\vecmean_k, \veccov_k}} \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
-  \label{eq:covariance-prior}
-\end{equation}
-
-% Casey: Reword following paragraph because it is clumsy.
-Following Eq. \ref{eq:I_component_params}, we require the logarithm of the
-determinant of the expected Fisher information matrix for $\vecmean_k$ and
-$\veccov_k$ in order to encode the full message. In practice the determinant
-of the Fisher information is non-trivial to evaluate, and as such we
-approximate the determinant of the Fisher information
- $\detfisher{{\vecmean_k, \veccov_k}}$ 
-as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
-\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
-we take the second derivative of the log-likelihood function 
-$\likelihood(\data|\vectheta)$, which yields:
-\begin{equation}
-  \detfisher{\vecmean_k} = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
-\end{equation}
-We make use of an analytical expression for $\detfisher{\veccov_k}$
-\cite{Dwyer_1967,Magnus_1988}:
-\begin{equation}
-  \detfisher{\veccov_k} = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad .
-\end{equation}
-Combining these expressions gives our approximation for the determinant of
-the expected Fisher information $\detfisher{{\vecmean_k,\veccov_k}}$:
-\begin{eqnarray}
-  \detfisher{{\vecmean_k,\veccov_k}} & \approx & \detfisher{\vecmean_k} \cdot \detfisher{\veccov_k} \nonumber \\
-  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
-  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
-  \label{eq:mixture-component-det-fisher}
-\end{eqnarray}
-
-Substituting Eqs. \ref{eq:covariance-prior} and \ref{eq:mixture-component-det-fisher} 
-into Eq. \ref{eq:I_component_params} gives the message length to encode the component
-parameters $\vecmean$ and $\veccov$ for all $K$ mixtures:
-
-\begin{eqnarray}
-  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \nonumber \\
-  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
-    + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
-%  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
-    %-\frac{KD}{2}\log{2} \nonumber \\
-%  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
-  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
-\end{eqnarray}
-
-\subsubsection{The data given the model}
-If we assume that the data have homoskedastic noise properties, then the 
-precision of each measurement $\mathcal{E}$ relates the probability of a
-datum $p(\datum_n)$ can be related to the probability density of
-the datum, given the model 
-$p(\datum_n) = \mathcal{E}^{D}p(y_i|\mathcal{M})$.
-In this work we adopt $\mathcal{E} = 0.01$, but we note that the value of
-$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
-a datum given the model is,
-\begin{equation}
-  I(\data_n) = -\log{p(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
-\end{equation}
-
-\noindent{}and for the entire data:
-\begin{eqnarray}
-  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
-\end{eqnarray}
-
-The final part of the message is to encode the lattice quantisation constant,
-which arises from the approximation to the (so-called) strict MML, where
-parameters are quantised into intervals in high dimensional space in order to
-be losslessly encoded.  We use an approximation of the lattice constant
-\citep[see Sections 5.1.12 and 3.3.4 of ]
-%[]
-{Wallace_2005} such that,
-\begin{equation}
-  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
-\end{equation}
-
-\noindent{}where $Q$ is the total number of free parameters in the model. Each
-Gaussian component in the mixture requires $D$ parameters to specify $\vecmean$,
-and a covariance matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ 
-parameters. Because only $K - 1$ weights need to be encoded, the total number of free
-parameters is
-\begin{equation}
-    Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1 \quad .
-    \label{eq:number-of-parameters}
-\end{equation}
-
-\subsection{Objective function}
-
-Combining the expressions for the message length (or information content) of
-each component, we arrive at our objective function $I$ to minimise across 
-all model parameters (including the number of mixtures, $K$):
-
-\begin{equation}
-\todo{I = 0}
-\end{equation}
-
-%\begin{eqnarray}
-%I_K & = & K\log{2} % I_k
-%    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
-%    +\mathcal{L}(\data|\vectheta) \nonumber \\
-%    & -& DN\log{y_{err}} % Likelihood
-%   +  \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right]\nonumber \\% I_t
-%    &-& \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
-%%\end{eqnarray}
-
-%\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
-%$\mathcal{L}$ is the log-likelihood of a multivariate Gaussian distribution.
-
-While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
-where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
-\vecmean, \veccov\}$, and we can meaningfully compare models with different 
-numbers of components or parameter values, we require a search strategy in 
-order to efficiently find the global solution: the minimum message length 
-over a class of finite mixture models (where $N > K$). Before we introduce and 
-compare search strategies, let us first make some comments and quantitative 
-descriptions on how the message length $I$ will vary with the model parameters. 
-Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
-sum of three smaller functions, and we will derive limits and predictions for 
-those quantites. In an ideal scenario, those limits and predictions would be 
-able to provide us with an estimate of the message length for a Gaussian 
-mixture before computing the mixture (i.e., before performing 
-expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
-as
-\begin{eqnarray}
-I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
-                          + I_{2}(K, N, \veccov)
-                          + I_{3}(\vec{\psi}) \quad ,
-\end{eqnarray}
-\noindent{}where:
-\begin{eqnarray}
-    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
-          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
-    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad \textrm{and},\\
-    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\mathcal{E}} \quad .
-\end{eqnarray}
-The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
-message that we introduced earlier to describe message length. Rather we are
-simply re-arranging the objective function because, as we will show, there are
-useful theoretical bounds and predictions that we can make for each part of
-this re-arranged expression.
-
-In Eq. \ref{eq:I1} it can be seen that for any finite mixture, $I_1$ can be
-almost directly computed just from knowing the number of mixture components
-$K$ and the properties of the data: the number of data points $N$ and its
-dimensionality $D$. The only ambiguous term is $\sum_{k=1}^{K}\log\weight_k$,
-which will depend on the data. However, for any finite mixture of $K$
-components we can consider the two extremes of $\sum_{k=1}^{K}\log\weight_k$:
-
-\begin{enumerate}
-    \item The weights are distributed uniformly such that $w_k = 1/K$ and
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
-          \end{equation}
-    \item The weights are distributed such that $K - 1$ components each 
-          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
-          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
-          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
-                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
-          \end{equation}
-          \noindent{}and by making use of the identity 
-            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
-          \end{equation}
-\end{enumerate}
-
-\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
-derived the bounds
-\begin{eqnarray}
-\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
-\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
-\end{eqnarray}
-\noindent{}which gives us analytic expressions for the upper and lower value 
-of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
-bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.
-
-Developing similar theoretical bounds for $I_2$ is not straightforward, because
-the determinants of the covariance matrices of Gaussian components will not
--- to our knowledge -- strictly follow some expected distribution, in the way 
-that the weights $\weights$ follow a multinomial distribution. Similarly, the
-\emph{scale} of $|\veccov_k|$ will depend entirely on the distribution of the
-data. In that sense, we cannot a priori state with any certainty about what 
-values $|\veccov_k|$ might take (or $\log{|\veccov_k|}$) without knowing
-something about the data first. Even demanding that the covariance matrices
-$\veccov_k$ are positive semi-definite only informs us that $\log{|\veccov_k|}$
-is finite. However, there are quantitative predictions we can make on the 
-bounds of $\sum_{k=1}^{K}\log{|\veccov_k|}$ if something is known about the
-data. Specifically if we consider the determinant of the sample covariance 
-matrix, which we denote $|\cov_{u}|$ (where the $u$ subscript denotes `upper',
-for the upper limit on the covariance matrix), then we can state with confidence that
-the sample covariance matrix describes the maximum amount of variance that 
-could be captured by the components in the mixture. For example, a limiting
-case of $\sum_{k=1}^{K}\log{|\veccov_k|}$ can be described when the determinant
-of the sample covariance matrix must be equally (uniformly) captured by $K$ 
-components in a mixture. In other words, the maximum variance in the data 
-must be equally spread between the components in the mixture. Uniformly 
-describing the variance in the data is the limiting case because any 
-non-uniformity would lead to smaller $|\veccov|$ for one mixture (and larger 
-for another), and that asymmetry would lead to a higher $I_2$ because 
-$\log(x - \epsilon) + \log(x + \epsilon) < 2\log{x}$, and because of the 
-$-(1 + \frac{D}{2})$ term that enters multiplicatively in the calculation of 
-$I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
-the sample covariance matrix $|\cov_{u}|$:
-
-\begin{eqnarray}
-    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{\frac{|\cov_u|}{K}} \nonumber \\
-    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\left(K\log{|\cov_u|} - K\log{K}\right) 
-\end{eqnarray}
-
-The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
-possible variance that could be captured by any single component. In theory
-this approaches zero, but in practice it will be set by the minimum pairwise
-distance in $D$ dimensions. Specifically, if we consider two data points to be
-captured by a component in the mixture, an those two data points are the
-closest two data points (in high dimensional space) of any data points, then
-the variance captured by that Gaussian component will represent the \emph{minimum}
-possible variance that can be captured by any single component.  
-In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
-in $D$ dimensions, average it over the dimensions, and approximate the minimum
-determinant of the covariance matrix $|\cov_l|$ as
-\begin{equation}
-    |\cov_{l}| \approx \prod_{d=1}^{D}\left(\frac{d_{min}}{2}\right)^2 \quad ,
-\end{equation}
-\noindent{}which approximately represents the variance between two data points,
-but ignores correlations and assumes only a mean distance in $D$ dimensions
-(for comuptational cost). With these approximations,
-\begin{eqnarray}
-    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\left(\frac{d_{min}}{2}\right)^2}  \nonumber \\
-    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
-\end{eqnarray}
-\noindent{}and the upper limit on $I_2$ is:
-\begin{eqnarray}
-    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
-    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
-\end{eqnarray}
-
-Here our bounds on $I_2$ are only dependent on $K$, $D$, the minimum euclidean
-pair-wise distance $d_{min}$, and the determinant of the sample covariance
-matrix $|\cov_u|$. In practice it is the lower bound on $I_{2}$ that is 
-arguably more useful for search and optimisation, but one could conceive
-situations where there is great utility in also having upper bounds on $I_1$
-and $I_2$.
-
-There is no finite upper bound on $I_{3}$ because the negative log-likelihood
-$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
-data. A useful lower limit, however, does exist. The negative log-likelihood
-in Eq. \ref{eq:log-likelihood} can be expanded as,
-\begin{equation}
-    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
-\end{equation}
-\noindent{}and we can define
-\begin{equation}
-\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
-\end{equation}
-For the sake of placing a lower bound on $I_{3}$, let us assume that each data
-point will be `mostly' explained by one of the components in the 
-mixture\footnote{We model the data with partial assignment, but here we merely 
-seek to place a lower bound on the negative log-likelihood.}, but we do not 
-necessarily know which component, or what the parameters of that component are. 
-In that situation, we can state that the component that \emph{does} explain
-the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
-whereas the other components that do not explain the data will have 
-$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
-this simplification we can write an approximation for the negative log-likelihood
-that does not require knowledge about which data points are associated to what
-component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
-will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
-approximate lower bound is:
-\begin{eqnarray}
-    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
-    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad .
-\nonumber
-    \label{eq:logl-lower-bound}
-\end{eqnarray}
-\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
-for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
-to the smallest realistic values, where those mixtures encapsulate only two
-data points and $w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N$ such that
-%\begin{eqnarray}
-%	\sum_{k=1}^{K}\weight_{k}\log\weight_k = \left[(K-1)\left(\frac{2}{N}\log{\frac{2}{N}}\right) + \left(1-\frac{2(K-1)}{N}\right)\log{\left(1-\frac{2(K-1)}{N}\right)}\right] \nonumber
-%\end{eqnarray}`
-%\noindent{}and
-\begin{eqnarray}
-	-N\sum_{k=1}^{K}\weight_{k}\log\weight_k \geq 2(1-K)\log{2} + N\log{N} - (N-2K+2)\log{\left(N - 2K + 2\right)}
-\end{eqnarray}
-%  log(a-b) = log(a) + log(1 - b/a)
-% log(1 - b/a) = log(a-b) - log(a)
-% log(1 - b/a) = log(N - 2K + 2) - log(N)
-
-% 2(1-K)log(2) - 2(1-K)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
-% 2(1-K)log(2) - 2log(N) + 2Klog(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
-% 2(1-K)log(2) + (2K - 2)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - 2 - N)log(N)
-% 2(1-K)log(2) + Nlog(N) + (2K - N - 2)log(N - 2K + 2)
-
-The remaining term in Eq. \ref{eq:logl-lower-bound}, $\frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log|\vec{\cov}_k|$, requires some  consideration.
-
-\andy{here}
-
-\section{Search Methods}
-
-\subsection{The MML$_{K+D}$ Method}
-Say we wanted to calculate whether another mixture was warranted. If another
-mixture were preferred then we would want:
-
-\begin{eqnarray}
-  \Delta{}I_{K+1} - I_{K} < 0
-\end{eqnarray}
-
-The expression is given as:
-\begin{eqnarray*}
-\Delta{I_{K+1} - I_K} & = & (K + 1)\log{2} - K\log{2} \\ % I_k^(new) - %I_k^(old)
-  &&+ \frac{(K)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K+1}\log{w_k}^{(new)} - \log{|\Gamma(K+1)|} \\ % I_w^(new)
-  &&- \frac{(K - 1)}{2}\log{N} + \frac{1}{2}\sum_{k=1}^{K}\log{w_k} + \log{|\Gamma(K)|}\\ % I_w^(old)
-  &&+ \mathcal{L}^{(new)} - DN\log{y_{err}} \\ % Likelihood (new)\\
-  &&- \mathcal{L}^{(old)} + DN\log{y_{err}} \\ % Likelihood (old) \\
-  &&+ \frac{1}{2}\sum_{k=1}^{K+1(new)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(new)
-  &&- \frac{1}{2}\sum_{k=1}^{K(old)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(old)
-  &&- \frac{Q^{(new)}}{2}\log(2\pi) + \frac{\log{Q^{(new)}\pi}}{2} \\ % lattice, minus the lattice of part 2
-  &&+ \frac{Q^{(old)}}{2}\log(2\pi) - \frac{\log{Q^{(old)}\pi}}{2} % lattice, minus the lattice of part 2
-\end{eqnarray*}
-
-\noindent{}By making use of $\log{\Gamma(K)} - \log{\Gamma(K + 1)} = -\log{K}$ and re-arranging the expression:
-
-\begin{eqnarray}
-\Delta{}I_{K+1} - I_K &=& \log{2} % \Delta{}I_k
-    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
-& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
-& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber\\
-&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
-& +& \frac{\log(2\pi)}{2}(Q^{(old)} - Q^{(new)}) + \frac{\pi}{2}\left(\log{Q^{(new)}} - \log{Q^{(old)}}\right)
-\label{eq:13}
-\end{eqnarray}
-
-Expanding the $Q$ terms:
-
-\begin{eqnarray}
-Q^{(old)} - Q^{(new)} &=& \frac{1}{2}DK(D + 3) + K - 1 - \frac{1}{2}D(K + 1)(D + 3) + (K + 1) - 1 \nonumber \\
-Q^{(old)} - Q^{(new)} &=& -\frac{1}{2}D(D+3) + 2K  - 1
-\label{eq:14}
-\end{eqnarray}
-
-\noindent{}And making use of the following logarithmic identities,
-\begin{eqnarray}
-  \log{Q^{(new)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3) + K\right)} \nonumber \\
-                  &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} + \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \\
-  \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}DK(D + 3) + K - 1\right)} \nonumber \\
-                  &=& \log{\left(\frac{1}{2}DK(D + 3)\right)} + \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}
-\end{eqnarray}
-
-\noindent{}gives us,
-
-\begin{eqnarray}
-  \log{Q^{(new)}} &- \log{Q^{(old)}} = \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
-                                    &+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}.
-\end{eqnarray}
-The second row of terms can be ignored because they are very small (typically less than 1 bit).
-This is because as $K \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 1$, thus $\log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \rightarrow \log{2}$.
-Similarly as $D \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 0$.
-
-As $K \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 1$ and as $D \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 0$ and thus $\log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \approx 0$.
-
-\noindent{}Ignoring these minor terms:
-
-\begin{eqnarray}
-  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
-  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{(K + 1)} - \log{K}
-  \label{eq:19}
-\end{eqnarray}
-
-\noindent{}Substituting Eqs. \ref{eq:19} and \ref{eq:14} into \ref{eq:13} yields:
-
-\begin{eqnarray}
-\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
-    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
-& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
-& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber \\
-&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
-& +& \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
-\end{eqnarray}
+In Appendix \ref{app:mml_derivation} we define our priors and describe the 
+individual contributions to the mixture message length. For the sake of brevity we
+denote the full set of model parameters as $\vec{\psi} = \{K, \vec{\weight}, \vecmean, \veccov\}$.
+The total message length $I(\vec{\data}|\vec{\psi}, N, D)$ to model some $N \times D$ dimensional data $\vec{\data}$ with a mixture of multivariate Gaussian components is given by:
+\begin{eqnarray}
+	I(\vec{\data}|\vec{\psi}, N, D) &=& -\log\mathcal{L}\left(\vec{\data}|\vectheta\right) - \frac{(D + 2)}{2}\sum_{k=1}^{K}\log{|{\cov}_k|} \nonumber \\
+	&& \cdots\ + \left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum_{k=1}^{K}\log{\weight_k} + K\left(1-\frac{D}{2}\right)\log{2} \nonumber \\ && \cdots\ + \log\Gamma{(K)} + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N} \,\,\,\, .
+	\label{eq:objective_function}
+\end{eqnarray}
+
+\subsection{Expectation-Maximization}
+
+\todo{Derivations and updates to the mixture parameters through E-M}
+
+\subsection{Theoretical lower bounds on the message length of future mixtures}
+
+Knowing how some of the model parameters are distributed allows us to derive theoretical
+bounds on the message length for individual components. These bounds allow us to describe
+the lower and upper bounds on \emph{some components of} the message length of a Gaussian mixture
+model, before estimating the model parameters (or running expectation-maximisation). For the
+purposes of this work we will describe such mixtures as \emph{future mixtures} as they have
+not been trialled yet.
+
+In Appendix \ref{app:mml_limits} we describe four useful bounds. The first two are lower and upper bounds on the   message length to encode the sum of the log of the weights $\sum_{k=1}^{K}\weight_k$
+for any future mixture.
+The third is an approximation to the lower bound of the negative log-likelihood $-\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right)$, and the fourth is an approximation on the message length required to encode
+the sum of the log of the determinant of the covariance matrices $\sum_{k=1}^{K}\log|\cov_k|$. These are the  most important components in the objective function described
+in Equation \ref{eq:objective_function}, because their values are not known until some set of model parameters $\vec{\psi}$ is trialled -- and the optimal set of model parameters is not known \emph{a priori} -- and all other terms in Equation \ref{eq:objective_function} are deterministic, depending only on $K$, $N$, and $D$.
+While we do provide both a lower and upper bound on the message length to encode the weights, no such upper bound exists on the
+negative log-likelihood because this could be arbitrarily high for an increasingly poor estimate of the mixture
+parameters.
+Some upper limit likely exists for $\sum\log|\cov_k|$, but that limit would depend on how the data are separated.
+Nevertheless, it is the lower bound on these quantities that is most useful for optimisation.
+
+Empirically we found that when the dimensionality of the data $D$ is small, these theoretical limits can efficiently describe a lower bound to the message length for any given Gaussian mixture of $K$ components.
+This allows, in some situations, to efficiently place a useful upper limit on the maximum number of components $K$, $K_{max}$ (useful in that $K_{max} << N$).
+
+\subsection{Predicting the message length of future mixtures}
+
+The objective function in Equation \ref{eq:objective_function} remains valid across a class of 
+Gaussian mixture models where $N > K$. However, we still require a strategy to minimise this
+objective function over all possible mixture parameters $\vec{\psi}$. Here we describe methods to predict
+the \emph{minimum} message length of future (untrialled) mixtures, which we will then use in conjunction 
+with search strategies (Section \ref{sec:search}).
+
+% TODO: Predicting the sum of the log of the weights.
+
+% TODO: Predicting the negative log-likelihood.
+
+% TODO: Predicting the sum of the log of the determinant of the covariance matrices.
+
+% Combining it together.
+
+\section{Search Strategies} \label{sec:search}
+
+We introduce and evaluate two search strategies that make use of the objective function in Equation \ref{eq:objective_function}.
+These search strategies are compared against an existing state-of-the-art method, which we describe in Section \ref{sec:experimental-design}.
+
+\subsection{The Message-Breaking Method} \label{sec:mbm}
+The first strategy, \emph{The Message-Breaking Method}, does not make any predictions on
+the message length of future mixtures. Instead it starts from 
+a $K=1$ mixture and iteratively splits the component with the largest message length until 
+the total message length no longer improves (Algorithm~\ref{alg:MBM}). We evaluate the message length of each component
+by re-ordering Equation \ref{eq:objective_function} and weighting the negative log-likelihood
+$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ with entries
+\begin{eqnarray}
+	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
+\end{eqnarray}
+\noindent{}such that the message length of the $k$th component is
+\begin{eqnarray}
+	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
+		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
+		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
+\end{eqnarray}
+\noindent{}and the total message length remains the sum of the message lengths of the components
+\begin{eqnarray}
+	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
+\end{eqnarray}
+
+The component-wise calculation of the message length is performed as part of the 
+expectation step of the expectation-maximisation algorithm, adding negligible computational overhead.
+
+When splitting the component with the largest message length, we split this component into two child components. Initially we 
+set the means of the child mixtures as positive and negative eigenvalues along the principal
+eigenvector of the component covariance matrix. A matrix of child responsibilities is then
+calculated based on the Euclidean distance between each datum and the child means (i.e., hard
+association), which we use to estimate the covariance matrices and weights of the child
+components. After the parameters of the child components are initialised, we run \emph{partial}
+expectation-maximization to iteratively update the parameter estimates of the child components. 
+By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
+all other components fixed, and only updating the estimates of the parameters of the child
+components \citep{kasarapu2015minimum}.
+After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
+
+One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
+One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$.
+As a result, the algorithm would converge to a local minima, not the global one.
+However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.
 
 \begin{algorithm}[H]
-\caption{The MML$_{K+D}$ Method.}
-\label{alg:MMLKD}
+\caption{The Message-Breaking Method.}
+\label{alg:MBM}
+{\bf Input:} $\vec{\data}$ \Comment{The $N \times D$ dimensional data.}\\
+{\bf Output:} $\vec{\psi}$ \Comment{The optimal mixture parameters.} 
 \begin{algorithmic}[1]
-\State {\bf \sc input:} $K$ \Comment{Number of initial components.}
-\State current $\leftarrow$ $K$ component mixture
+\State current $\leftarrow$ one component mixture
 \While{TRUE}
+\State $m$ $\leftarrow$ \Call{ComponentMessageLength}{current}
+\State $i$ $\leftarrow$ \Call{ArgMax}{$m$} 
+\State perturbation $\leftarrow$ \Call{Split}{current, components[i]} \label{line:mbm-split}
+\State $\Delta{}I \leftarrow$ \Call{MessageLength}{perturbation} - \Call{MessageLength}{current} 
+\If{$\Delta I < 0$} \Comment{Check for improvement.}
+\State current $\leftarrow$ perturbation
+\State {\bf \sc Continue}
+\EndIf
+\State {\bf \sc Break} \label{line:end}
 \EndWhile
-\State {\bf \sc output:} 
 \end{algorithmic}
@@
 
-\subsection{Method of Moments}
-% \alda{This is a copy paste from David's email with minor changes}
-%
-% Re the case of two (or more) classes with the same mean:
-% In 2D, a simple case in point is two classes with the same mean
-% such that $\sigma_{1, x} = \sigma_{2, y} >> \sigma_{2, x} = \sigma_{1, y}$.
-% Andy said that this and higher dimensional stuff can be picked up
-% by dropping to 1 dimension along the eigenvector of largest eigenvalue.
-%
-%
-% I wonder also about the merits of dropping to 2 dimensions (if we have at least 2 attributes), being the eigenvectors of the two largest eigenvalues. 
-%
-% One could possibly check whether uniform radially directly from the original 2D projection without having to project it to a circle (i.e., it might be pretty easy to work out what the angle of each point would be post-transformation without having to do the transformation).
-The 2D projection is 1D in the sense that we only care about the angle around the circle.
-And then some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform in radial distribution.
-If not uniform, then perhaps grounds for splitting.
-%
-% Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
-% as above but rather projecting on eigenvector of largest eigenvalue), we
-% discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution}.
-% So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
-%
-
-We informally outline here the method of moments form of search heuristic and then look at how it might be applied.
-
-Let's suppose we have a 1D component.
-We'll shift it (without loss of generality, w.l.o.g.) so that its mean is 0 and we'll shrink or expand it (again, w.l.o.g.) so that its s.d.
-($\sigma$) is 1 (and variance v is 1).
-We consider now splitting it into K components with mean 0,
-mixing weights (or proportions) $w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
-Let's set $K = 2$, so $w_2 = 1 - w_1$.
-
-Noting that the 2nd, 4th, 6th and 8th moments of a true Gaussian distribution are 1, 3, 15, 105,
-suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6$ and $105 \alpha_8, ... $.
-We get some equations - if we have the correct number of equations,
-they're simultaneous equations.
-
-If we get too many equations (by using too many moments), then it
-becomes something of a fitting or regression problem to choose
-$w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.
-
-Our simultaneous equations are:
-
-\begin{align*}
-w_1 v_1       + (1 - w_1) v_2       &=    1\\
-w_1 (v_1)^2  + (1 - w_1) (v_2)^2  &=    3 \alpha_4\\
-w_1 (v_1)^3  + (1 - w_1) (v_2)^3  &=   15 \alpha_6\\
-w_1 (v_1)^4  + (1 - w_1) (v_2)^4  &=  105 \alpha_8
-\end{align*}
-
-% These might be messy to solve on the fly  but one could pre-process by solving many of these beforehand (actually, approximately solving these beforehand, as it's only
-% a heuristic, anyway) at the very start of the program while loading in the data.
-
-So, for $1000 \times 1000 = 10^6$ different values of $\alpha_4$ and $\alpha_6$
-or for $100 \times 100 \times 100 = 10^6$ different values of $\alpha_4, \alpha_6$ and $\alpha_8$
-one could pre-compute {\em approximate} 1,000,000ish `solutions' to the above simultaneous equations.
-
-Then, when running the program with real actual values of $\alpha_4$ and $\alpha_6$ (and $\alpha_8$),
-one could just grab something from nearby in this pre-computed look-up table
-as putative values for $w_1, v_1, v_2$ (and etc., if need be).
-
-% As an example of this algorithm, consider the mixture $0.6 N(-3, 2^2) + 0.4 N(7, 3^2)$.
-% The best fitting single Normal distribution to this would be $N(1, \sqrt{30}^2)$.
-%
-As an
-% another
-example of this algorithm, consider a first moment of $0$, a $2^{\rm nd}$ moment of $6$ and a $4^{\rm th}$ moment of $3 \times 37.2 = 111.6$.
-Solving the simultaneous equations gives us $0.6 N(0, 2^2) +  0.4 N(0, 3^2)$.
-
-\subsection{The k-means++ algorithm}
-
-The k-means++ algorithm extends the k-means method with an efficient way of choosing centres.
-Let $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected.
-The k-means++ has the following steps:
-
-\begin{enumerate}
-    \item Selects a centre $c_1$ uniformly at random from the $N$ data points.
-    \item Selects a new centre $c_i$ from the data points with probability $\frac{D(x)^2}{\sum_{x\in X}}D(x)^2$
-    \item Repeat Step 2. until k centres have been selected.
-    \item Proceed as with the standard k-means algorithm.
-\end{enumerate}
-
-k-means++ is $O(\log k)$-Competitive.
+\subsection{Component-Jumping Method}
+
+The \emph{Component-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.
+
+\begin{algorithm}[H]
+
+\algblock[TryCatchFinally]{try}{endtry}
+\algcblock[TryCatchFinally]{TryCatchFinally}{finally}{endtry}
+\algcblockdefx[TryCatchFinally]{TryCatchFinally}{catch}{endtry}
+	[1]{\textbf{catch} #1}
+	{\textbf{end try}}
+
+\caption{The Component-Jumping Method.}
+\label{alg:component-jumping-method}
+{\bf Input:} $\vec\data$, $K_{i}$, $I_\epsilon$ \\
+{\bf Output:} $\vec\psi$ \Comment{The optimal mixture parameters.}
+\begin{algorithmic}[1]
+\State s $\leftarrow (K_{i})^{-1}\log_{10}{N}$ \Comment{Initial step size in $K$.}
+\State v $\leftarrow [0, 1, ..., K_{i}]$ 
+\State $K_{inits} \leftarrow$ \Call{Power}{10, vs} \Comment{Set initial $K$ values to trial.}
+\State list $K_{t}$ \Comment{Create a list to save trialled $K$ values.}
+\For{$k\in K_{inits}$} 
+  \State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
+  \try 
+	  \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
+	  \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
+  \catch
+  	  \State {\bf \sc Break} \Comment{Do not trial any higher $K$ values.}
+  \finally 
+      \State mixture[k] $\leftarrow$ mix 
+      \State state[k] $\leftarrow$ \Call{MessageLength}{mixture[k]}
+  \endtry
+\EndFor
+\State converged $\leftarrow$ FALSE
+\While{not converged}
+	\State $K_{predict} \leftarrow [1, \cdots, 2\Call{Max}{K_{t}}]$ \Comment{Predict future message lengths.}
+	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$}
+	% argsort.
+	\State $K_{next} \leftarrow$ \Call{ArgSort}{$I_{predict}$} \Comment{Sort $K$ by predicted message length.}
+	\For{$k\in K_{next}$}
+		% Check if the mixture at K was good?
+		\If{$k\in K_{t}$}
+			\State converged $\leftarrow$ TRUE
+			\State {\bf \sc Break}
+		\Else
+			\State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
+	   	  \try 
+		    \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
+	   	    \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
+		  \catch
+  	        \State {\bf \sc Continue} \Comment{Move on to next $K$ trial.}
+		  \endtry 
+		\EndIf
+	\EndFor
+\EndWhile
+\end{algorithmic}
+\end{algorithm}
 
@@
 The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.
+\andy{Huh?}
 
@@
 
-We employ two criteria for comparing the proposed search methods to the benchmark: the message length and the Kullback Leibler divergence~\cite{kullback1951information}.
-
-\paragraph{Message Length}
-
-\paragraph{Kullback Leibler divergence}
-
 \section{Results}
@@
 
+\section{Appendices}
+\subsection{The MML objective function}
+\label{app:mml_derivation}
+In the following subsections we provide expressions for the message length of
+each component listed above, and our prior beliefs, before describing our
+full objective function.
+
+\subsubsection{The number of mixture components (or groups) $K$}
+
+We assume a prior on the number of components $K$ of $\prior{K} \propto 2^{-K}$
+\cite[][p279, sec. 6.8.2]{Wallace05} such that the length of the 
+optimal lossless message to encode $K$ is 
+\begin{equation}
+    I(K) = -\log{\prior{K}} = K\log{2} + \textrm{constant} \quad .
+\end{equation}
+This  expresses our prior belief that there ought to be fewer components.
+
+\subsubsection{The relative weight $w_k$ of each component}
+
+We assume a uniform prior on the individual weights $\weight_{k}$. The weights
+$\weights$ are drawn from a multinomial distribution, such that only $K - 1$
+weights need to be encoded because $\sum_{k=1}^{K}\weight_k = 1$. Following
+from Equation \ref{eq:taylor}, here we describe how our estimate of the message 
+length of the multinomial parameter $\vec{\weight}$ quantifies the trade-off 
+between stating the parameters $\vec{\weights}$ imprecisely, so as to keep the 
+first part of the message short and place higher prior probability in the 
+relevant uncertainty region, and the goodness-of-fit to the data, which is 
+typically better if we state the estimates of the parameters more precisely. 
+By balancing this trade-off, the length of this component is equal (within a 
+small constant) to the negative log of the product of the prior probabilities 
+of each group ($\prior{\vec{\weight}}$) divided by the square root of the 
+expected Fisher information ($\sqrt{\fisher{\vec{\weight_k}}}$), where the 
+Fisher information matrix contains the \emph{expected} second-order partial 
+derivatives of the log-likelihood function. This describes how sharply the 
+likelihood function peaks, and as a result it quantifies how precisely the 
+parameters should be stated. In the case of uniform prior probabilities,
+\begin{equation}
+    \prior{\vec{\weight}} = (K - 1)! \quad ,
+\end{equation}
+\noindent{}the Fisher information is
+\begin{equation}
+\fisher{\vec{w}} = \frac{N^{K - 1}}{\prod_{k=1}^{K} w_k} \quad ,
+\end{equation}
+\noindent{}which gives the message length of this part to be
+\begin{eqnarray}
+I(\vec{w}) &=& -\log\left(\frac{\prior{\vec{w}}}{\sqrt{{\kappa_K}^K \fisher{\vec{w}}}}\right)  \nonumber \\
+I(\vec{w}) &=& -\log{(K - 1)!} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k} \nonumber \\
+I(\vec{w}) &=& -\log{\Gamma(K)} + \frac{K}{2}\log{\kappa_K} + \frac{K-1}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{\weight_k}
+\end{eqnarray}
+\noindent{}where $\kappa_K$ is a lattice constant that varies between $1/12$
+and $1/(2\pi e)$ \cite{BoultonWallace1969}.
+
+\subsubsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
+               $K$ Gaussian components}
+
+In order to properly encode the means $\vecmean$ and covariance matrices
+$\veccov$ for all $K$ mixtures, we must encode both our prior belief on 
+those parameters and the determinant of the expected Fisher information 
+matrix. For the $k$-th mixture the joint message length is
+\begin{eqnarray}
+  I(\vecmean_k,\veccov_k) &=& -\log{\left(\frac{\prior{{\vecmean_k,\veccov_k}}}{\sqrt{\detfisher{{\vecmean_k,\veccov_k}}}}\right)} \nonumber \\ 
+  I(\vecmean_k,\veccov_k) &=& -\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\log{\detfisher{{\vecmean_k,\veccov_k}}}
+\end{eqnarray}
+
+\noindent{}and for all mixtures:
+\begin{equation}
+  I(\vecmean,\veccov) = -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \quad .
+  \label{eq:I_component_params}
+\end{equation}
+
+We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
+that 
+    $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
+However,  the MML principle requires proper priors. In practice we make 
+this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
+we do so without actually specifying the value of the bounds because they add 
+only a constant value to the total message length.
+We adopt a conjugate inverted Wishart prior for the covariance matrix of an
+individual mixture \cite[Section 5.2.3 of][]{Schafer_1997}:
+
+\begin{equation}
+  \prior{{\vecmean_k, \veccov_k}} \propto |\veccov_k|^{\frac{D+1}{2}} \quad .
+  \label{eq:covariance-prior}
+\end{equation}
+
+% Casey: Reword following paragraph because it is clumsy.
+Following Eq. \ref{eq:I_component_params}, we require the logarithm of the
+determinant of the expected Fisher information matrix for $\vecmean_k$ and
+$\veccov_k$ in order to encode the full message. In practice the determinant
+of the Fisher information is non-trivial to evaluate, and as such we
+approximate the determinant of the Fisher information
+ $\detfisher{{\vecmean_k, \veccov_k}}$ 
+as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
+\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
+we take the second derivative of the log-likelihood function 
+$\likelihood(\data|\vectheta)$, which yields:
+\begin{equation}
+  \detfisher{\vecmean_k} = (N\weight_k)^{D}|\veccov_k|^{-1} \quad .
+\end{equation}
+We make use of an analytical expression for $\detfisher{\veccov_k}$
+\cite{Dwyer_1967,Magnus_1988}:
+\begin{equation}
+  \detfisher{\veccov_k} = (N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \quad .
+\end{equation}
+Combining these expressions gives our approximation for the determinant of
+the expected Fisher information $\detfisher{{\vecmean_k,\veccov_k}}$:
+\begin{eqnarray}
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & \detfisher{\vecmean_k} \cdot \detfisher{\veccov_k} \nonumber \\
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^{D}|\veccov_k|^{-1}(N\weight_k)^\frac{D(D+1)}{2}2^{-D}|\veccov_k|^{-(D+1)} \nonumber \\
+  \detfisher{{\vecmean_k,\veccov_k}} & \approx & (N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)} \quad .
+  \label{eq:mixture-component-det-fisher}
+\end{eqnarray}
+
+Substituting Eqs. \ref{eq:covariance-prior} and \ref{eq:mixture-component-det-fisher} 
+into Eq. \ref{eq:I_component_params} gives the message length to encode the component
+parameters $\vecmean$ and $\veccov$ for all $K$ mixtures:
+
+\begin{eqnarray}
+  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{\prior{{\vecmean_k,\veccov_k}}} + \frac{1}{2}\sum_{k=1}^{K}\log{\detfisher{{\vecmean_k,\veccov_k}}} \nonumber \\
+  I(\vecmean,\veccov) &=& -\sum_{k=1}^{K}\log{|\veccov_k|^{\frac{D + 1}{2}}}
+    + \frac{1}{2}\sum_{k=1}^{K}\log{\left[(N\weight_k)^\frac{D(D+3)}{2}2^{-D}|\veccov_k|^{-(D+2)}\right]} \nonumber \\
+%  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{(D + 1)}{2}\sum_{k=1}^{K}\log{|\veccov_k|}
+    %-\frac{KD}{2}\log{2} \nonumber \\
+%  && -\frac{D+2}{2}\sum_{k=1}^{K}\log{|\veccov_k|} \nonumber \\
+  I(\vecmean,\veccov) &=& \frac{D(D+3)}{4}\sum_{k=1}^{K}\log{N\weight_k} -\frac{2D+3}{2}\sum_{k=1}^{K}\log{|\veccov_k|} - \frac{KD}{2}\log{2} 
+\end{eqnarray}
+
+\subsubsection{The data given the model}
+If we assume that the data have homoskedastic noise properties, then the 
+precision of each measurement $\mathcal{E}$ relates the probability of a
+datum $p(\datum_n)$ can be related to the probability density of
+the datum, given the model 
+$p(\datum_n) = \mathcal{E}^{D}p(y_i|\mathcal{M})$.
+In this work we adopt $\mathcal{E} = 0.01$, but we note that the value of
+$\mathcal{E}$ has no effect on our inferences.  The message length of encoding
+a datum given the model is,
+\begin{equation}
+  I(\data_n) = -\log{p(\data_n)} = -D\log\mathcal{E} - \log\sum_{k=1}^{K}\weight_{k}f_{k}(\data_n|\vecmean_k,\veccov_k)
+\end{equation}
+
+\noindent{}and for the entire data:
+\begin{eqnarray}
+  I(\data|\vectheta) &=& -ND\log\mathcal{E} - \sum_{n=1}^{N}\log\sum_{k=1}^{K}w_{k}f_k(\data_n|\vecmean_k,\veccov_k) \quad .
+\end{eqnarray}
+
+The final part of the message is to encode the lattice quantisation constant,
+which arises from the approximation to the (so-called) strict MML, where
+parameters are quantised into intervals in high dimensional space in order to
+be losslessly encoded.  We use an approximation of the lattice constant
+\citep[see Sections 5.1.12 and 3.3.4 of ]
+%[]
+{Wallace_2005} such that,
+\begin{equation}
+  \log\kappa(Q) = \frac{\log{Q\pi}}{Q} - \log{2\pi} - 1 \quad ,
+\end{equation}
+
+\noindent{}where $Q$ is the total number of free parameters in the model. Each
+Gaussian component in the mixture requires $D$ parameters to specify $\vecmean$,
+and a covariance matrix with off-diagonal terms requires $\frac{1}{2}D(D+1)$ 
+parameters. Because only $K - 1$ weights need to be encoded, the total number of free
+parameters is
+\begin{equation}
+    Q = K\left[\frac{D(D+3)}{2} + 1\right] - 1 \quad .
+    \label{eq:number-of-parameters}
+\end{equation}
+
+\subsection{Objective function}
+
+Combining the expressions for the message length (or information content) of
+each component, we arrive at our objective function $I$ to minimise across 
+all model parameters (including the number of mixtures, $K$):
+
+\begin{equation}
+\todo{I = 0}
+\end{equation}
+
+\section{Bounds on components of the message length}
+
+In Eq. \ref{eq:I1} it can be seen that for any finite mixture, $I_1$ can be
+almost directly computed just from knowing the number of mixture components
+$K$ and the properties of the data: the number of data points $N$ and its
+dimensionality $D$. The only ambiguous term is $\sum_{k=1}^{K}\log\weight_k$,
+which will depend on the data. However, for any finite mixture of $K$
+components we can consider the two extremes of $\sum_{k=1}^{K}\log\weight_k$:
+
+\begin{enumerate}
+    \item The weights are distributed uniformly such that $w_k = 1/K$ and
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
+          \end{equation}
+    \item The weights are distributed such that $K - 1$ components each 
+          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
+          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
+          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
+          %\end{equation}
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
+                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
+          \end{equation}
+
+          \noindent{}and by making use of the identity 
+            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
+          \end{equation}
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
+          %\end{equation}
+\end{enumerate}
+
+\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
+derived the bounds
+\begin{eqnarray}
+\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
+\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
+\end{eqnarray}
+\begin{equation}
+-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
+\end{equation}
+
+\noindent{}which gives us analytic expressions for the upper and lower value 
+of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
+bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.
+
+Developing similar theoretical bounds for $I_2$ is not straightforward, because
+the determinants of the covariance matrices of Gaussian components will not
+-- to our knowledge -- strictly follow some expected distribution, in the way 
+that the weights $\weights$ follow a multinomial distribution. Similarly, the
+\emph{scale} of $|\veccov_k|$ will depend entirely on the distribution of the
+data. In that sense, we cannot a priori state with any certainty about what 
+values $|\veccov_k|$ might take (or $\log{|\veccov_k|}$) without knowing
+something about the data first. Even demanding that the covariance matrices
+$\veccov_k$ are positive semi-definite only informs us that $\log{|\veccov_k|}$
+is finite. However, there are quantitative predictions we can make on the 
+bounds of $\sum_{k=1}^{K}\log{|\veccov_k|}$ if something is known about the
+data. Specifically if we consider the determinant of the sample covariance 
+matrix, which we denote $|\cov_{u}|$ (where the $u$ subscript denotes `upper',
+for the upper limit on the covariance matrix), then we can state with confidence that
+the sample covariance matrix describes the maximum amount of variance that 
+could be captured by the components in the mixture. For example, a limiting
+case of $\sum_{k=1}^{K}\log{|\veccov_k|}$ can be described when the determinant
+of the sample covariance matrix must be equally (uniformly) captured by $K$ 
+components in a mixture. In other words, the maximum variance in the data 
+must be equally spread between the components in the mixture. Uniformly 
+describing the variance in the data is the limiting case because any 
+non-uniformity would lead to smaller $|\veccov|$ for one mixture (and larger 
+for another), and that asymmetry would lead to a higher $I_2$ because 
+$\log(x - \epsilon) + \log(x + \epsilon) < 2\log{x}$, and because of the 
+$-(1 + \frac{D}{2})$ term that enters multiplicatively in the calculation of 
+$I_2$. Thus, we can specify a lower bound on $I_2$ based on the determinant of
+the sample covariance matrix $|\cov_{u}|$:
+
+\begin{eqnarray}
+    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{\frac{|\cov_u|}{K}} \nonumber \\
+    I_{2} & \geq & -\left(1 + \frac{D}{2}\right)\left(K\log{|\cov_u|} - K\log{K}\right) 
+\end{eqnarray}
+
+The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
+possible variance that could be captured by any single component. In theory
+this approaches zero, but in practice it will be set by the minimum pairwise
+distance in $D$ dimensions. Specifically, if we consider two data points to be
+captured by a component in the mixture, an those two data points are the
+closest two data points (in high dimensional space) of any data points, then
+the variance captured by that Gaussian component will represent the \emph{minimum}
+possible variance that can be captured by any single component.  
+In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
+in $D$ dimensions, average it over the dimensions, and approximate the minimum
+determinant of the covariance matrix $|\cov_l|$ as
+\begin{equation}
+    |\cov_{l}| \approx \prod_{d=1}^{D}\left(\frac{d_{min}}{2}\right)^2 \quad ,
+\end{equation}
+\noindent{}which approximately represents the variance between two data points,
+but ignores correlations and assumes only a mean distance in $D$ dimensions
+(for comuptational cost). With these approximations,
+\begin{eqnarray}
+    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\left(\frac{d_{min}}{2}\right)^2}  \nonumber \\
+    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
+\end{eqnarray}
+\noindent{}and the upper limit on $I_2$ is:
+\begin{eqnarray}
+    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
+    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
+\end{eqnarray}
+
+Here our bounds on $I_2$ are only dependent on $K$, $D$, the minimum euclidean
+pair-wise distance $d_{min}$, and the determinant of the sample covariance
+matrix $|\cov_u|$. In practice it is the lower bound on $I_{2}$ that is 
+arguably more useful for search and optimisation, but one could conceive
+situations where there is great utility in also having upper bounds on $I_1$
+and $I_2$.
+
+There is no finite upper bound on $I_{3}$ because the negative log-likelihood
+$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
+data. A useful lower limit, however, does exist. The negative log-likelihood
+in Eq. \ref{eq:log-likelihood} can be expanded as,
+\begin{equation}
+    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
+\end{equation}
+\noindent{}and we can define
+\begin{equation}
+\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
+\end{equation}
+For the sake of placing a lower bound on $I_{3}$, let us assume that each data
+point will be `mostly' explained by one of the components in the 
+mixture\footnote{We model the data with partial assignment, but here we merely 
+seek to place a lower bound on the negative log-likelihood.}, but we do not 
+necessarily know which component, or what the parameters of that component are. 
+In that situation, we can state that the component that \emph{does} explain
+the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
+whereas the other components that do not explain the data will have 
+$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
+this simplification we can write an approximation for the negative log-likelihood
+that does not require knowledge about which data points are associated to what
+component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
+will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
+approximate lower bound is:
+\begin{eqnarray}
+    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
+    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad .
+\nonumber
+    \label{eq:logl-lower-bound}
+\end{eqnarray}
+\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
+for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
+to the smallest realistic values, where those mixtures encapsulate only two
+data points and $w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N$ such that
+%\begin{eqnarray}
+%	\sum_{k=1}^{K}\weight_{k}\log\weight_k = \left[(K-1)\left(\frac{2}{N}\log{\frac{2}{N}}\right) + \left(1-\frac{2(K-1)}{N}\right)\log{\left(1-\frac{2(K-1)}{N}\right)}\right] \nonumber
+%\end{eqnarray}`
+%\noindent{}and
+\begin{eqnarray}
+	-N\sum_{k=1}^{K}\weight_{k}\log\weight_k \geq 2(1-K)\log{2} + N\log{N} - (N-2K+2)\log{\left(N - 2K + 2\right)}
+\end{eqnarray}
+%  log(a-b) = log(a) + log(1 - b/a)
+% log(1 - b/a) = log(a-b) - log(a)
+% log(1 - b/a) = log(N - 2K + 2) - log(N)
+
+% 2(1-K)log(2) - 2(1-K)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
+% 2(1-K)log(2) - 2log(N) + 2Klog(N) + (2K - N - 2)log(N - 2K + 2) - (2K - N - 2)log(N)
+% 2(1-K)log(2) + (2K - 2)log(N) + (2K - N - 2)log(N - 2K + 2) - (2K - 2 - N)log(N)
+% 2(1-K)log(2) + Nlog(N) + (2K - N - 2)log(N - 2K + 2)
+
+The remaining term in Eq. \ref{eq:logl-lower-bound}, $\frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log|\vec{\cov}_k|$, requires some  consideration.
+
+\andy{here}
+
+\subsection{Reproducibility} \label{sec:reproducibility}
+
+\noindent{}This project was developed in a public \texttt{git} repository that is 
+accessible at:
+\begin{center}
+\giturl
+\end{center}
+\noindent{}This repository includes a well-tested \texttt{Python 3.6} implementation
+of the algorithms described here, scripts to reproduce our results, and \LaTeX\ to compile this manuscript. This research can be reproduced in full using
+the following commands in a modern terminal:\\\\
+\bashcommand{git clone https://github.com/andycasey/gmmmml.git gmmmml}
+\bashcommand{cd gmmmml/}
+\bashcommand{git checkout \githash}
+\bashcommand{./reproduce}
+
+Reproducing these results will require at least \todo{X}\,Gb of free disk space
+and \todo{Y}\,hours of compute time, mostly because of the extensive algorithm 
+comparisons performed.
+
+%---------
+% COMMENTS
+%---------
+
+\begin{comment}
+We employ two criteria for comparing the proposed search methods to the benchmark: the message length and the Kullback Leibler divergence~\cite{kullback1951information}.
+
+\paragraph{Message Length}
+
+\paragraph{Kullback Leibler divergence}
+\end{comment}
+
+\begin{comment}
+
+\subsection{The MML$_{K+D}$ Method}
+Say we wanted to calculate whether another mixture was warranted. If another
+mixture were preferred then we would want:
+
+\begin{eqnarray}
+  \Delta{}I_{K+1} - I_{K} < 0
+\end{eqnarray}
+
+The expression is given as:
+\begin{eqnarray*}
+\Delta{I_{K+1} - I_K} & = & (K + 1)\log{2} - K\log{2} \\ % I_k^(new) - %I_k^(old)
+  &&+ \frac{(K)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K+1}\log{w_k}^{(new)} - \log{|\Gamma(K+1)|} \\ % I_w^(new)
+  &&- \frac{(K - 1)}{2}\log{N} + \frac{1}{2}\sum_{k=1}^{K}\log{w_k} + \log{|\Gamma(K)|}\\ % I_w^(old)
+  &&+ \mathcal{L}^{(new)} - DN\log{y_{err}} \\ % Likelihood (new)\\
+  &&- \mathcal{L}^{(old)} + DN\log{y_{err}} \\ % Likelihood (old) \\
+  &&+ \frac{1}{2}\sum_{k=1}^{K+1(new)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(new)
+  &&- \frac{1}{2}\sum_{k=1}^{K(old)}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right] \\ % I_t^(old)
+  &&- \frac{Q^{(new)}}{2}\log(2\pi) + \frac{\log{Q^{(new)}\pi}}{2} \\ % lattice, minus the lattice of part 2
+  &&+ \frac{Q^{(old)}}{2}\log(2\pi) - \frac{\log{Q^{(old)}\pi}}{2} % lattice, minus the lattice of part 2
+\end{eqnarray*}
+
+\noindent{}By making use of $\log{\Gamma(K)} - \log{\Gamma(K + 1)} = -\log{K}$ and re-arranging the expression:
+
+\begin{eqnarray}
+\Delta{}I_{K+1} - I_K &=& \log{2} % \Delta{}I_k
+    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
+& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
+& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber\\
+&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
+& +& \frac{\log(2\pi)}{2}(Q^{(old)} - Q^{(new)}) + \frac{\pi}{2}\left(\log{Q^{(new)}} - \log{Q^{(old)}}\right)
+\label{eq:13}
+\end{eqnarray}
+
+Expanding the $Q$ terms:
+
+\begin{eqnarray}
+Q^{(old)} - Q^{(new)} &=& \frac{1}{2}DK(D + 3) + K - 1 - \frac{1}{2}D(K + 1)(D + 3) + (K + 1) - 1 \nonumber \\
+Q^{(old)} - Q^{(new)} &=& -\frac{1}{2}D(D+3) + 2K  - 1
+\label{eq:14}
+\end{eqnarray}
+
+\noindent{}And making use of the following logarithmic identities,
+\begin{eqnarray}
+  \log{Q^{(new)}} &=& \log{\left(\frac{1}{2}D(K+1)(D + 3) + K\right)} \nonumber \\
+                  &=& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} + \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \\
+  \log{Q^{(old)}} &=& \log{\left(\frac{1}{2}DK(D + 3) + K - 1\right)} \nonumber \\
+                  &=& \log{\left(\frac{1}{2}DK(D + 3)\right)} + \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}
+\end{eqnarray}
+
+\noindent{}gives us,
+
+\begin{eqnarray}
+  \log{Q^{(new)}} &- \log{Q^{(old)}} = \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
+                                    &+ \log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} - \log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)}.
+\end{eqnarray}
+The second row of terms can be ignored because they are very small (typically less than 1 bit).
+This is because as $K \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 1$, thus $\log{\left(1 + \frac{K}{\frac{1}{2}D(K+1)(D + 3)}\right)} \rightarrow \log{2}$.
+Similarly as $D \rightarrow \infinity$, $2K/D(K+1)(D+3) \rightarrow 0$.
+
+As $K \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 1$ and as $D \rightarrow \infinity$ then $2(K-1)/DK(D+3) \rightarrow 0$ and thus $\log{\left(1 + \frac{K - 1}{\frac{1}{2}DK(D + 3)}\right)} \approx 0$.
+
+\noindent{}Ignoring these minor terms:
+
+\begin{eqnarray}
+  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{\left(\frac{1}{2}D(K+1)(D + 3)\right)} - \log{\left(\frac{1}{2}DK(D + 3)\right)} \nonumber \\
+  \log{Q^{(new)}} - \log{Q^{(old)}} &\approx& \log{(K + 1)} - \log{K}
+  \label{eq:19}
+\end{eqnarray}
+
+\noindent{}Substituting Eqs. \ref{eq:19} and \ref{eq:14} into \ref{eq:13} yields:
+
+\begin{eqnarray}
+\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
+    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
+& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
+& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber \\
+&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
+& +& \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
+\end{eqnarray}
+\end{comment}
+
+\begin{comment}
+\subsection{Method of Moments}
+% \alda{This is a copy paste from David's email with minor changes}
+%
+% Re the case of two (or more) classes with the same mean:
+% In 2D, a simple case in point is two classes with the same mean
+% such that $\sigma_{1, x} = \sigma_{2, y} >> \sigma_{2, x} = \sigma_{1, y}$.
+% Andy said that this and higher dimensional stuff can be picked up
+% by dropping to 1 dimension along the eigenvector of largest eigenvalue.
+%
+%
+% I wonder also about the merits of dropping to 2 dimensions (if we have at least 2 attributes), being the eigenvectors of the two largest eigenvalues. 
+%
+% One could possibly check whether uniform radially directly from the original 2D projection without having to project it to a circle (i.e., it might be pretty easy to work out what the angle of each point would be post-transformation without having to do the transformation).
+The 2D projection is 1D in the sense that we only care about the angle around the circle.
+And then some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform in radial distribution.
+If not uniform, then perhaps grounds for splitting.
+%
+% Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
+% as above but rather projecting on eigenvector of largest eigenvalue), we
+% discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution}.
+% So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
+%
+
+We informally outline here the method of moments form of search heuristic and then look at how it might be applied.
+
+Let's suppose we have a 1D component.
+We'll shift it (without loss of generality, w.l.o.g.) so that its mean is 0 and we'll shrink or expand it (again, w.l.o.g.) so that its s.d.
+($\sigma$) is 1 (and variance v is 1).
+We consider now splitting it into K components with mean 0,
+mixing weights (or proportions) $w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$ (and variances $v_i = (\sigma_i)^2$).
+Let's set $K = 2$, so $w_2 = 1 - w_1$.
+
+Noting that the 2nd, 4th, 6th and 8th moments of a true Gaussian distribution are 1, 3, 15, 105,
+suppose the moments of our current component are 1, $3 \alpha_4, 15 \alpha_6$ and $105 \alpha_8, ... $.
+We get some equations - if we have the correct number of equations,
+they're simultaneous equations.
+
+If we get too many equations (by using too many moments), then it
+becomes something of a fitting or regression problem to choose
+$w_1, ..., w_i, ..., w_K$ and s.d.s $\sigma_i$, variances $v_i = (\sigma_i)^2$.
+
+Our simultaneous equations are:
+
+\begin{align*}
+w_1 v_1       + (1 - w_1) v_2       &=    1\\
+w_1 (v_1)^2  + (1 - w_1) (v_2)^2  &=    3 \alpha_4\\
+w_1 (v_1)^3  + (1 - w_1) (v_2)^3  &=   15 \alpha_6\\
+w_1 (v_1)^4  + (1 - w_1) (v_2)^4  &=  105 \alpha_8
+\end{align*}
+
+% These might be messy to solve on the fly  but one could pre-process by solving many of these beforehand (actually, approximately solving these beforehand, as it's only
+% a heuristic, anyway) at the very start of the program while loading in the data.
+
+So, for $1000 \times 1000 = 10^6$ different values of $\alpha_4$ and $\alpha_6$
+or for $100 \times 100 \times 100 = 10^6$ different values of $\alpha_4, \alpha_6$ and $\alpha_8$
+one could pre-compute {\em approximate} 1,000,000ish `solutions' to the above simultaneous equations.
+
+Then, when running the program with real actual values of $\alpha_4$ and $\alpha_6$ (and $\alpha_8$),
+one could just grab something from nearby in this pre-computed look-up table
+as putative values for $w_1, v_1, v_2$ (and etc., if need be).
+
+% As an example of this algorithm, consider the mixture $0.6 N(-3, 2^2) + 0.4 N(7, 3^2)$.
+% The best fitting single Normal distribution to this would be $N(1, \sqrt{30}^2)$.
+%
+As an
+% another
+example of this algorithm, consider a first moment of $0$, a $2^{\rm nd}$ moment of $6$ and a $4^{\rm th}$ moment of $3 \times 37.2 = 111.6$.
+Solving the simultaneous equations gives us $0.6 N(0, 2^2) +  0.4 N(0, 3^2)$.
+\end{comment}
+
+\begin{comment}
+\subsection{The k-means++ algorithm}
+
+The k-means++ algorithm extends the k-means method with an efficient way of choosing centres.
+Let $D(x)$ denote the shortest distance from a data point $x$ to the closest centre that has already been selected.
+The k-means++ has the following steps:
+
+\begin{enumerate}
+    \item Selects a centre $c_1$ uniformly at random from the $N$ data points.
+    \item Selects a new centre $c_i$ from the data points with probability $\frac{D(x)^2}{\sum_{x\in X}}D(x)^2$
+    \item Repeat Step 2. until k centres have been selected.
+    \item Proceed as with the standard k-means algorithm.
+\end{enumerate}
+
+k-means++ is $O(\log k)$-Competitive.
+\end{comment}
+
 \end{document}

### 2018-10-02 article/ms.tex
fix travis

@@
 %\usepackage{boldsymbol}
-\usepackage{minted}
 \usepackage{color,colortbl} % for \todo command
@@
 The \emph{Component-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.
+
+\todo{initialisation, how kmeans++ breaks when too big unless covariance regularisation\\}
+
+\todo{Run E-M, make predictions\\}
+
+\todo{Advantage in being able to jump many $\Delta{}K$ at once}
 
@@
 
-The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms.
-The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.
-\andy{Huh?}
+\andy{The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms.
+The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.}
 
 \subsection{Benchmark Method}
-We use the search algorithm proposed by Kasarapu and Allison as a benchmark for comparison, which
-has been shown to be the state-of-the art
+We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison, which has been shown to be the state-of-the-art.
 % in what sense has this been shown?

### 2018-10-03 gmmmml/gmm.py
minor changes to scripts

@@ -7,17 +7,10 @@ __all__ = ["GaussianMixture"]
 
 import logging
 import numpy as np
-import scipy
-from collections import (defaultdict, OrderedDict)
+from collections import OrderedDict
 from time import time
-from tqdm import tqdm
-from scipy.special import erf
-from scipy.signal import find_peaks_cwt
-from sklearn import cluster
-from sklearn.utils import check_random_state
-from sklearn.utils.extmath import row_norms
 
-from . import (mml, em, operations, strategies)
+from . import (mml, em, strategies)
 
 logger_name, *_ = __name__.split(".")
 logger = logging.getLogger(logger_name)
@@ -331,11 +324,11 @@ class GaussianMixture(object):
                 search_strategy_class = getattr(strategies, search_strategy)
 
             except AttributeError:
-                raise ValueError("unrecognised strategy: '{}'".format(strategy))
+                raise ValueError(f"unrecognised strategy: '{search_strategy}'")
 
         else:
             # Assume strategy is a strategy class.
-            search_strategy_class = strategy
+            search_strategy_class = search_strategy
 
         logger.info(f"Using search strategy '{search_strategy_class.__name__}'")
 

### 2018-10-03 gmmmml/utils.py

@@ -354,22 +354,28 @@ def generate_data(N, D, K=None, dirichlet_concentration=1, isotropy=10, psi=0.05
     phi = np.abs(np.random.normal(0, psi, size=(K, D)))
 
     I = np.eye(D)
-    covariances = np.empty((K, D, D))
+    covs = np.empty((K, D, D))
     for k, diag in enumerate(phi):
 
-        covariances[k] = I * diag
+        covs[k] = I * diag
 
         for i, j in zip(*np.tril_indices(D, -1)):
             value = rho(1) * diag[i] * diag[j]
-            covariances[k, i, j] = covariances[k, j, i] = value
+            covs[k, i, j] = covs[k, j, i] = value
 
     # Draw samples from each component to generate the data.
     members = np.round(N * weights).astype(int)
+    if members.sum() > N:
+        # Drop a few.
+        diff = members.sum() - N
+        idx = np.random.choice(np.where(members > 2)[0], diff, replace=False)
+        members[idx] -= 1
+
 
     X = np.empty((N, D))
     R = np.zeros(N, dtype=int)
 
-    for i, (m, mean, cov) in enumerate(zip(members, means, covariances)):
+    for i, (m, mean, cov) in enumerate(zip(members, means, covs)):
         si = int(np.sum(members[:i]))
         X[si:si + m] = np.random.multivariate_normal(mean, cov, size=m)[:N-si]
         R[si:si + m] = i
@@ -383,7 +389,7 @@ def generate_data(N, D, K=None, dirichlet_concentration=1, isotropy=10, psi=0.05
                 responsibilities=R, truths=dict(
                     means=means,
                     weights=weights,
-                    covariances=covariances
+                    covs=covs
                 ))
 
     return (X, meta)

### 2018-10-03 scripts/evaluate_search_methods.py

@@ -1,0 +1,87 @@
+
+"""
+Evaluate the performance of search strategies on generated data.
+"""
+
+import numpy as np
+import os
+import pickle
+from collections import OrderedDict
+from glob import glob
+from time import time
+
+from gmmmml import GaussianMixture
+
+gmm_kwds = dict(threshold=1e-5, 
+                max_em_iterations=1000,
+                covariance_regularization=1e-10)
+
+search_strategies = OrderedDict([
+    ("BayesStepper", dict()),
+    ("KasarapuAllison2015", dict())
+])
+
+data_paths = glob("data/*.data")
+
+overwrite = False
+raise_if_not_converged = True
+
+# Initialise the evaluations.
+N = len(data_paths)
+ml = lambda I: np.sum(np.hstack(I.values()))
+
+results_path_format = "{data_path}.{search_strategy}.output"
+
+for search_strategy, search_kwds in search_strategies.items():
+
+    for i, data_path in enumerate(data_paths):
+
+        results_path = results_path_format.format(data_path=data_path, 
+                                          search_strategy=search_strategy)
+
+        if os.path.exists(results_path) and not overwrite:
+            print(f"Skipping {i}/{N} ({data_path}) using {search_strategy} "\
+                  f"because {results_path} exists and not overwriting")
+            continue
+
+        print(f"Running {i}/{N} ({data_path}) using {search_strategy}")
+
+        with open(data_path, "rb") as fp:
+            draws = pickle.load(fp)
+
+        results = []
+        for j, (X, meta) in enumerate(draws):
+
+            model = GaussianMixture(**gmm_kwds)
+
+            tick = time()
+
+            try:
+                model.search(X, search_strategy=search_strategy, **search_kwds)
+
+            except:
+                I = np.inf
+
+            else:
+                I = ml(model.message_lengths_)
+
+            finally:
+                tock = time()
+
+            # Did we converge to the global solution?
+            # (We gauge this by evaluating if we found a message length longer
+            # than what is required to encode with the 'true' parameters)
+            R_t, ll_t, I_t = model.expectation(X, **meta["truths"])
+            I_t = ml(I_t)
+
+            converged = (I <= I_t)
+
+            results.append((converged, I, I_t, tock - tick))
+
+            print(converged, I, I_t, tock - tick)
+
+            assert converged or not raise_if_not_converged
+
+        with open(results_path, "wb") as fp:
+            pickle.dump(results, fp)
+

### 2018-10-03 scripts/generate_test_data.py

@@ -10,33 +10,46 @@ import pickle
 from gmmmml import utils
 
 
-random_seed = 42
+np.random.seed(123)
 
-path_format = "data/{K:05d}_data.pkl"
+M = 1 # number of draws per unit of N, K, D
+
+max_N, max_D = 10**4, 30
+
+orders = np.atleast_2d(np.logspace(0, np.log10(max_N), 1 + np.log10(max_N))[1:])
+steps = np.atleast_2d(np.round(np.logspace(0, 1, 4)).astype(int))
+
+Nu = np.unique((orders.T @ steps).flatten())
+
+Du = np.logspace(np.log10(2), np.log10(max_D), 6).astype(int)
+
+N, D = np.meshgrid(Nu, Du)
+
+K = np.clip(np.round([np.random.uniform(2, 0.1 * n) for n in N]),
+            2, np.inf)
+
+path_format = "data/{K:05d}_{N:06d}_{D:01d}.data"
 
 dirname = os.path.dirname(path_format)
 if not os.path.exists(dirname):
     os.makedirs(dirname, exist_ok=True)
 
 
-Ks = np.round(np.logspace(0, 3, 7)).astype(int)
+common_kwds = dict(dirichlet_concentration=100, isotropy=10, psi=0.01)
 
-common_kwds = dict(D=5, dirichlet_concentration=100, isotropy=10, 
-                   psi=0.001, random_seed=random_seed)
+N, D, K = (ea.flatten().astype(int) for ea in (N, D, K))
 
+for i, (n, d, k) in enumerate(zip(N, D, K)):
 
+    print(f"Generating data for N = {n}, D = {d}, K = {k}")
 
-for K in Ks:
-    print(f"Generating data with K = {K} and {common_kwds}")
+    draws = []
+    for m in range(M):
+        draws.append(utils.generate_data(N=n, D=d, K=k, **common_kwds))
 
-    X, meta = utils.generate_data(K=K, N=10*K, **common_kwds)
-
-    with open(path_format.format(K=K), "wb") as fp:
-        pickle.dump((X, meta), fp, -1)
-
-
-
-
+    kwds = dict(N=n, D=d, K=k)
+    with open(path_format.format(**kwds), "wb") as fp:
+        pickle.dump(draws, fp)
 
 
 

### 2018-10-03 scripts/plot_evaluations.py

@@ -1,0 +1,4 @@
+
+"""
+Plot the results from the evaluations on artificial data.
+"""

### 2018-10-03 article/ms.tex
reorder benchmark method

@@
 We introduce and evaluate two search strategies that make use of the objective function in Equation \ref{eq:objective_function}.
-These search strategies are compared against an existing state-of-the-art method, which we describe in Section \ref{sec:experimental-design}.
+These strategies are benchmarked against an existing state-of-the-art search method, which we have adapted to use the same objective function in Equation~\ref{eq:objective_function} in order to facilitate a fair comparison.
+
+\subsection{Benchmark Method}
+
+We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison.
+The method is presented in Algorithm \ref{alg:kasarapu}.
+It starts with a single component mixture and iteratively splits (lines \ref{line:splitBegin}-\ref{line:splitEnd}), deletes (lines \ref{line:deleteBegin}-\ref{line:deleteEnd}), and merges (lines \ref{line:MergeBegin}-\ref{line:MergeEnd}) components until no improvements in the message length is observed (line~\ref{line:end}).
+This is a greedy approach, which is advantageous in that it is guaranteed to only accept perturbations that minimise the message length, but that guarantee comes with a computational cost.
+% in what sense has this been shown?
+\begin{algorithm}[H]
+\caption{The benchmark method~\cite{kasarapu2015minimum}.}
+\label{alg:kasarapu}
+{\bf Input:} $\vec\data$ \\
+{\bf Output:} $\vec\psi$
+\begin{algorithmic}[1]
+\State current $\leftarrow$ one component mixture
+\While{TRUE}
+\State components $\leftarrow$ current
+  \For{$i\in K$}\label{line:splitBegin}\Comment{Exhaustively split all components}
+    \State splits[i]$\leftarrow$ \Call{Split}{current,components[i]}
+  \EndFor
+\State  bestSplit$\leftarrow$\Call{Best}{splits}\label{line:splitEnd} \Comment{Record the best split.}
+\If{K>1}
+  \For{$i\in K$}\label{line:deleteBegin}\Comment{Exhaustively delete all components}
+    \State deletes[i]$\leftarrow$ \Call{Delete}{current,components[i]}
+  \EndFor
+\State  bestDelete$\leftarrow$\Call{Best}{deletes} \label{line:deleteEnd}\Comment{Record the best delete.}
+\EndIf
+  \For{$i\in K$}\label{line:MergeBegin}\Comment{Exhaustively merge all components}
+    \State j$\leftarrow$ \Call{closestComponent}{i}
+    \State merges[i]$\leftarrow$ \Call{Merge}{merges}
+  \EndFor
+\State  bestMerge$\leftarrow$\Call{Best}{merges} \label{line:MergeEnd}\Comment{Record the best merge.}
+\State bestPerturbation $\leftarrow$ \Call{Best}{bestSplit, bestDelete,bestMerge} \Comment{Select the best perturbation.}
+\State $\Delta I \leftarrow$ \Call{MessageLength}{bestPerturbation}-\Call{MessageLength}{current} \Comment{Check for improvement.}
+\If{$\Delta I < 0$}
+\State current $\leftarrow$ bestPerturbation
+\State {\bf \sc Continue}
+\EndIf
+\State {\bf \sc Break} \label{line:end}
+\EndWhile
+\end{algorithmic}
+\end{algorithm}
 
 \subsection{The Message-Breaking Method} \label{sec:mbm}
-The first strategy, \emph{The Message-Breaking Method}, does not make any predictions on
-the message length of future mixtures. Instead it starts from 
+The simplest strategy we introduce here, \emph{The Message-Breaking Method}, does not use any predictions of the message length of future mixtures. Instead it starts from 
 a $K=1$ mixture and iteratively splits the component with the largest message length until 
-the total message length no longer improves (Algorithm~\ref{alg:MBM}). We evaluate the message length of each component
-by re-ordering Equation \ref{eq:objective_function} and weighting the negative log-likelihood
-$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ with entries
+the total message length no longer improves (Algorithm~\ref{alg:MBM}).
+This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split components (instead of splitting, merging, and deleting all components).
+Similarly, in \emph{The Message-Breaking Method} we only split the component with the largest message length instead of exhaustively trying to split all components at every mixture.
+
+We evaluate the message length of each component by rearranging Equation \ref{eq:objective_function} and multiplying the negative log-likelihood
+$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ that has entries
 \begin{eqnarray}
@@
 The component-wise calculation of the message length is performed as part of the 
-expectation step of the expectation-maximisation algorithm, adding negligible computational overhead.
+expectation step of the expectation-maximisation algorithm, adding negligible computational overhead to the algorithm.
 
@@
 
+\andy{Should we move/remove the following paragraph?}
 One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
@@
 
-\subsection{Benchmark Method}
-We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison, which has been shown to be the state-of-the-art.
-% in what sense has this been shown?
-~\cite{kasarapu2015minimum}. 
-\begin{algorithm}[H]
-\caption{The benchmark method~\cite{kasarapu2015minimum}.}
-\label{alg:kasarapu}
-\begin{algorithmic}[1]
-\State {\bf \sc input:} $K$ \Comment{Number of components.}
-\State current $\leftarrow$ one component mixture
-\While{TRUE}
-\State components $\leftarrow$ current
-  \For{$i\in K$}\label{line:splitBegin}\Comment{Exhaustively split all components}
-    \State splits[i]$\leftarrow$ \Call{Split}{current,components[i]}
-  \EndFor
-\State  bestSplit$\leftarrow$\Call{Best}{splits}\label{line:splitEnd} \Comment{Record the best split.}
-\If{K>1}
-  \For{$i\in K$}\label{line:deleteBegin}\Comment{Exhaustively delete all components}
-    \State deletes[i]$\leftarrow$ \Call{Delete}{current,components[i]}
-  \EndFor
-\State  bestDelete$\leftarrow$\Call{Best}{deletes} \label{line:deleteEnd}\Comment{Record the best delete.}
-\EndIf
-  \For{$i\in K$}\label{line:MergeBegin}\Comment{Exhaustively merge all components}
-    \State j$\leftarrow$ \Call{closestComponent}{i}
-    \State merges[i]$\leftarrow$ \Call{Merge}{merges}
-  \EndFor
-\State  bestMerge$\leftarrow$\Call{Best}{merges} \label{line:MergeEnd}\Comment{Record the best merge.}
-\State bestPerturbation $\leftarrow$ \Call{Best}{bestSplit, bestDelete,bestMerge} \Comment{Select the best perturbation.}
-\State $\Delta I \leftarrow$ \Call{MessageLength}{bestPerturbation}-\Call{MessageLength}{current} \Comment{Check for improvement.}
-\If{$\Delta I < 0$}
-\State current $\leftarrow$ bestPerturbation
-\State {\bf \sc Continue}
-\EndIf
-\State {\bf \sc Break} \label{line:end}
-\EndWhile
-\State {\bf \sc output:} current
-\end{algorithmic}
-\end{algorithm}
-
-The method is presented in Algorithm~\ref{alg:kasarapu}.
-It starts with one component and iteratively splits (lines \ref{line:splitBegin}-\ref{line:splitEnd}), deletes (lines \ref{line:deleteBegin}-\ref{line:deleteEnd}), and merges (lines \ref{line:MergeBegin}-\ref{line:MergeEnd}) components until no improvements in the message length is observed (line~\ref{line:end}).
-
-\subsection{Dataset}
-
-\alda{Andy, can you please include the details?}
+\subsection{Generated Datasets}
+
+\todo{How we generated data sets}
+
+\todo{Range of generated data sets}
+
+\todo{Figure: predictions for future mixtures}
+
+\todo{Figure: computational cost of various methods}
+
+\todo{Figure: corner plot showing chemical abundance data}
+
+\todo{Figure: corner plot showing chemical abundance data, coloured by their responsibility}
+
+\todo{Figures showing the concentration of chemically-identified data}
+
 \paragraph{Simulations}
 
-\paragraph{GA data}
-
-\subsection{Criteria for Comparison}
+\paragraph{Criteria for Comparison}
+
+\subsection{Galactic Archaeology data}
 
 \section{Results}
+
+\todo{Convergence criteria: how often do we reach the `true' solution?}
+
+\todo{What is the variance in the results of our method?}
+
+\todo{Speed comparisons}
+
+\todo{What do we find for the chemical abundance data}
+
+\todo{Recovering known star clusters that are still gravitationally bound?}
+
+\todo{Scientific insight from identifying groups in these data?}
+
+\todo{Estimate of the number of star-forming sites in the galaxy?}
+
 \section{References}
@@
 
+\subsubsection{Search Strategy Implementation}
+
+The astute reader will recognise that each search `strategy' we have described
+is implemented in the \texttt{gmmmml} codebase as a set of `policies' that govern
+different behaviour of the search:
+\begin{itemize}
+\item \emph{Initialisation}: how many mixtures should be initialised, and how?
+\item \emph{Prediction}: when should we make predictions about the message length of future mixtures, and how far ahead should we make predictions?
+\item \emph{Movement}: \emph{what} should the next $K$ be?
+\item \emph{Repartition}: \emph{how} should we move to the next $K$? Should we repartition an existing mixture, initialise a new mixture, or should we make a probabilistic decision?
+\item \emph{Convergence}: when should we stop searching?
+\end{itemize}
+
+As an example, the \emph{Message-Breaking Method} can equivalently be described
+by a set of policies that: 
+	\emph{initialise} a single mixture $K = 1$; 
+	makes no \emph{predictions};
+	\emph{moves} a single step towards the minimum message length;
+	\emph{repartitions} the closest existing mixture by splitting the component with the maximum message length; and 
+	describes \emph{convergence} at the point when the message length of the current mixture is worse than the previous mixture. 
+This policy-strategy factorisation allows for the trivial implementation of flexible search strategies with complex behaviour. We encourage the development of new search strategies.
+
 %---------

### 2018-10-03 scripts/compare_search_methods.py
scripts to make and plot evaluations

@@ -1,91 +1,0 @@
-
-"""
-Compare methods on generated data.
-"""
-
-from glob import glob
-import numpy as np
-import matplotlib.pyplot as plt
-import pickle
-
-from gmmmml import GaussianMixture
-
-
-# Methods:
-# - greedy-kmeans
-# - bayes-jumper
-# - kasarapu
-
-data_paths = glob("data/?????_data.pkl")
-
-data_paths = sorted(data_paths)
-
-gmm_kwds = dict(threshold=1e-5, 
-                max_em_iterations=1000,
-                covariance_regularization=1e-10)
-
-comparison_kwds = dict([
-    ("KasarapuAllison2015", dict()),
-    ("BayesJumper", dict())
-
-
-times = {}
-K_est = {}
-
-
-for i, data_path in enumerate(data_paths):
-
-    with open(data_path, "rb") as fp:
-        X, meta = pickle.load(fp)
-
-    times.setdefault(data_path, dict(meta=meta))
-    K_est.setdefault(data_path, dict())
-
-
-    for j, (strategy, search_kwds) in enumerate(comparison_kwds.items()):
-        assert strategy != "meta", "This will stuff things up"
-
-        model = GaussianMixture(**gmm_kwds)
-        try:
-            model.search(X, search_strategy=strategy, **search_kwds)
-
-        except:
-            result = np.nan
-            K_ = np.nan
-
-        else:
-            result = model.meta_["t_search"]
-            K_ = model.weights_.size
-
-        times[data_path][strategy] = result
-        K_est[data_path][strategy] = K_
-
-
-# Plot times.
-strategies = comparison_kwds.keys()
-
-K = dict([(k, times[k]["meta"]["K"]) for k in times.keys()])
-t = dict()
-for strategy, kwds in comparison_kwds.items():
-    t.setdefault(strategy, [])
-
-x = []
-y = []
-for data_path, k in K.items():
-    x.append(k)
-
-    y_row = []
-    for strategy in strategies:
-        y_row.append(times[data_path][strategy])
-
-    y.append(y_row)
-
-x, y = (np.array(x), np.array(y))
-
-fig, ax = plt.subplots()
-for i, strategy in enumerate(strategies):
-    ax.scatter(x, y.T[i], label=strategy)
-
-plt.legend()
-
-raise a

### 2018-10-03 scripts/evaluate_search_methods.py

@@ -21,10 +21,10 @@ search_strategies = OrderedDict([
     ("KasarapuAllison2015", dict())
 ])
 
-data_paths = glob("data/*.data")
+data_paths = sorted(glob("data/*.data"))
 
 overwrite = False
-raise_if_not_converged = True
+raise_if_not_converged = False
 
 # Initialise the evaluations.
 N = len(data_paths)
@@ -76,7 +76,11 @@ for search_strategy, search_kwds in search_strategies.items():
 
             converged = (I <= I_t)
 
-            results.append((converged, I, I_t, tock - tick))
+            result = dict(K=meta["K"], N=meta["N"], D=meta["D"],
+                          converged=(I <= I_t),
+                          I=I, I_t=I_t, time=tock - tick)
+
+            results.append(result)
 
             print(converged, I, I_t, tock - tick)
 

### 2018-10-03 scripts/plot_evaluations.py

@@ -2,3 +2,130 @@
 """
 Plot the results from the evaluations on artificial data.
 """
+
+# TODO: Get these from somewhere else?
+
+import numpy as np
+import matplotlib.pyplot as plt
+import scipy.optimize as op
+import pickle
+from collections import OrderedDict
+
+from glob import glob
+
+search_strategies = OrderedDict([
+    ("BayesStepper", dict()),
+    ("KasarapuAllison2015", dict())
+])
+
+results_path_template = "data/*{search_strategy}.output"
+
+times = dict()
+result_keys = ["K", "N", "D", "I", "I_t", "time"]
+        
+for search_strategy in search_strategies.keys():
+
+    results_paths = glob(
+        results_path_template.format(search_strategy=search_strategy))
+    N = len(results_paths)
+
+    print(f"Collecting {N} result files on {search_strategy}")
+
+    if N < 1:
+        continue
+
+    times[search_strategy] = []
+
+    for i, results_path in enumerate(results_paths):
+
+        print(f"At {i}/{N}: {results_path}")
+
+        with open(results_path, "rb") as fp:
+            results = pickle.load(fp)
+
+        # Average draws?
+        assert len(results) == 1
+
+        for result in results:
+            times[search_strategy].append([result[k] for k in result_keys])
+
+
+    times[search_strategy] = np.array(times[search_strategy])
+
+
+# Plot as a function of N, D, K
+
+x_labels = ["N", "D", "K", "ND", "NK"]
+y_label = "time"
+
+fig, axes = plt.subplots(1, len(x_labels))
+
+
+max_y = 0
+upper_lim = lambda existing: 10**(1 + np.ceil(np.log10(np.max(existing))))
+
+for i, (ax, x_label) in enumerate(zip(axes, x_labels)):
+
+    try:
+        x_idx = result_keys.index(x_label)
+    
+    except ValueError:
+        x_idx = np.array([result_keys.index(_) for _ in x_label])
+
+    y_idx = result_keys.index(y_label)
+
+    max_x = 0
+    for search_strategy, data in times.items():
+
+        x = data.T[x_idx]
+        y = data.T[y_idx]
+
+        if len(x.shape) > 1:
+            x = np.product(x, axis=0)
+
+        idx = np.argsort(x)
[…]

### 2018-10-03 scripts/evaluate_search_methods.py
betterised plots for evaluating search strategy

@@ -17,8 +17,8 @@ gmm_kwds = dict(threshold=1e-5,
                 covariance_regularization=1e-10)
 
 search_strategies = OrderedDict([
+    ("KasarapuAllison2015", dict()),
     ("BayesStepper", dict()),
-    ("KasarapuAllison2015", dict())
 ])
 
 data_paths = sorted(glob("data/*.data"))
@@ -28,7 +28,7 @@ raise_if_not_converged = False
 
 # Initialise the evaluations.
 N = len(data_paths)
-ml = lambda I: np.sum(np.hstack(I.values()))
+ml = lambda I: I if isinstance(I, float) else np.sum(np.hstack(I.values()))
 
 results_path_format = "{data_path}.{search_strategy}.output"
 

### 2018-10-03 scripts/plot_evaluations.py

@@ -58,7 +58,8 @@ for search_strategy in search_strategies.keys():
 x_labels = ["N", "D", "K", "ND", "NK"]
 y_label = "time"
 
-fig, axes = plt.subplots(1, len(x_labels))
+L = len(x_labels)
+fig, axes = plt.subplots(1, L, figsize=(4 * L, 4))
 
 
 max_y = 0
@@ -83,13 +84,19 @@ for i, (ax, x_label) in enumerate(zip(axes, x_labels)):
         if len(x.shape) > 1:
             x = np.product(x, axis=0)
 
+        converged = data.T[result_keys.index("I")] <= data.T[result_keys.index("I_t")]
+
+
         idx = np.argsort(x)
-        x, y = x[idx], y[idx]
+        x, y, converged = x[idx], y[idx], converged[idx]
 
         max_y = max(y.max(), max_y)
         max_x = max(x.max(), max_x)
 
-        ax.scatter(x, y, label=search_strategy)
+
+        scat = ax.scatter(x[converged], y[converged], label=search_strategy)
+
+        ax.scatter(x[~converged], y[~converged], alpha=0.5, c=scat.get_facecolor())
 
     ax.loglog()
     ax.set_xlim(1e-0, upper_lim([max_x]))
@@ -101,9 +108,10 @@ for i, (ax, x_label) in enumerate(zip(axes, x_labels)):
 for ax in axes:
     ax.set_ylim(1e-2, upper_lim([max_y]))
 
+fig.tight_layout()
 
 
-x_labels = ["N", "D", "K"]
+x_labels = ["K", "N", "D"]
 
 
 # Fit the cost.
@@ -113,9 +121,6 @@ for search_strategy, data in times.items():
     y_idx = result_keys.index(y_label)
 
     log_x = np.log(data.T[x_idx])
-
-    # Include N*D as a param.
-    log_x = np.vstack([log_x, np.product(data.T[x_idx[:2]], axis=0).reshape(1, -1)])
     log_y = np.log(data.T[y_idx])
 
     f = lambda _, *p: p @ log_x
@@ -124,8 +129,8 @@ for search_strategy, data in times.items():
     p_opt, p_cov = op.curve_fit(f, np.ones(j), log_y, p0=np.ones(i))
 
     # Plot the time relative to K
+    order_repr = "".join([f"[{x}^{p:.1f}]" for x, p in zip(x_labels, p_opt)])
 
-
-    raise a
+    print(f"{search_strategy}: O({order_repr})")
 
 raise a

### 2018-10-15 gmmmml/policies/convergence.py
updates

@@ -17,6 +17,8 @@ class BaseConvergencePolicy(Policy):
 
 class DefaultConvergencePolicy(BaseConvergencePolicy):
     
+    num_consecutively_worse_trials = 2
+
     @property
     def converged(self):
 
@@ -32,22 +34,32 @@ class DefaultConvergencePolicy(BaseConvergencePolicy):
         _, __ = np.unique(K_all, return_index=True)
         K_unique = K_all[np.sort(__)]
 
-        prev_K, K = K_unique[-2:]
-        
-        
-        I = np.min(I_all[np.where(K_all == K)[0]])
-        prev_I = np.min(I_all[np.where(K_all == prev_K)[0]])
+        N = self.num_consecutively_worse_trials
+        prev_K_trials = K_unique[-N:]
+        best_I_trials = np.array([np.min(I_all[K_all==K]) for K in prev_K_trials])
 
-        delta = I - prev_I
-        
+
         # Stop if the current message length is more than the best from the
-        # previous mixture.
-        converged = delta > 0
+        # previous mixture(s)
+        diffs = np.diff(best_I_trials)
+        converged = np.all(diffs > 0)
 
-        logger.info(f"Convergence threshold: {I:.0f} (@ K = {K}) < {prev_I:.0f}"\
-                    f" (@ prev_K = {prev_K}) ({delta:.0f})")
+        prev_I, I = best_I_trials[-2:]
+        prev_K, K = prev_K_trials[-2:]
+        delta = diffs[-1]
+
+        logger.info(f"Convergence threshold (checked last {N} unique trials): "\
+                    f"{I:.0f} (@ K = {K}) < {prev_I:.0f}"\
+                    f" (@ prev_K = {prev_K}) ({delta:.0f}; {diffs})")
         
         return converged
+
+
+class StrictConvergencePolicy(DefaultConvergencePolicy):
+
+    num_consecutively_worse_trials = 5
+
+
 
 
 class ConvergedWithSuccessivelyWorseIterations(BaseConvergencePolicy):

### 2018-10-15 gmmmml/strategies.py

@@ -14,6 +14,13 @@ class BayesStepper(initialisation.DefaultInitialisationPolicy,
     pass
 
 
+class StrictBayesStepper(initialisation.DefaultInitialisationPolicy,
+                         prediction.DefaultPredictionPolicy,
+                         movement.StepTowardsMMLMixtureMovementPolicy,
+                         repartition.IterativelyRepartitionFromNearestMixturePolicy,
+                         convergence.StrictConvergencePolicy):
+    pass
+
 class BayesJumper(initialisation.LogarithmicInitialisationPolicy,
                   prediction.DefaultPredictionPolicy,
                   movement.JumpToMMLMixtureMovementPolicy,

### 2018-10-15 scripts/run_apogee.py

@@ -11,15 +11,15 @@ from astropy.table import Table
 
 from gmmmml import GaussianMixture
 
-gmm_kwds = dict(threshold=1, 
-                max_em_iterations=100,
+gmm_kwds = dict(threshold=1e-3, 
+                max_em_iterations=10000,
                 covariance_regularization=1e-10)
-search_kwds = dict(search_strategy="BayesStepper")
+search_kwds = dict(search_strategy="StrictBayesStepper")
 
 apogee_input_path = "../ct/catalogs/tc-apogee-dr12-regularized-release.fits"
 
 overwrite = False
-output_path = "apogee-results.pkl"
+output_path = "apogee-results-strict.test.pkl"
 
 if os.path.exists(output_path) and not overwrite:
     raise IOError(f"output path '{output_path}' exists and not overwriting")
@@ -45,7 +45,7 @@ result = dict(gmm_kwds=gmm_kwds, search_kwds=search_kwds, time_taken=tock - tick
               message_lengths=model.message_lengths_,
               meta=model.meta_)
 
-with open(output_path, "rb") as fp:
+with open(output_path, "wb") as fp:
     pickle.dump(result, fp)
 
 print(f"Search {search_kwds} found K = {model.weights_.size}")

### 2018-10-15 article/ms.tex
words

@@
 % Have some maaaath.
+\def\transpose{^\intercal}
 \def\infinity{\rotatebox{90}{8}}
@@
 As a consequence, where a star is now does not reflect its formation site, 
-complicating our understanding of galaxy formation and evolution. However, the
+which complicates our understanding of galaxy formation and evolution. However, the
 observable chemical composition of a star remains largely the same throughout
 that star's lifetime, providing an immutable marker (a chemical tag, or
-`fingerprint') that carries with it information about where and when star
+`fingerprint') that carries with it information about where and when every star
 formed.
@@
 clusters can be identified solely on the basis of stellar chemical
-abundances~\cite{hogg2016kmeans}. Astronomers have precisely measured the
-chemical composition for nearly one million stars in the disk of our galaxy,
-where the natal formation site for most stars is not known~\cite{galah}.
+abundances~\cite{Hogg:2016}. Astronomers have precisely measured the
+chemical composition for hundreds of thousands of stars in the disk of our galaxy,
+where the natal formation site for most stars is not known~\cite{Buder:2018}.
 Indeed, from those data it is not known how many star formation sites we might
@@
 might to observe from each cluster (the so-called stellar mass function).
-Theoretical simulations can provide a guide to these quantities (e.g., of the
-order X star forming sites that have total masses less than Y), but the
-predictions have a large variance.
+Theoretical simulations can provide a guide to these quantities, but the
+predictions are still uncertain.
+% large variance.
 
@@
 \noindent{}Mixture modelling is composed of three main parts:
-\begin{itemize}
-    \item an estimator of component parameters of a mixture,
-    \item an objective functions that scores a mixture, and 
+\begin{enumerate}
+    \item an estimator of component parameters of a mixture;
+    \item an objective functions that scores a mixture; and 
     \item a search strategy which identifies the best number of components and
           their relative weights.
-\end{itemize}
+\end{enumerate}
 
@@
 over-fitting. In this work, we focus on the Bayesian information-theoretic 
-Minimum Message Length (MML) principle~\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a, Wallace05}.
+Minimum Message Length (MML) principle~\cite{WallaceBoulton1968, WallaceFreeman1987, WallaceDowe1999a, Wallace:2005}.
 Unlike MLE or MAP estimators, the MML principle is invariant under non-linar
@@
 %
-% In this work, we focus on the Bayesian Minimum Message Length Principle~\cite{Wallace05}, which unlike MAP estimators, are invariant under non-linear data transformations.
+% In this work, we focus on the Bayesian Minimum Message Length Principle~\cite{Wallace:2005}, which unlike MAP estimators, are invariant under non-linear data transformations.
 %, and different from MLE, can be used with different probability distributions.  
@@
 The classically-described princple of MML is that the best explanation of the
-data is the one that leads to the shortest so-called two-part message~\cite{Wallace05}, 
+data is the one that leads to the shortest so-called two-part message~\cite{Wallace:2005}, 
 where a \textit{message} takes into account both the complexity of the model 
@@
 \cite{viswanathan1999finding,fitzgibbon2004minimum}, with references to 
-further examples in \cite{Wallace05,dowe2007bayes,Dowe2008a,Dowe2011a}).
+further examples in \cite{Wallace:2005,dowe2007bayes,Dowe2008a,Dowe2011a}).
 
@@
 
-%Unlike the rival Bayesian approach MAP, for continuous parameters MML is a statistically invariant (meaning that $1$-to-$1$ transformations of co-ordinates, such as polar $(r, \theta)$ to Cartesian $(x, y)$, leave the answer unchanged) Bayesian approach~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}.
+%Unlike the rival Bayesian approach MAP, for continuous parameters MML is a statistically invariant (meaning that $1$-to-$1$ transformations of co-ordinates, such as polar $(r, \theta)$ to Cartesian $(x, y)$, leave the answer unchanged) Bayesian approach~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace:2005,dowe2007bayes}.
 It allows the modelling of prior beliefs in the calculation of the best model parameters.
@@
 number of components, and iteratively deletes components. A mirror extension 
-of this method was proposed by Kasarapu and Allison~\cite{kasarapu2015minimum}, 
+of this method was proposed by Kasarapu and Allison~\cite{Kasarapu|Allison:2015}, 
 which starts with one components, and selectively splits, deletes, or merges 
@@
 %Things that could also be added include
-%ConwaySloane1988
+%!TEX encoding = UTF-8 UnicodeConwaySloane1988
 % the above is a book on lattices and lattice constants
@@
 simplicity and truth, as beautifully elucidated by the Occam's Razor principle. 
-MML is a statistical technique that formalizes and quantifies this principle.
+MML is a statistical technique that attempts to formalize and quantify this principle.
 
@@
 
-%The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace05}, 
+%The MML principle states that the best explanation of the data is the one that leads to the shortest message~\cite{Wallace:2005}, 
 %where a \textit{message} takes into account both the complexity of the model, and its explanatory power.
@@
 The \textit{message} must encode two parts: the model, and the data given the
-model. The encoding of the message is based on Shannon's information theory~\cite{shannon}. 
+model. The encoding of the message is based on Shannon's information theory~\cite{Shannon:1948}. 
 The information gained from an event $e$ occurring, where $p(e)$ is the
@@
 I think that, here, all we need to do is mention Morse code and (e.g.) the Huffman code example from \cite[sec.
-2.1.4]{Wallace05}.
+2.1.4]{Wallace:2005}.
 % Consider the simplified illustrative example of rolling an unbiased four-sided die (d4), where each of the four outcomes $(1,2,3,4,5,6)$ have equal probability $p_i=0.25$.
@@
 here correspond to probabilities, but with the benefit of statistical 
-invariance under re-parameterisation (e.g., ~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace05,dowe2007bayes}).
+invariance under re-parameterisation (e.g., ~\cite{WallaceFreeman1987,WallaceDowe1999a,Wallace:2005,dowe2007bayes}).
 
@@
     \item[\textbf{b.}]  The relative weight $w_k$ (or contribution, or mixing 
-                        proportion) of the amount(s) of data in each group.
+                        proportion) of the amount(s) of data in each group.\footnote{In practice only $K - 1$ weights must be encoded because $\sum\weight_k = 1$.}
 
@@
 %minimising (over cluster indices $k$) $ -\log (p_k(x)) $ (which is the cost of encoding the particular cluster, $k$) plus the negative log-likelihood given cluster $k$'s parameters.
-%However, as per \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}\cite[sec. 6.8]{Wallace05}, the cost can be reduced to
+%However, as per \cite{WallaceDowe1994b,WallaceDowe1997a,WallaceDowe2000}\cite[sec. 6.8]{Wallace:2005}, the cost can be reduced to
 %$ -\log ( \Sigma_{k=1}^{K} p_k(x) ~ \frac{1}{\sqrt{2 \pi} \sigma_k} ~ e^{-\frac{(x-\mu_k)^2}{2\sigma_k^2}}) $
@@
 \end{eqnarray}
-\noindent{}$\kappa(Q)$ is a function to approximate the lattice quantisation
-constant for $Q$ free parameters\cite{ConwaySloane1984}, and $|\fisher{\vec{\theta}}|$
+\noindent{}and $\kappa(Q)$ is a function to approximate the lattice quantisation
+constant for $Q$ free parameters\cite{Conway|Sloane:1984}, and $|\fisher{\vec{\theta}}|$
 is the determinant of the \emph{expected} Fisher information matrix (the second
 order partial derivatives of the negative log-likelihood function $-\log\likelihood\left(\vec{\data}|\vec{\theta}\right)$).
-A common that arises after listing the components above is that although we
-have expliticly encoded the model parameters, we have not encoded `the model'
+A common issue that arises after listing the components above is that although we
+have explicitly encoded the model parameters, we have not encoded `the model'
 in the sense that we have not described how these encoded parameters relate 
@@
 
-\todo{Derivations and updates to the mixture parameters through E-M}
-
+Given a mixture model with $K$ components, we iteratively update the model parameters $\vec\theta$
+by expectation-maximization \cite{dempster1977maximum}. In the expectation step we calculate the partial memberships of the  data
+to each component, as specified by  the responsibility matrix
+\begin{eqnarray}
+	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
+\end{eqnarray}
+
+\noindent{}where $\mathcal{R}_{nk}$ represents the fractional assignment of the $n$th data point to the
+$k$th component.  At the maximization step we update our estimates of the mixture parameters to minimize
+the total message length (Equation \ref{eq:objective_function}). The updated estimates for the relative
+weights are obtained by differentiating Equation \ref{eq:objective_function} with respect to $\weight_k$ under the constraint $\sum\weight_k = 1$ (see \cite{Kasarapu|Allison:2015} for derivation) such that:
+\begin{equation}
+	\weight_k^{(t+1)} = \frac{2n_k^{(t)} + 1}{2N + K}
+\end{equation}
+\noindent{}where $n_k$ is the effective membership for the $k$th component
+\begin{equation}
+	n_k = \sum_{i=1}^{N} \mathcal{R}_{nk} \quad .
+\end{equation}
+
+\noindent{}The updates of the mean and covariance matrix are given by:
+\begin{eqnarray}
+\vecmean_k^{(t+1)} &=& \frac{1}{n_k^{(t)}}\sum_{i=1}^{N}\mathcal{R}_{nk}^{(t)}\vecmean_k \\
+\veccov_k^{(t+1)} &=& \frac{1}{n_k^{(t)} - 1}\sum_{i=1}^{N}\mathcal{R}_{nk}^{(t)}\left(\vec\data_i - \vecmean_k^{(t+1)}\right)\left(\vec\data_i - \vecmean_k^{(t+1)}\right)\transpose \quad .
+\end{eqnarray}
+
+Throughout this work we consider the mixture to be converged when the change in message length
+is less than $10^{-5}$ nats.
+
+\subsection{Predicting the message length of future mixtures}
+
+The objective function in Equation \ref{eq:objective_function} is valid for a class of 
+Gaussian mixture models,  and the EM updates that act to minimize Equation \ref{eq:objective_function} are valid for a mixture with a given number of $K$ components.
+However, we require a search strategy to minimise Equation \ref{eq:objective_function} across all possible numbers of components.
+Existing search strategies will greedily  seek to trial different numbers of components.
+Here we describe methods to predict the minimum message length of a future (untrialled) mixture, which help to guide our search strategies (Section \ref{sec:search}).
+We separately describe the predictions for different items in Equation \ref{eq:objective_function}:
+\begin{enumerate}
+	\item predicting the sum of the log of the weights;
+	\item predicting the negative log-likelihood; and
+	\item predicting the sum of the log of the determinant of the covariance matrices.
+\end{enumerate}
+
+These items directly contribute to Equation \ref{eq:objective_function}, and their values are not known before trialling a mixture.
+All other contributions to Equation \ref{eq:objective_function} can be calculated directly given $K$, $N$, and $D$.
+
+\subsubsection{Predicting $\sum\log\weight_k$}
+
+We consider the two extremes of $\sum\log\weight_k$ for a mixture of $K$ components:
+\begin{enumerate}
+    \item The weights are distributed uniformly such that $w_k = 1/K$ and
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
+          \end{equation}
+    \item The weights are distributed such that $K - 1$ components each 
+          describe just two data points $w_{2..K} = 2/N$, and the remaining 
+          component describes all other data points $w_{1} = 1 - 2(K-1)/N$
+          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
+                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
+		    \label{eq:slw_uniform}
+          \end{equation}
+
+          \noindent{}and by making use of logarithmic identities:
+          %  $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
+          \begin{equation}
+                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad . \label{eq:slw_nonuniform}
+          \end{equation}
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
+          %\end{equation}
+\end{enumerate}
+
+\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
+derived the bounds
+%\begin{eqnarray}
+%\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
+%\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
+%\end{eqnarray}
+\begin{equation}
+-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
+\end{equation}
+\noindent{}which  describe the smallest and largest contribution that the weights can contribute to the message length for any untrialled mixture with $K$ components.
+When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
+that describes the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$).
+We use estimates of $f_u$ from previously trialled mixtures to make predictions for future, untrialled mixtures.
+
+\subsubsection{Predicting $-\log{\mathcal{L}(\data|\vectheta)}$}
+
+\subsubsection{Predicting $\sum\log|\veccov_k|$}
+
+\subsubsection{Combining the predictions}
+
+% TODO: Predicting the negative log-likelihood.
+
+% TODO: Predicting the sum of the log of the determinant of the covariance matrices.
+
+% Combining it together.
+
+\begin{comment}
 \subsection{Theoretical lower bounds on the message length of future mixtures}
@@
 
-\subsection{Predicting the message length of future mixtures}
-
-The objective function in Equation \ref{eq:objective_function} remains valid across a class of 
-Gaussian mixture models where $N > K$. However, we still require a strategy to minimise this
-objective function over all possible mixture parameters $\vec{\psi}$. Here we describe methods to predict
-the \emph{minimum} message length of future (untrialled) mixtures, which we will then use in conjunction 
-with search strategies (Section \ref{sec:search}).
-
-% TODO: Predicting the sum of the log of the weights.
-
-% TODO: Predicting the negative log-likelihood.
-
-% TODO: Predicting the sum of the log of the determinant of the covariance matrices.
-
-% Combining it together.
+\end{comment}
 
@@
 
-We use the search algorithm proposed \citep{kasarapu2015minimum} as a benchmark for comparison.
+We use the search algorithm proposed \citep{Kasarapu|Allison:2015} as a benchmark for comparison.
 The method is presented in Algorithm \ref{alg:kasarapu}.
@@
 \begin{algorithm}[H]
-\caption{The benchmark method~\cite{kasarapu2015minimum}.}
+\caption{The benchmark method~\cite{Kasarapu|Allison:2015}.}
 \label{alg:kasarapu}
@@
 We evaluate the message length of each component by rearranging Equation \ref{eq:objective_function} and multiplying the negative log-likelihood
-$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ that has entries
-\begin{eqnarray}
-	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
-\end{eqnarray}
-\noindent{}such that the message length of the $k$th component is
+$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ 
+%\begin{eqnarray}
+%	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
+%\end{eqnarray}
+such that the message length of the $k$th component is
 \begin{eqnarray}
@@
 set the means of the child mixtures as positive and negative eigenvalues along the principal
-eigenvector of the component covariance matrix. A matrix of child responsibilities is then
+eigenvector of the component covariance matrix \citep{Kasarapu|Allison:2015}.
+A matrix of child responsibilities is then
 calculated based on the Euclidean distance between each datum and the child means (i.e., hard
@@
 all other components fixed, and only updating the estimates of the parameters of the child
-components \citep{kasarapu2015minimum}.
+components \citep{Kasarapu|Allison:2015}.
 After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
@@
 We assume a prior on the number of components $K$ of $\prior{K} \propto 2^{-K}$
-\cite[][p279, sec. 6.8.2]{Wallace05} such that the length of the 
+\cite[][p279, sec. 6.8.2]{Wallace:2005} such that the length of the 
 optimal lossless message to encode $K$ is 
@@
 \noindent{}where $\kappa_K$ is a lattice constant that varies between $1/12$
-and $1/(2\pi e)$ \cite{BoultonWallace1969}.
+and $1/(2\pi e)$ \cite{Boulton|Wallace:1969}.
 
@@
 We adopt a conjugate inverted Wishart prior for the covariance matrix of an
-individual mixture \cite[Section 5.2.3 of][]{Schafer_1997}:
+individual mixture \cite[Section 5.2.3 of][]{Knorr-Held:2000}:
 
@@
 as the product of $\detfisher{\vecmean_k}$ and $\detfisher{\veccov_k}$ 
-\cite{Oliver_1996,Roberts_1998}. In order to derive $\detfisher{\vecmean_k}$
+\cite{oliver1996unsupervised,roberts1998bayesian}. In order to derive $\detfisher{\vecmean_k}$
 we take the second derivative of the log-likelihood function 
@@
 We make use of an analytical expression for $\detfisher{\veccov_k}$
-\cite{Dwyer_1967,Magnus_1988}:
+\cite{Dwyer:1967,Magnus|Neudecker:1988}:
 \begin{equation}
@@
 %[]
-{Wallace_2005} such that,
+{Wallace:2005} such that,
 \begin{equation}

### 2018-10-17 article/ms.tex
add cost figure
werds

@@
 \item Minimum Message Length (MML) formulation
-\item 2 search methods
+\item $\mathcal{O}\left(K\right)$ search methods for gaussian mixture  modelling
 \end{itemize}
 \section{Galactic Archaeology}
-% "Galaxy produces ‘molecular forests’"
-% https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
 Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
@@
 \subsection{The Message-Breaking Method} \label{sec:mbm}
-The simplest strategy we introduce here, \emph{The Message-Breaking Method}, does not use any predictions of the message length of future mixtures. Instead it starts from 
+The simplest strategy we introduce here, \emph{The Message-Breaking Method}, 
+is a direct improvement upon the Benchmark Method that does not use predictions of the message length of future mixtures. Instead the algorithm begins with 
 a $K=1$ mixture and iteratively splits the component with the largest message length until 
 the total message length no longer improves (Algorithm~\ref{alg:MBM}).
-This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split components (instead of splitting, merging, and deleting all components).
-Similarly, in \emph{The Message-Breaking Method} we only split the component with the largest message length instead of exhaustively trying to split all components at every mixture.
+This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split the component with the largest message length instead of splitting, merging, and deleting all components (and running EM at each operation).
 
@@
 expectation step of the expectation-maximisation algorithm, adding negligible computational overhead to the algorithm.
-
-When splitting the component with the largest message length, we split this component into two child components. Initially we 
-set the means of the child mixtures as positive and negative eigenvalues along the principal
-eigenvector of the component covariance matrix \citep{Kasarapu|Allison:2015}.
-A matrix of child responsibilities is then
-calculated based on the Euclidean distance between each datum and the child means (i.e., hard
-association), which we use to estimate the covariance matrices and weights of the child
-components. After the parameters of the child components are initialised, we run \emph{partial}
-expectation-maximization to iteratively update the parameter estimates of the child components. 
-By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
-all other components fixed, and only updating the estimates of the parameters of the child
-components \citep{Kasarapu|Allison:2015}.
-After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
-
-\andy{Should we move/remove the following paragraph?}
-One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
-One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$.
-As a result, the algorithm would converge to a local minima, not the global one.
-However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.
 
@@
 
-\subsection{Component-Jumping Method}
-
-The \emph{Component-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.
+When splitting the component with the largest message length, we split this component into two child components. Initially we 
+set the means of the child mixtures as positive and negative eigenvalues along the principal
+eigenvector of the component covariance matrix \citep{Kasarapu|Allison:2015}.
+A matrix of child responsibilities is then
+calculated based on the Euclidean distance between each datum and the child means (i.e., hard
+association), which we use to estimate the covariance matrices and weights of the child
+components. After the parameters of the child components are initialised, we run \emph{partial}
+expectation-maximization to iteratively update the parameter estimates of the child components. 
+By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
+all other components fixed, and only updating the estimates of the parameters of the child
+components \citep{Kasarapu|Allison:2015}.
+After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
+
+One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
+One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$.
+As a result, the algorithm would converge to a local minima, not the global one.
+In the code implementation that we release with this work, we have implemented a \emph{Strict Message-Breaking Method} variant that allows the user to specify the number of successive $K$ trials that increased the message length, before adopting the mixture with the shortest message length.
+By default the Strict Message-Breaking Method requires five successive mixtures where the message length did not improve.
+
+\subsection{Message-Jumping Method}
+
+The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.
 
@@
 To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets.
-As the output of the algorithms depends on the starting seed, and some of the methods can be affected by randomness, we perform 50 runs for each search scheme and dataset.
-
-\andy{The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms.
+
+\subsection{Generated Datasets}
+
+We set the data dimensionality $D = 2$ after previous experiments found that the data dimensionality had only a small effect on the differential cost on the search strategies tested.
+We chose $N$ to vary uniformly in log space from $N = 10$ data points to $N = 10^{4.5}$.
+For each $N$ we set $K = N/10$, and then used these values of $N$, $D$, and $K$ to generate a synthetic dataset.
+The weights for each mixture were drawn from a Dirichlet distribution with concentration $\alpha = 100$ while ensuring that $\weight_k \in \left(\frac{2}{N}, 1-\frac{2(K-1)}{N}\right)$.
+The means for each component were drawn from $\mathcal{U}\left(-10, +10\right)$.
+The correlation coefficients  for the component covariance matrices were drawn from
+\begin{equation}
+	\rho \sim \frac{\left[1-\mathcal{U}(-1, +1)^2\right]^{\frac{1}{2}(s - 4)}}{\mathcal{B}\left(\frac{1}{2}, \frac{1}{2}(s - 4)\right)}
+\end{equation}
+\noindent{}where $s$ is an isotropy hyperparameter that describes the probability of correlations between data dimensions, which we fix as $s=100$.
+The diagonal variances for the component covariance matrices were drawn from $|\mathcal{N}\left(0, 0.1\right)|$.
+Given these randomly drawn parameters we generated $N\weight_k$ members for each component, rounding as necessary to ensure the total number of generated data points is kept at $N$.
+
+We evaluated the following methods on all generated datasets: the benchmark method (Algorithm \ref{alg:kasarapu}), the Message-Breaking Method (Algorithm \ref{alg:MBM}), and \todo{X}.
+The cost (in seconds) is shown as a function of the number of data points $N$ and true number of clusters $K$ in Figure \ref{fig:algorithm_costs}.
+\todo{Cost is approximately linear for MBM, but of O(3) for kasarpu.
+Not surprising because Kasarapu is a greedy search method, requiring at least 3*K operations (split, merge, delete) at each mixture.}
+
+We considered the search algorithm to be converged to the global minima if the calculated message length is shorter than the message length given the true mixture parameters.
+\todo{Something about when converged and not}
+
+\todo{Figure: predictions for future mixtures for one generated data set}
+
+\andy{As the output of the algorithms depends on the starting seed, and some of the methods can be affected by randomness, we perform 50 runs for each search scheme and dataset.
+The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms.
 The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.}
 
-\subsection{Generated Datasets}
-
-\todo{How we generated data sets}
-
-\todo{Range of generated data sets}
-
-\todo{Figure: predictions for future mixtures}
-
-\todo{Figure: computational cost of various methods}
+\subsection{Galactic Archaeology data}
+
+\todo{describe APOGEE data}
 
@@
 all model parameters (including the number of mixtures, $K$):
-
-\begin{equation}
-\todo{I = 0}
-\end{equation}
+\begin{eqnarray}
+	I(\vec{\data}|\vec{\psi}, N, D) &=& -\log\mathcal{L}\left(\vec{\data}|\vectheta\right) - \frac{(D + 2)}{2}\sum_{k=1}^{K}\log{|{\cov}_k|} \nonumber \\
+	&& \cdots\ + \left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum_{k=1}^{K}\log{\weight_k} + K\left(1-\frac{D}{2}\right)\log{2} \nonumber \\ && \cdots\ + \log\Gamma{(K)} + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N} \,\,\,\, .
+	\label{eq:objective_function}
+\end{eqnarray}
 
 \section{Bounds on components of the message length}
-
-In Eq. \ref{eq:I1} it can be seen that for any finite mixture, $I_1$ can be
-almost directly computed just from knowing the number of mixture components
-$K$ and the properties of the data: the number of data points $N$ and its
-dimensionality $D$. The only ambiguous term is $\sum_{k=1}^{K}\log\weight_k$,
-which will depend on the data. However, for any finite mixture of $K$
-components we can consider the two extremes of $\sum_{k=1}^{K}\log\weight_k$:
-
-\begin{enumerate}
-    \item The weights are distributed uniformly such that $w_k = 1/K$ and
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
-          \end{equation}
-    \item The weights are distributed such that $K - 1$ components each 
-          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
-          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
-          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
-          %\begin{equation}
-          %      \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
-          %\end{equation}
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
-                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
-          \end{equation}
-
-          \noindent{}and by making use of the identity 
-            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
-          \end{equation}
-          %\begin{equation}
-          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
-          %\end{equation}
-\end{enumerate}
-
-\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
-derived the bounds
-\begin{eqnarray}
-\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
-\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
-\end{eqnarray}
-\begin{equation}
--K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
-\end{equation}
-
-\noindent{}which gives us analytic expressions for the upper and lower value 
-of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
-bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.
 

### 2018-10-17 article/ms.tex
words

@@
 \usepackage{algpseudocode}
+\usepackage{mathtools}
+
+\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
+\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
 
@@
           %  $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
-          \begin{equation}
-                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad . \label{eq:slw_nonuniform}
-          \end{equation}
+          %\begin{equation}
+          %      \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + %\log{N} + \log{\left(N - 2K + 2\right)} \quad . \label{eq:slw_nonuniform}
+          %\end{equation}
+          \begin{equation}
+            \sum_{k=1}^{K}\log\weight_k = (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
+          \end{equation}
+
           %\begin{equation}
@@
 \begin{equation}
--K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
+-K\log{K} \quad \geq \quad \sum_{k=1}^{K}\log\weight_k \quad \geq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
 \end{equation}
@@
 When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
-that describes the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$).
-We use estimates of $f_u$ from previously trialled mixtures to make predictions for future, untrialled mixtures.
+\begin{equation}
+f_u = \frac{\alpha - \sum\log\weight_k}{\alpha + K\log{K}}
+\end{equation}
+\noindent{}where for brevity $\alpha = (K-1)\log{2} + \log(N-2K+2)-K\log{N}$,
+and $f_u$ the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$).
+We use estimates of $f_u$ from previously trialled mixtures to make predictions for the uniformity of future,
+untrialled mixtures, and the message length required to encode the weights.
+Given the theoretical bounds in Equations \ref{eq:slw_uniform} and \ref{ew:slw_nonuniform},
+we clip the predictions of the expectation value on the message length required to encode the weights by these bounds.
 
@@
 
-The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.
-
-\todo{initialisation, how kmeans++ breaks when too big unless covariance regularisation\\}
-
-\todo{Run E-M, make predictions\\}
-
-\todo{Advantage in being able to jump many $\Delta{}K$ at once}
+The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of computations.
+In the \emph{Message-Breaking Method} we
+try to accomplish this using partial E-M to minimise the total number of computations
+required to converge on a $K+1$ component mixture from a $K$ component mixture.
+However, in the \emph{Message-Jumping Method} we seek to minimise the total number
+of computations by optimizing a fewer number of mixtures. Instead of iteratively
+perturbing a mixture from say $K = 10$ to $K = 30$, we use predictions of the
+message length of untrialled mixtures and immediately `jump' to a component (
+e.g., $K = 27$) that will provide the most improvement in our predictions of the
+minimum message length over \emph{all} components. In doing so we seek a global 
+message length minima with a fewer number of total E-M operations.
+
+We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:mbm}. In this
+method we seek to predict the minimum message length for any number of $K$ components,
+and then converge to the optimal mixture as quickly as possible. For this reason,
+we initialise many mixtures that are evenly separated in $\log_{10}$ space from
+$K = 1$ to $K = N/2$ in order to
+quickly describe the minimum message length across all $K$. Specifically we
+introduce a search hyperparameter $K_{inits}$ (defaults to 10) that sets the
+number of mixtures to initialise -- and those initial $K$ values. 
+We initialise each mixture using the K-means++ algorithm\cite{kmeanspp} and
+optimize the mixture using EM as described earlier. We trial each of the $K_{init}$
+mixtures in rank order of increasing $K$. If one of these mixtures fails to initialise
+or optimise, this is likely because we could not perform Choleksy decomposition
+of one (or more) of the covariance matrices. In practice this is typically because
+the estimate of $K$ is much higher than $K_{true}$, leading to near-zero values
+along the diagonal of the covariance matrix. In this situation we do not attempt
+to initialise any more of the $K_{init}$ mixtures -- although higher $K$ values
+can be trialled later -- and we being the search method.
+
+The motivation for this initialisation process is to mimic the search strategy
+that a human might take: What is an order-of-magnitude estimate on the number of
+$K$? Or what is a ballpark number on the maximum number of $K$ that we should
+trial? After we have that number, and an estimate on the message length for a
+few sparsely-sampled numbers of components, we can make reasonable predictions
+on the global minimum.
+
+We record the message length for the optimal mixture at each trialled $K$, and
+the model parameters for that mixture. With these quantities we predict the
+message length of untrialled mixtures for every $K$ value from $K = 1$ to $K = \ceil{1.1\max(K_{trialled})}$ -- e.g., up to 10\% ahead of the largest previously trialled
+$K$ value. The message length predicted from the optimised Gaussian processes
+provide an expectation value on the message length for an untrialled mixture, as
+well as a variance in that prediction (Figure \ref{fig:search-predictions}).
+Before selecting the next value of $K$ to trial, we use the expectation value and
+associated variance to compute the acquisition function (or the "expected
+improvement") at each value of $K$ \cite{Jones:1998}.
+
+We select the next $K$ value by choosing the value that will maximise the expected
+improvement \todo{but this forces us out to large K, so ....}
+
+\todo{initialisation with kmeans++, if it fails,....}
 
@@
 
-To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets.
+We conducted a set of experiments and real datasets to evaluate the proposed
+search algorithms.
 
@@
 
-\todo{describe APOGEE data}
+We use detailed chemical abundances from the Apache Point Observatory Galactic Evolution Experiment (APOGEE), an experiment run by the Sloan Digital Sky Survey (SDSS) collaboration.
+We make use of SDSS Data Release 12, which includes high-resolution, high S/N infrared spectra for $\sim$163,000 stars collected by APOGEE.
+Each observation
+is an array of 8,575 pixels that have an associated rest-frame wavelength, and a
+measurement of the stellar flux that is normalised relative to the spectral energy
+distribution such that a flux value of 1 indicates no stellar absorption and a
+flux value near 0 implies strong stellar absorption. All else being equal, a stronger
+amount of absorption (lower normalised flux value) at a particular wavelength 
+indicates that the star contains more of the chemical element that absorbs and
+emits light at that wavelength. For example, sodium emits and absorbs light 
+at X\,\AA\ when an electron moves between the XY transitions. All else being
+equal, a star with lower normalised flux at a rest-frame wavelength of X\,\AA\
+indicates that the star has a higher abundance of sodium.
+
+We make use of the detailed chemical abundances calculated by \citep{casey}
+from 96,648 APOGEE spectra. This data set includes 15 measured chemical abundances
+for all stars (e.g., no missing data), with estimated uncertainties. All chemical
+abundances are reported as the logarithmic count of the number of atoms relative
+to hydrogen, with a correction to account for the Sun's chemical composition such
+that "[Na/H] = 0" indicates a star has the same abundance of sodium relative to
+hydrogen as the Sun. \todo{To provide context on the fidelity of the data, in Figure \ref{fig:gce} we show the measured abundance
+of [Na/H] with respect to the measured abundance of [Fe/H] for all XX,XXX stars,
+where the Sun's abundance is marked.}
+
+The data set also contains other measured properties of the
+stars, including their temperature and surface gravity, and whether they are 
+likely associated with any gravitationally-bound star cluster. Most stars are
+not likely members of any known star cluster; most stars in this dataset are 
+described as `field' stars because they have no obvious association.
+We do not make use of any of this ancillary data (e.g., temperature, cluster
+association) during our search, but we do make use of it to evaluate the performance
+of our search.
+
+\section{Results}
+
+Costs of generated data sets.
+Convergence of generated data sets.
+When ther eis a switchover and we are faster.
+
+Expectation that kasarapu and allison should be O(3) and we are O(1) with some overhead.
+
+\begin{figure}
+    \includegraphics[width=1.0\textwidth]{cost.pdf}
+    \caption{Elapsed wall clock time for trialled search strategies to converge 
+             on generated datasets as a function of the true number of clusters
+             in the data. The approximate cost functions are shown.}
+    \label{fig:cost}
+\end{figure}
 
@@
 \todo{Figures showing the concentration of chemically-identified data}
-
-\paragraph{Simulations}
-
-\paragraph{Criteria for Comparison}
-
-\subsection{Galactic Archaeology data}
-
-\section{Results}
-
-\todo{Convergence criteria: how often do we reach the `true' solution?}
-
-\todo{What is the variance in the results of our method?}
-
-\todo{Speed comparisons}
 
@@
 \label{app:mml_derivation}
-In the following subsections we provide expressions for the message length of
-each component listed above, and our prior beliefs, before describing our
+In the following subsections we provide expressions for the message length to
+encode a gaussian mixture model, and our prior beliefs on the parameters of that model, before describing our
 full objective function.

### 2018-10-17 scripts/mpl_utils.py
plot script

@@ -1,0 +1,92 @@
+
+# A matplotlib style based on the gala package by @adrn:
+# github.com/adrn/gala
+
+import numpy as np
+
+mpl_style = {
+
+    # Lines
+    'lines.linewidth': 1.7,
+    'lines.antialiased': True,
+    'lines.marker': '.',
+    'lines.markersize': 5.,
+
+    # Patches
+    'patch.linewidth': 1.0,
+    'patch.facecolor': '#348ABD',
+    'patch.edgecolor': '#CCCCCC',
+    'patch.antialiased': True,
+
+    # images
+    'image.origin': 'upper',
+
+    # colormap
+    'image.cmap': 'viridis',
+
+    # Font
+    'font.size': 12.0,
+    'text.usetex': True,
+    'text.latex.preamble': r'\usepackage{amsmath}',
+    'text.latex.preview': True,
+    'axes.unicode_minus': False,
+
+    # Axes
+    'axes.facecolor': '#FFFFFF',
+    'axes.edgecolor': '#333333',
+    'axes.linewidth': 1.0,
+    'axes.grid': False,
+    'axes.titlesize': 'x-large',
+    'axes.labelsize': 'large',
+    'axes.labelcolor': 'k',
+    'axes.axisbelow': True,
+
+    # Ticks
+    'xtick.major.size': 8,
+    'xtick.minor.size': 4,
+    'xtick.major.pad': 6,
+    'xtick.minor.pad': 6,
+    'xtick.color': '#333333',
+    'xtick.direction': 'in',
+    'ytick.major.size': 8,
+    'ytick.minor.size': 4,
+    'ytick.major.pad': 6,
+    'ytick.minor.pad': 6,
+    'ytick.color': '#333333',
+    'ytick.direction': 'in',
+    'xtick.labelsize': 'medium',
+    'ytick.labelsize': 'medium',
+
+    # Legend
+    'legend.fancybox': True,
+    'legend.loc': 'best',
+
+    # Figure
+    'figure.figsize': [6, 6],
+    'figure.facecolor': '1.0',
+    'figure.edgecolor': '0.50',
+    'figure.subplot.hspace': 0.5,
+
+    # Other
+    'savefig.dpi': 300,
+}
+
+
+
+
+def plot_histogram_steps(ax, x_bins, y, y_err, **kwargs):
+
+    xx = np.array(x_bins).repeat(2)[1:]
[…]

### 2018-10-17 scripts/plot_evaluations_2d.py

@@ -6,12 +6,17 @@ Plot the results from the evaluations on artificial data.
 # TODO: Get these from somewhere else?
 
 import numpy as np
+import matplotlib
 import matplotlib.pyplot as plt
 import scipy.optimize as op
 import pickle
 from collections import OrderedDict
 
 from glob import glob
+
+from mpl_utils import mpl_style
+
+matplotlib.style.use(mpl_style)
 
 search_strategies = OrderedDict([
     ("BayesStepper", dict()),
@@ -56,11 +61,20 @@ for search_strategy in search_strategies.keys():
 
 # Plot as a function of N, K
 
-x_labels = ["N", "K"]
+x_labels = ["K"]
 y_label = "time"
 
+latex_labels = dict(time=r"$\textrm{time}\,/\,\textrm{seconds}$",
+                    K=r"$\textrm{number of true clusters}$ $K$",
+                    KasarapuAllison2015=r"$\textrm{Kasarapu \& Allison (2015)}$",
+                    BayesStepper=r"$\textrm{Message-Breaking Method}$")
+
+scat_kwds = dict(KasarapuAllison2015=dict(marker="s", s=50),
+                 BayesStepper=dict(s=50))
+
 L = len(x_labels)
-fig, axes = plt.subplots(1, L, figsize=(4 * L, 4))
+fig, axes = plt.subplots(1, L, figsize=(5 * L, 5))
+axes = np.atleast_1d([axes]).flatten()
 
 
 max_y = 0
@@ -103,11 +117,26 @@ for i, (ax, x_label) in enumerate(zip(axes, x_labels)):
         xi = np.array([x.min(), x.max()])
         yi = 10**np.polyval(mu[::-1], np.log10(xi))
 
-        ax.plot(xi, yi, label=f"$\mathcal{{O}}({x_label}^{{{mu[1]:.1f}}})$")
+        ax.plot(xi, yi, "-", label=f"$\mathcal{{O}}({x_label}^{{{mu[1]:.1f}}})$")
+        _kwds = dict(label=latex_labels.get(search_strategy, search_strategy))
+        _kwds.update(scat_kwds.get(search_strategy, dict()))
 
-        scat = ax.scatter(x[converged], y[converged], label=search_strategy)
+        scat = ax.scatter(x[converged], y[converged], **_kwds)
 
[…]

### 2018-10-18 gmmmml/gmm.py
bayesjumper method

@@ -271,7 +271,7 @@ class GaussianMixture(object):
         return em.expectation_maximization(y, means, covs, weights, **kwds)
 
 
-    def search(self, y, search_strategy="BayesStepper", **kwargs):
+    def search(self, y, search_strategy="MessageBreaking", **kwargs):
         r"""
         Search for the optimal number of multivariate Gaussian components, and
         the parameters of that mixture, given the data.
@@ -297,11 +297,12 @@ class GaussianMixture(object):
             Some strategies include:
 
             - `strategies.KasarapuAllison2015`
-            - `strategies.BayesStepper`
-            - `strategies.BayesJumper`
+            - `strategies.MessageBreaking`
+            - `strategies.StrictMessageBreaking`
+            - `strategies.MessageJumping`
             - `strategies.GreedyKMeans`
 
-            Defaults to `strategies.BayesStepper`, a fast and possibly convex
+            Defaults to `strategies.MessageBreaking`, a fast and possibly convex
             optimisation search strategy.
         """
 

### 2018-10-18 gmmmml/policies/base.py

@@ -1,2 +1,6 @@
+
 class Policy(object):
-    pass
+
+    def __init__(self, *args, **kwargs):
+        self.meta = dict()
+        

### 2018-10-18 gmmmml/policies/convergence.py

@@ -60,6 +60,15 @@ class StrictConvergencePolicy(DefaultConvergencePolicy):
     num_consecutively_worse_trials = 5
 
 
+class NeverConvergePolicy(BaseConvergencePolicy):
+
+    @property
+    def converged(self):
+        """ Continue indefinitely because we assume convergence is over-ridden
+        by another policy. """
+
+        return False
+
 
 
 class ConvergedWithSuccessivelyWorseIterations(BaseConvergencePolicy):

### 2018-10-18 gmmmml/policies/initialisation.py

@@ -17,8 +17,10 @@ class BaseInitialisationPolicy(Policy):
 
     def __init__(self, model, K_init=1, **kwargs):
 
+        super(BaseInitialisationPolicy, self).__init__(model, **kwargs)
+
         self.model = model
-        self.K_init = K_init
+        self.meta.update(K_init=K_init)
         return None
 
 
@@ -34,7 +36,7 @@ class DefaultInitialisationPolicy(BaseInitialisationPolicy):
         y = np.atleast_2d(y)
         N, D = y.shape
 
-        K_inits = np.logspace(0, np.log10(N/2.0), self.K_init, dtype=int)
+        K_inits = np.logspace(0, np.log10(N/2.0), self.meta["K_init"], dtype=int)
 
         for i, K in enumerate(K_inits):
 
@@ -50,12 +52,3 @@ class DefaultInitialisationPolicy(BaseInitialisationPolicy):
 
 
 
-
-
-
-
-class LogarithmicInitialisationPolicy(DefaultInitialisationPolicy):
-
-    def __init__(self, model, K_init=10, **kwargs):
-        return super(DefaultInitialisationPolicy, self).__init__(model, K_init, 
-                                                                 **kwargs)

### 2018-10-18 gmmmml/policies/movement.py

@@ -1,6 +1,8 @@
 
 import logging
 import numpy as np
+from scipy.special import erf
+from scipy.signal import find_peaks_cwt
 from .base import Policy
 
 logger_name, *_ = __name__.split(".")
@@ -57,6 +59,75 @@ class JumpToMMLMixtureMovementPolicy(BaseMovementPolicy):
 
         return None
 
+
+class ConvergeToMixtureWithMaximumExpectationImprovementPolicy(BaseMovementPolicy):
+
+    def move(self, y, **kwargs):
+
+        width = 2
+
+        old_min = None
+
+        while True:
+
+            # Check if converged.
+
+            Kp, I, I_var, I_lower = self.predict(y)
+
+            # The implicit assumption is that the bounds of Kp are not changing
+            # between predictions, otherwise the policy will *always* want to
+            # expand the bounds of K (rather than finding K minimum).
+
+            # Calculate the acquisition function.
+            I_min = np.min(self.model._state_I)
+            I_diff = I_min - I
+            chi = I_diff / np.sqrt(I_var)
+            Phi = 0.5 * (1.0 + erf(chi / np.sqrt(2)))
+            phi = np.exp(-0.5 * chi**2) / np.sqrt(2 * np.pi * I_var)
+            A_ei = I_diff * Phi + I_var * phi
+
+            # Prefer the point that maximizes the expectation improvement.
+            indices = [np.argmax(A_ei)]
+            indices.extend(find_peaks_cwt(A_ei, width * np.ones_like(A_ei)))
+            indices.extend(np.argsort(A_ei)[::-1])
+            indices = np.array(indices)
+
+            # Unique without sorting.
+            sidx = np.unique(indices, return_index=True)[1]
+            indices = np.array([indices[si] for si in sorted(sidx)])
+
+            # Next values to try, excluding trialled.
+            K_next = [Kp[i] for i in indices \
+                if  Kp[i] not in self.model._results.keys() \
+                and Kp[i] not in self.model._state_K]
+
+            # OK, take best.
+            if len(K_next) == 0:
+                logger.warn("No new movement to make. May have not converged")
+                break
+
+            K = K_next[0]
+
+            # Check for convergence.
+            current_min = Kp[np.argmin(I)]
+
+            # We could actually claim convergence before actually trialling the
+            # K_min mixture, but we will need the best mixture returned, so we
+            # effectively *have* to trial it.
+            # However, I don't want to make the search heuristic more complex,
[…]

### 2018-10-18 gmmmml/policies/prediction.py

@@ -9,6 +9,10 @@ logger = logging.getLogger(logger_name)
 
 
 class BasePredictionPolicy(Policy):
+    
+    def __init__(self, *args, **kwargs):
+        super(BasePredictionPolicy, self).__init__(*args, **kwargs)
+    
 
     def predict(self, y, **kwargs):
         raise NotImplementedError("should be implemented by the sub-classes")
@@ -22,10 +26,53 @@ class DefaultPredictionPolicy(BasePredictionPolicy):
 
         # Predict a little bit ahead.
         Kp = 1 + np.arange(2 * np.max(self.model._state_K))
-        return self.model._predict_message_length(Kp, N, D, **kwargs)
+
+        logger.info("Predicting between K = {0} and K = {1}".format(Kp[0], Kp[-1]))
+        
+        K, I, I_var, I_lower = self.model._predict_message_length(Kp, N, D, **kwargs)
+
+        K_min = K[np.argmin(I)]
+        logger.info(f"Predicted minimum message length at K = {K_min}")
+
+        return (K, I, I_var, I_lower)
+
+
+
+class LookaheadFromInitialisationPredictionPolicy(BasePredictionPolicy):
+
+    def __init__(self, *args, **kwargs):
+        super(LookaheadFromInitialisationPredictionPolicy, self).__init__(*args, **kwargs)
+
+
+    def predict(self, y, **kwargs):
+        """
+        Predict the message length only up to the K value that was trialled
+        during the initialisation procedure.
+        """
+
+        N, D = y.shape
+        """
+        K_inits = np.logspace(0, np.log10(N/2.0), self.meta["K_init"], dtype=int)
+        K_max = K_inits[1 + self.model._num_initialisations]
+        """
+        K_max = int(np.ceil(1.5 * [*self.model._results][self.model._num_initialisations - 1]))
+
+
+        Kp = np.arange(1, 1 + K_max).astype(int)
+        logger.info("Predicting between K = {0} and K = {1}".format(Kp[0], Kp[-1]))
+        
+        K, I, I_var, I_lower = self.model._predict_message_length(Kp, N, D, **kwargs)
+
+        K_min = K[np.argmin(I)]
+        logger.info(f"Predicted minimum message length at K = {K_min}")
+
+        return (K, I, I_var, I_lower)
+            
 
 
 class NoPredictionPolicy(BasePredictionPolicy):
+
+
 
     def predict(self, y, **kwargs):
 

### 2018-10-18 gmmmml/strategies.py

@@ -3,31 +3,37 @@
 from .policies import (initialisation, prediction, movement, repartition, 
                        convergence)
 
-__all__ = ["BayesStepper", "BayesJumper", "GreedyKMeans", "KasarapuAllison2015"]
+__all__ = ["MessageBreaking", "MessageJumping", "GreedyKMeans", "KasarapuAllison2015"]
 
 
-class BayesStepper(initialisation.DefaultInitialisationPolicy,
-                   prediction.DefaultPredictionPolicy,
-                   movement.StepTowardsMMLMixtureMovementPolicy,
-                   repartition.IterativelyRepartitionFromNearestMixturePolicy,
-                   convergence.DefaultConvergencePolicy):
+class MessageBreaking(initialisation.DefaultInitialisationPolicy,
+                      prediction.DefaultPredictionPolicy,
+                      movement.StepTowardsMMLMixtureMovementPolicy,
+                      repartition.IterativelyRepartitionFromNearestMixturePolicy,
+                      convergence.DefaultConvergencePolicy):
     pass
 
 
-class StrictBayesStepper(initialisation.DefaultInitialisationPolicy,
-                         prediction.DefaultPredictionPolicy,
-                         movement.StepTowardsMMLMixtureMovementPolicy,
-                         repartition.IterativelyRepartitionFromNearestMixturePolicy,
-                         convergence.StrictConvergencePolicy):
+class StrictMessageBreaking(initialisation.DefaultInitialisationPolicy,
+                            prediction.DefaultPredictionPolicy,
+                            movement.StepTowardsMMLMixtureMovementPolicy,
+                            repartition.IterativelyRepartitionFromNearestMixturePolicy,
+                            convergence.StrictConvergencePolicy):
     pass
 
-class BayesJumper(initialisation.LogarithmicInitialisationPolicy,
-                  prediction.DefaultPredictionPolicy,
-                  movement.JumpToMMLMixtureMovementPolicy,
-                  repartition.RepartitionMixtureUsingKMeansPP,
-                  convergence.ConvergedWithSuccessivelyWorseIterations):
-    pass
 
+class MessageJumping(initialisation.DefaultInitialisationPolicy,
+                     prediction.LookaheadFromInitialisationPredictionPolicy,
+                     movement.ConvergeToMixtureWithMaximumExpectationImprovementPolicy,
+                     repartition.RepartitionMixtureUsingKMeansPP):
+    
[…]

### 2018-10-18 scripts/evaluate_search_methods_2d.py

@@ -13,16 +13,21 @@ from time import time
 from gmmmml import GaussianMixture
 
 gmm_kwds = dict(threshold=1e-5, 
-                max_em_iterations=10000,
+                max_em_iterations=100,
                 covariance_regularization=1e-10)
 
 search_strategies = OrderedDict([
-    #("BayesJumper", dict()),
-    ("BayesStepper", dict()),
-    ("KasarapuAllison2015", dict()),
+    ("MessageJumping", dict(covariance_regularization=0)),
+    #("BayesStepper", dict()),
+    #("KasarapuAllison2015", dict()),
 ])
 
 data_paths = sorted(glob("data_2d/*.data"))
+
+#data_paths = [
+#    "data_2d/01000_010000_2.data",
+#    "data_2d/01000_020000_2.data"
+#]
 
 overwrite = False
 raise_if_not_converged = True

### 2018-10-18 scripts/plot_evaluations_2d.py

@@ -21,6 +21,7 @@ matplotlib.style.use(mpl_style)
 search_strategies = OrderedDict([
     ("BayesStepper", dict()),
     ("KasarapuAllison2015", dict()),
+    ("MessageJumping", dict())
 #    ("BayesJumper", dict())
 ])
 
@@ -67,7 +68,8 @@ y_label = "time"
 latex_labels = dict(time=r"$\textrm{time}\,/\,\textrm{seconds}$",
                     K=r"$\textrm{number of true clusters}$ $K$",
                     KasarapuAllison2015=r"$\textrm{Kasarapu \& Allison (2015)}$",
-                    BayesStepper=r"$\textrm{Message-Breaking Method}$")
+                    BayesStepper=r"$\textrm{Message-Breaking Method}$",
+                    MessageJumping=r"$\textrm{Message-Jumping Method}$")
 
 scat_kwds = dict(KasarapuAllison2015=dict(marker="s", s=50),
                  BayesStepper=dict(s=50))

### 2018-10-30 article/ms.tex
words

@@
 
-\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
-derived the bounds
+\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have the bounds
 %\begin{eqnarray}
@@
 
+The negative log-likelihood could be infinitely high given the wrong component parameters.
+When predicting the negative log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$ we seek to
+predict the \emph{minimum} $-\log{\mathcal{L}(\data|\vectheta)}$ for an untrialled mixture with
+$K$ components. We assume that the improvement in $-\log{\mathcal{L}(\data|\vectheta}$ between
+subsequent trials of $K$ improves of order $1/K$ \cite{Li|Barron:2000}. We assume a mean model for the minimal message lengths $I$ of previously trialled mixtures
+with $K$ components as
+\begin{equation}
+	-\log{\mathcal{L}(\data|\vectheta)} = \frac{\alpha}{K} + \beta
+\end{equation}
+\noindent{}where $\alpha$ and $\beta$ are unknowns, and fit these data using a Gaussian
+Process \citep{GPs} where the covariance between the data ($K$, $I$) is modelled by a squared exponential kernel 
+function \citep{GPs;expsquared} with white noise component added to the diagonal of the covariance matrix. 
+We simultaneously optimized the parameters $\alpha$, $\beta$ and the kernel hyperparameters
+\cite{BFGS,Geoge}.
+Given this Gaussian Process model with the optimized parameters, we make predictions for the minimum message length of untrialled mixtures.
+Specifically we are making predictions for the expectation value of the minimum message length of untrialled mixtures, and each prediction of the expectation value has an associated variance in its prediction.
+Given the kernel adopted, the predicted variance will be higher for untrialled mixtures that are farthest away (in $K$) than any trialled mixture (Figure~\ref{fig:mml-predictions}).
+
 \subsubsection{Predicting $\sum\log|\veccov_k|$}
 
+The remaining component whose contribution to the message length is unknown is the sum of
+the log of the determinant of the covariance matrices $\sum\log|\veccov_k|$. For most
+practical scenarios this component is a much smaller order of magnitude than the negative
+log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$. Through experiments with generated
+data sets, we found that the message length to encode $\sum\log|\veccov_k|$
+\begin{equation}
+	I_{sldc} = -\frac{1}{2}(D + 2)\sum\log|\veccov_k|
+\end{equation}
+\noindent{}tended to decrease with increasing $K$ components when the value of $K$ is low
+(e.g., $\approx \frac{1}{5}K_{true}$), and then began to increase linearly with $K$ until
+$K_{true}$. Our experiments suggested that this is merely a consequence of the search
+process, and cannot be reliably used for a heuristic for predicting $K_{true}$, as the
+turnover point occurred when the first true data component is well-described by a single
+model component. That is to say that by $K \approx \frac{1}{5}K_{true}$, there is one
+component covariance matrix that will no longer change significantly, and therefore the
+determinants of the component covariance matrices that are added to the mixture will only
+tend to decrease on average, producing a near linear increase in $I_{sldc}$. Before the
+turnover point, as components are added the average determinants of the covariance matrices
+continues to increase.
+
+ % Remove above paragraph?
+We trialled several heuristics to approximate the change in $\sum\log|\veccov_k|$ between
+mixtures. While some heuristics based on the distribution of distances between data points
+appeared promising, we found good results by simply modelling $I_{sldc}$ as a function
+of the number of trialled components $K$ using a Gaussian Process. We used a Mat\'ern-3/2
+kernel \citep{matern,GPs} function to model the covariance between data points ($K$, $I_{sldc}$) and added
+hyperparameters to fit the data mean, and a white noise component that was added to the
+diagonal of the covariance matrix. We optimized these hyperparameters using the L-BFGS-B
+algorithm \citep{bfgs,scipy,george}, and made predictions for the message length required
+to encode the sum of the log of the determinant of covariance matrices of untrialled
+mixtures.
+
+Like the predictions from our Gaussian Process model for $-\log{\mathcal{L}(\data|\vectheta)}$,
+our predictions for the expectation value of $-\frac{1}{2}(D + 2)\sum\log|\veccov_k|$ (for
+the optimal mixture) come with an associated variance on that prediction.
+
 \subsubsection{Combining the predictions}
 
-% TODO: Predicting the negative log-likelihood.
-
-% TODO: Predicting the sum of the log of the determinant of the covariance matrices.
-
-% Combining it together.
+We sum the predictions in the message length required to describe $\sum\log\weight_k$, the
+negative log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$, and the sum of the log of 
+the determinant of the covariance matrices $\sum\log|\veccov_k|$, as well as constant and
+deterministic terms, to predict the (minimum) message length of an untrialled mixture.
+We sum the variances in the predictions of each component. This process is illustrated in
+Figure~\ref{fig:mml-predictions} where we show the predicted (blue) and actual message lengths 
+(black) for various mixtures. The
+minimum message length over all models $\vec{\mathcal{M}}$ is predicted to be near
+$K \approx \todo{X}$, even though only a few mixtures have been trialled, and the nearest
+trialled mixture is $K = \todo{X}$.
 
@@
 		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
-		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
+		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right] \label{eq:component-message-length}
 \end{eqnarray}
@@
 components \citep{Kasarapu|Allison:2015}.
-After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
-
-One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
-One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$.
+After partial E-M has converged to the same specified tolerance given for the search, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.
+
+One advantage of the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point.
+One possible disadvantage to this approach is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$.
 As a result, the algorithm would converge to a local minima, not the global one.
@@
 The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of computations.
-In the \emph{Message-Breaking Method} we
-try to accomplish this using partial E-M to minimise the total number of computations
+This objective is shared by the \emph{Message-Breaking Method} and the \emph{Message-Jumping Method}, but they differ in their approach.
+
+In the \emph{Message-Breaking Method} we seek to minimize the total number of computations by using  partial E-M to minimise the total number of (effective) E-M steps
 required to converge on a $K+1$ component mixture from a $K$ component mixture.
-However, in the \emph{Message-Jumping Method} we seek to minimise the total number
+In the \emph{Message-Jumping Method} we seek to minimise the total number
 of computations by optimizing a fewer number of mixtures. Instead of iteratively
@@
 
-We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:mbm}. In this
+We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:MBM}. In this
 method we seek to predict the minimum message length for any number of $K$ components,
@@
 we initialise many mixtures that are evenly separated in $\log_{10}$ space from
-$K = 1$ to $K = N/2$ in order to
+$K = 1$ to $K = N/2$ (the maximum possible number of components) in order to
 quickly describe the minimum message length across all $K$. Specifically we
@@
 number of mixtures to initialise -- and those initial $K$ values.
+
 We initialise each mixture using the K-means++ algorithm\cite{kmeanspp} and
 optimize the mixture using EM as described earlier. We trial each of the $K_{init}$
-mixtures in rank order of increasing $K$. If one of these mixtures fails to initialise
-or optimise, this is likely because we could not perform Choleksy decomposition
-of one (or more) of the covariance matrices. In practice this is typically because
-the estimate of $K$ is much higher than $K_{true}$, leading to near-zero values
-along the diagonal of the covariance matrix. In this situation we do not attempt
-to initialise any more of the $K_{init}$ mixtures -- although higher $K$ values
-can be trialled later -- and we being the search method.
+mixtures in order of increasing $K$. If one of these mixtures fails to initialise
+or optimize, this is likely because of zero entries in the diagonal of the covariance
+matrices. In practice this is typically because
+the estimate of $K$ is much higher than $K_{true}$. In this situation we do not attempt
+to initialise any more of the $K_{init}$ mixtures, and we being the search method.
 
@@
 that a human might take: What is an order-of-magnitude estimate on the number of
-$K$? Or what is a ballpark number on the maximum number of $K$ that we should
-trial? After we have that number, and an estimate on the message length for a
+$K$? Or rather: what is a ballpark number on the maximum number of $K$ that we should
+trial? After we have that estimate, and we have an estimate on the message length for a
 few sparsely-sampled numbers of components, we can make reasonable predictions
@@
 the model parameters for that mixture. With these quantities we predict the
-message length of untrialled mixtures for every $K$ value from $K = 1$ to $K = \ceil{1.1\max(K_{trialled})}$ -- e.g., up to 10\% ahead of the largest previously trialled
-$K$ value. The message length predicted from the optimised Gaussian processes
+message length of untrialled mixtures for every $K$ value from $K = 1$ to 
+\todo{some number}. The message length predicted from the optimised Gaussian processes
 provide an expectation value on the message length for an untrialled mixture, as
@@
 Before selecting the next value of $K$ to trial, we use the expectation value and
-associated variance to compute the acquisition function (or the "expected
-improvement") at each value of $K$ \cite{Jones:1998}.
-
-We select the next $K$ value by choosing the value that will maximise the expected
-improvement \todo{but this forces us out to large K, so ....}
-
-\todo{initialisation with kmeans++, if it fails,....}
+associated variance to compute the acquisition function\cite{Jones:1998} (or the "expected
+improvement") at each value of $K$ 
+\begin{equation}
+	\todo{\Psi = X}
+\end{equation}
+\noindent{}and select the next $K$ value by choosing the $K$ that will maximize
+the expected improvement.
+
+After the next $K$ value has been chosen, we initialise a new mixture with $K$
+components using the K-means++ algorithm\citep{kmeanspp} and run E-M until
+convergence. If the initialisation and optimization process fails, we reselect
+the next $K$ value that will maximise the expected improvement. After the mixture
+has converged, we update our predictions for the message length of untrialled
+mixtures and choose the next $K$ value to trial.
 
@@
   \catch
+  	  \State $K_{max} \leftarrow k$ \Comment{Record the maximum $K$ value.}
   	  \State {\bf \sc Break} \Comment{Do not trial any higher $K$ values.}
   \finally 
-      \State mixture[k] $\leftarrow$ mix 
-      \State state[k] $\leftarrow$ \Call{MessageLength}{mixture[k]}
+      \State mixtures[k] $\leftarrow$ mix 
+      \State states[k] $\leftarrow$ \Call{MessageLength}{mix}
   \endtry
@@
 \While{not converged}
-	\State $K_{predict} \leftarrow [1, \cdots, 2\Call{Max}{K_{t}}]$ \Comment{Predict future message lengths.}
-	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$}
+	\State $K_{predict} \leftarrow [1, \cdots, K_{max}]$ \Comment{Predict future message lengths.}
+	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$,mixtures,states}
 	% argsort.
+	\State $A \leftarrow$ \Call{Acquisition}{$K_{predict},I_{predict}$} \Comment{Calculate the expected improvement.}
 	\State $K_{next} \leftarrow$ \Call{ArgSort}{$I_{predict}$} \Comment{Sort $K$ by predicted message length.}
@@
   	        \State {\bf \sc Continue} \Comment{Move on to next $K$ trial.}
+	      \finally
+	      	\State mixtures[k] $\leftarrow$ mix
+			\State states[k] $\leftarrow$ \Call{MessageLength}{mix}
 		  \endtry 
@@
 
+\begin{comment}
 We set the data dimensionality $D = 2$ after previous experiments found that the data dimensionality had only a small effect on the differential cost on the search strategies tested.
@@
 Given these randomly drawn parameters we generated $N\weight_k$ members for each component, rounding as necessary to ensure the total number of generated data points is kept at $N$.
+\end{comment}
 
@@
 emits light at that wavelength. For example, sodium emits and absorbs light 
-at X\,\AA\ when an electron moves between the XY transitions. All else being
-equal, a star with lower normalised flux at a rest-frame wavelength of X\,\AA\
+at 1639.33\,nm in vacuum conditions. All else being
+equal, a star with lower normalised flux at a rest-frame wavelength of 1639.33\,nm
 indicates that the star has a higher abundance of sodium.
 
-We make use of the detailed chemical abundances calculated by \citep{casey}
-from 96,648 APOGEE spectra. This data set includes 15 measured chemical abundances
+We make use of detailed chemical abundances measured for 96,648 APOGEE spectra\citep{Casey:2016}.
+This data set includes 15 measured chemical abundances
 for all stars (e.g., no missing data), with estimated uncertainties. All chemical
@@
 to hydrogen, with a correction to account for the Sun's chemical composition such
-that "[Na/H] = 0" indicates a star has the same abundance of sodium relative to
-hydrogen as the Sun. \todo{To provide context on the fidelity of the data, in Figure \ref{fig:gce} we show the measured abundance
-of [Na/H] with respect to the measured abundance of [Fe/H] for all XX,XXX stars,
-where the Sun's abundance is marked.}
-
-The data set also contains other measured properties of the
-stars, including their temperature and surface gravity, and whether they are 
+that a value of [Na/H] = 0 indicates a star has the same abundance of sodium relative to
+hydrogen as the Sun. To provide some context for the data, in Figure \ref{fig:gce} 
+we show the measured abundance of various elements relative to the measured abundance
+of [Fe/H] for all 96,648 stars. The Sun's abundance is marked.
+
+The data set also contains other measured properties for each star. This 
+includes the effective temperature and surface gravity, and whether they are 
 likely associated with any gravitationally-bound star cluster. Most stars are
@@
 We do not make use of any of this ancillary data (e.g., temperature, cluster
-association) during our search, but we do make use of it to evaluate the performance
-of our search.
+association) during our search, but we do make use of it after the search is
+complete in order to evaluate the performance of our search.
 
@@
 
-Costs of generated data sets.
-Convergence of generated data sets.
-When ther eis a switchover and we are faster.
-
-Expectation that kasarapu and allison should be O(3) and we are O(1) with some overhead.
+We recorded the wall clock time required by each search algorithm on generated data sets, which we show in Figure~\ref{fig:cost}.
+Although wall-clock time metrics are subject to decisions made during implementation, all search and optimisation methods are implemented in the same way (using the same objective function), so the relative differences between methods is reliable.
+The \emph{Message-Breaking Method} and \emph{Message-Jumping Method} have approximately \todo{the same costs in wall time as a function of the number of clusters}.
+
+\todo{Costs of generated data sets.}
+\todo{Switchover in time and we are faster}
+
+\todo{Convergence of generated data sets.}
+
+\todo{Expected complexity of \cite{Kasarapu|Allison:2015} given the search setup}
+
+The \emph{Message-Breaking Method} splits the component with the largest message length (Equation~\ref{eq:component-message-length}), whereas the benchmark method~\cite{Kasarapu|Allison:2015} tries to split, merge, and delete each component.
+After repeated trials on generated data sets, we found that the benchmark method~\cite{Kasarapu|Allison:2015} would most frequently chose to split the same component that the \emph{Message-Breaking Method} would split (e.g., the component with the largest message length).
+We tracked the operation type (e.g., split, delete, merge) and which component was operated upon for \todo{X} operations in \todo{Y} generated data sets for the benchmark method~\cite{Kasarapu|Allison:2015} and the \emph{Message-Breaking Method}.
+
+The benchmark method chose to split in all \todo{X} operations (e.g., no delete or merge operations).
+We also found that the benchmark method usually (X/X; X\%) chose to split the same component that the \emph{Message-Breaking Method} split: the component with the largest message length.
+The results are shown in Figure~\ref{fig:path-operations}.
+Some X\% of the time the component with the second largest message length was split by the benchmark method, X\% of the time with the third-largest, and X\% with the fourth-largest.
+Despite there being X operations with more than 4 components available to split, the benchmark method consistently chose one of four components with the  largest message lengths.
+Despite the path differences taken between the benchmark method and the \emph{Message-Breaking Method}, the optimised mixture parameters and message lengths found by each search algorithm were usually the same within the E-M convergence tolerance.
+This indicates that the \emph{Message-Breaking Method} is just as effective as the benchmark method on generated data sets, and more efficient because it does not need to trial all possible operations on all possible components.
+
+%One advantage of greedily trying to operate on all components is that there is confidence that `the best' operation has been taken, as all other possible operations have been trialled.
+Cost is the most obvious disadvantage to such an approach.
 
@@
 
-\section{Appendices}
-\subsection{The MML objective function}
+\appendix
+
+\section{The MML objective function}
 \label{app:mml_derivation}
@@
 
-\subsubsection{The number of mixture components (or groups) $K$}
+\subsection{The number of mixture components (or groups) $K$}
 
@@
 
-\subsubsection{The relative weight $w_k$ of each component}
+\subsection{The relative weight $w_k$ of each component}
 
@@
 
-\subsubsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
+\subsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
                $K$ Gaussian components}
@@
 
-\subsubsection{The data given the model}
+\subsection{The data given the model}
 If we assume that the data have homoskedastic noise properties, then the 
@@
 be losslessly encoded.  We use an approximation of the lattice constant
-\citep[see Sections 5.1.12 and 3.3.4 of ]
-%[]
-{Wallace:2005} such that,
+(see Sections 5.1.12 and 3.3.4 of \citep{Wallace:2005}) such that,
 \begin{equation}
@@
 
-\subsection{Objective function}
+\subsection{The complete objective function}
 
@@
 
+\begin{comment}
 \section{Bounds on components of the message length}
@@
 \andy{here}
-
-\subsection{Reproducibility} \label{sec:reproducibility}
+\end{comment}
+
+\section{Reproducibility} \label{sec:reproducibility}
 
@@
 \end{center}
-\noindent{}This repository includes a well-tested \texttt{Python 3.6} implementation
+\noindent{}This repository includes a well-tested \texttt{Python 3.7} implementation
 of the algorithms described here, scripts to reproduce our results, and \LaTeX\ to compile this manuscript. This research can be reproduced in full using
@@
 
-\subsubsection{Search Strategy Implementation}
+\section{Code implementation of search strategies}
 

### 2018-10-30 article/ms.tex
track paths

@@
 
+\begin{figure}
+    \includegraphics[width=1.0\textwidth]{paths-splits.pdf}
+    \caption{The number of split operations performed by the benchmark method
+             \cite{Kasarapu|Allison:2015} on the top four components (by message
+             length) from 169 generated data sets. The \emph{Message-Breaking
+             Method} always splits the component with the largest message length.
+             Although the benchmark method tries to split, merge, and delete all
+             components, it most frequently decides to split the same component
+             that the \emph{Message-Breaking Method} splits -- without trialling
+             all operations. The benchmark method chose to split (instead of
+             merge or delete) in all 2029 operations.}
+    \label{fig:path-operations}
+\end{figure}
+
 %One advantage of greedily trying to operate on all components is that there is confidence that `the best' operation has been taken, as all other possible operations have been trialled.

### 2018-10-30 gmmmml/gmm.py

@@ -348,11 +348,14 @@ class GaussianMixture(object):
             # greedy repartition policies).
             self._results.update([strategy.repartition(y, K, **kwds)])
 
+            # This is to find the bug RE I not being a dict. No time now!
+            """
             for k, v in self._results.items():
                 try:
                     v[-1].values()
                 except AttributeError:
                     raise
+            """
 
             index = np.nanargmin(self._state_I)
             K_best, I_best = (self._state_K[index], self._state_I[index])
@@ -370,6 +373,7 @@ class GaussianMixture(object):
         self.means_, self.covs_, self.weights_ = state
         self.responsibilities_, self.log_likelihoods_, self.message_lengths_ = (R, ll, I)
         self.meta_ = meta
+        self.strategy_ = strategy
 
         return self
 

### 2018-10-30 gmmmml/operations.py

@@ -413,6 +413,8 @@ def iteratively_split_components(y, means, covs, weights, K, **kwargs):
         raise ValueError(f"the given mixture already has >={K} components")
 
 
+    meta = dict(I_components_chosen_for_split=[])
+
     Ks = []
     Is = []
     while K > weights.size:
@@ -424,6 +426,8 @@ def iteratively_split_components(y, means, covs, weights, K, **kwargs):
         
         # TODO: Allow us to get back the component-wise relative message lengths
         #       so we don't have to calculate the expectation step twice.
+
+        meta["I_components_chosen_for_split"].append([I_components[index], np.sum(I_components)])
 
         (state, responsibilities, ll, I) = split_component(
             y, means, covs, weights, R, index, **kwargs)
@@ -442,7 +446,10 @@ def iteratively_split_components(y, means, covs, weights, K, **kwargs):
     ax.scatter(Ks, Is)
     """
 
-    return (state, responsibilities, ll, I)
+    debug = kwargs.get("debug", False)
+
+    return (state, responsibilities, ll, I) if not debug \
+                                            else (state, responsibilities, ll, I, meta)
 
 
 def simultaneously_split_mixture(y, means, covs, weights, K, **kwargs):

### 2018-10-30 gmmmml/policies/repartition.py

@@ -6,7 +6,7 @@ from collections import defaultdict
 from tqdm import tqdm
 
 from .base import Policy
-from .. import operations as op
+from .. import (em, operations as op)
 
 logger_name, *_ = __name__.split(".")
 logger = logging.getLogger(logger_name)
@@ -112,15 +112,28 @@ class GreedilyPerturbNearestMixturePolicy(BaseRepartitionPolicy):
         bop, bp = min(best_perturbations.items(), key=lambda x: x[1][0])
         logger.debug(f"Best operation is {bop} on index {bp[1]}")
 
+
+        debug = kwargs.get("debug", False)
+
         if ml(bp[-1]) >= ml(prev_I):
             logger.info("All perturbations are worse.")
-            return (K, (state, R, ll, I))
+            return (K, (state, R, ll, prev_I, dict())) \
+                if debug else (K, (state, R, ll, prev_I))
+
+
+        if debug:
+            _, __, I_components = em._component_expectations(y, *state, **kwargs)
+            meta = dict(I_component_chosen=np.hstack([I_components[bp[1]], I_components]),
+                        operation=bop)
+
 
         _, __, state, R, ll, I = bp
 
         K = _K_from_state(state)
-        
-        return (K, (state, R, ll, I))
+
+        return (K, (state, R, ll, I, meta)) if debug else (K, (state, R, ll, I))
+
+
 
 
 def _K_from_state(state):

### 2018-10-30 gmmmml/strategies.py

@@ -14,6 +14,7 @@ class MessageBreaking(initialisation.DefaultInitialisationPolicy,
     pass
 
 
+
 class StrictMessageBreaking(initialisation.DefaultInitialisationPolicy,
                             prediction.DefaultPredictionPolicy,
                             movement.StepTowardsMMLMixtureMovementPolicy,
@@ -49,3 +50,50 @@ class KasarapuAllison2015(initialisation.DefaultInitialisationPolicy,
                           repartition.GreedilyPerturbNearestMixturePolicy,
                           convergence.DefaultConvergencePolicy):
     pass
+
+
+
+class DebugMessageBreaking(MessageBreaking):
+
+    def __init__(self, *args, **kwargs):
+
+        self._message_length_of_components_split = []
+        super(DebugMessageBreaking, self).__init__(*args, **kwargs)
+
+
+    def repartition(self, y, K, **kwargs):
+
+        kwds = kwargs.copy()
+        kwds["debug"] = True
+
+        K, result = super(DebugMessageBreaking, self).repartition(y, K, **kwds)
+        state, R, ll, I, meta = result
+
+        self._message_length_of_components_split.append(meta["I_components_chosen_for_split"])
+
+        return (K, (state, R, ll, I))
+
+
+class DebugKasarapuAllison2015(KasarapuAllison2015):
+
+    def __init__(self, *args, **kwargs):
+      self._message_length_of_components_split = []
+      super(KasarapuAllison2015, self).__init__(*args, **kwargs)
+
+
+    def repartition(self, y, K, **kwargs):
+
+        kwds = kwargs.copy()
+        kwds["debug"] = True
+
+        K, result = super(KasarapuAllison2015, self).repartition(y, K, **kwds)
+        state, R, ll, I, meta = result
+
+        if "I_component_chosen" in meta:
+
+            self._message_length_of_components_split.append(meta["I_component_chosen"])
+
+            # Just for running path traces! 
+            assert meta["operation"] == "split"
+
+        return (K, (state, R, ll, I))

### 2018-10-30 gmmmml/utils.py

@@ -440,7 +440,7 @@ def generate_isotropic_data(N=None, D=None, K=None, cluster_std=1.0,
 
 
     
-    responsibility, log_likelihood, lp = responsibilities(
+    responsibility, log_likelihood = responsibilities(
         X, mean, cov, weight, full_output=True, covariance_type="full")
 
     nll = -np.sum(log_likelihood)

### 2018-10-30 sandbox_plot_track_paths.py

@@ -1,0 +1,59 @@
+
+
+import numpy as np
+import pickle
+import matplotlib
+import matplotlib.pyplot as plt
+
+from mpl_utils import mpl_style
+
+matplotlib.style.use(mpl_style)
+
+with open("track_paths_results.pkl", "rb") as fp:
+    results = pickle.load(fp)
+
+
+# Make an indices plot.
+indices = np.vstack([ea[0] for ea in results if np.array(ea[0]).size > 0]).T[0]
+
+bins = np.arange(max(indices) + 2) - 0.5
+heights, _ = np.histogram(indices, bins=bins)
+x = np.arange(1 + max(indices)).astype(int)
+
+fig, ax = plt.subplots(figsize=(9, 3))
+ax.barh(x, heights, facecolor="tab:blue")
+ax.set_yticks(x)
+ax.set_yticklabels([
+    r"$\textrm{Component with largest} I_k$",
+    r"$\textrm{Component with second-largest} I_k$",
+    r"$\textrm{Component with third-largest} I_k$",
+    r"$\textrm{Component with fourth-largest} I_k$"
+])
+ax.set_ylim(ax.get_ylim()[::-1])
+
+ax.text(0.975, 0.05, 
+        r"$N_{{operations}} = {{{0}}}$".format(indices.size),
+        transform=ax.transAxes, 
+        verticalalignment="bottom",
+        horizontalalignment="right")
+
+ax.text(0.975, 0.13,
+        r"$N_{{datasets}} = {0}$".format(len(results)),
+        transform=ax.transAxes,
+        verticalalignment="bottom",
+        horizontalalignment="right")
+
+
+ax.set_xlabel(r"$\textrm{Number of component splits by benchmark method}$")
+ax.yaxis.set_tick_params(width=0)
+
+fig.tight_layout()
+
+
+
+# Difference between Message Breaking Method and Kasarapu:
+diff_I = np.array([(ea[1] - ea[4]) for ea in results])
+
+print("Max absolute difference in message length between methods: {0}".format(np.max(np.abs(diff_I))))
+
+fig.savefig("article/paths-splits.pdf", dpi=300)

### 2018-10-30 sandbox_track_paths.py

@@ -1,0 +1,101 @@
+
+"""
+Evaluate greedy paths compared to a policy where we only split the component
+with the largest message length.
+"""
+
+import os
+import numpy as np
+import pickle
+import matplotlib.pyplot as plt
+
+from gmmmml import utils, GaussianMixture
+
+random_seed = 123
+np.random.seed(random_seed)
+
+overwrite = True
+
+output_path = "track_paths_results.pkl"
+
+gmm_kwds = dict(threshold=1e-5, max_em_iterations=1000, covariance_regularization=0)
+search_kwds = dict(quiet=True)
+
+M = 1000
+
+D = lambda *_: 2
+K = lambda *_: int(np.random.uniform(1, 25))
+N = lambda *_: int(np.random.uniform(1000, 10000))
+
+ml = lambda I: I if isinstance(I, (int, float)) else np.sum(np.hstack(I.values()))
+
+
+data_kwds = dict(center_box=(-10, 10), cluster_std=0.1)
+
+if os.path.exists(output_path) and not overwrite:
+    raise IOError(f"output path exists ({output_path}) and told not to overwrite")
+
+
+results = []
+
+for m in range(M):
+
+    dk = data_kwds.copy()
+    dk.update(N=N(), D=D(), K=K())
+    print("DATA KWDS", dk)
+
+    X, y, target, meta = utils.generate_isotropic_data(**dk)
+
+    # Run Message-Breaking Search and track the path.
+
+    model = GaussianMixture(**gmm_kwds)
+    model.search(X, search_strategy="DebugMessageBreaking", **search_kwds)
+
+
+    # Run Kasarapu-Allison and track the path.
+    ka = GaussianMixture(**gmm_kwds)
+    ka.search(X, search_strategy="DebugKasarapuAllison2015", **search_kwds)
+
+    # Track the paths that were followed by Kasarapu & Allison.
+
+    # sorted_index_chosen
+    # diff_from_max_component_chosen
+    # percentage_diff_from_max_component_chosen
+
+    trace_path = []
+
+    for each in ka.strategy_._message_length_of_components_split:
+
+        I_chosen, *I_all = each
+        sorted_index_chosen = len(I_all) - 1 - np.searchsorted(np.sort(I_all), I_chosen)
+        diff_from_max_component_chosen = I_chosen - np.max(I_all)
+        percentage_diff_from_max_component_chosen = diff_from_max_component_chosen/np.sum(I_all)
+
+        denominator = np.max(I_all) - np.min(I_all)
[…]
